### 2023-06-27 doc/manuscript.tex
add manuscript to repo

\documentclass{article}
\usepackage[round]{natbib}

\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{color}

\usepackage{xr}
\externaldocument{supplement}

\newcommand{\aprcomment}[1]{{\textcolor{blue}{Comment: #1}}}
\newcommand{\E}{\mathbb{E}}

\title{Stabilizing selection, archaic introgression, and the distribution
of shared functional variation}
\author{APR}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
    .
\end{abstract}

\section*{Introduction}

Genomic studies have shown that admixture is a common occurrance between
diverged populations and closely related taxa \citep{}, and potentially has
been widespread in primate \citep{} and hominin \citep{} evolution.
Understanding the effects of admixture on phenotypic and molecular variation is
therefore relevant to many natural systems, including to studies of the genetic
basis of complex traits and diseases. In particular, archaic introgression in
\emph{Homo sapiens} involving Neanderthals and Denisovans has recently
attracted considerable attention, both in understanding the historical
processes leading to observed distributions of introgressed DNA in present-day
populations \citep{prufer2014complete,villanea20??,chen2020identifying} as well
as its contribution to quantitative traits \citep{sankararaman,wei,others}.

After their introduction through admixture, introgressed alleles may be
selected either for or against. Some introgressed haplotypes appear to have
been positively selected in modern humans
\citep{huerta2014altitude,gower2021detecting,Enard}, possibly because such
alleles are locally adaptive and provide a fitness advantage when encountering
a novel environment. Despite these cases of adaptive introgression,
introgressed alleles were more likely to have been negatively selected in
modern humans \citep{harris2016genetic,juric2016strength}. As Neanderthal and
Denisovan population sizes are thought to have been relatively small for
hundreds of thousands of years, population genetics theory predicts that they
would accumulate deleterious variation at a faster rate than larger
populations.  Introgressed haplotypes would then carry more deleterious
variants, which would quickly be selected against after admixture. In mapping
the distribution of haplotypes identified as introgressed from Neanderthals,
there is a reduction of Neanderthal-related ancestry in enhancers and
regulatory regions
\citep{petr2019limits,telis2020selection,yermakovich2023long}. This
Neanderthal-allelic depletion, or ``deserts,'' support the hypothesis that
introgressed functional alleles were selected against
\citep{sankararaman2014genomic,sankararaman2016combined}.

There is genetic evidence that early \emph{H. sapiens} reciprocally contributed
to Neanderthal genomes \citep{kuhlwilm2016ancient,hubisz2020mapping}, with
introgression in the reverse direction occurring tens or hundreds of thousands
of years before the Neanderthal-to-human event surrounding the out-of-Africa
expansion around 60 thousand years ago (ka). This is supported by early ``near
modern'' \emph{H.  sapiens} outside of African around 120--100ka or earlier
\citep{schwarcz1988esr,grun2005u,beyer2021climatic}, potentially overlapping
with Neanderthals and providing opportunities for early contacts. While
estimates of the genomic contribution of early \emph{H. sapiens} to
Neanderthals vary \citep{}, up to XX\% of later Neanderthal genomes may have
been contributed through introgression from humans to Neanderthals. Under the
same argument that \emph{H. sapiens}-related haplotypes carry fewer deleterious
alleles, human-introgressed DNA may have been favored in Neanderthal genomes.
The replacement of Neanderthal mitochondrial and Y chromosomes by early human
haplotypes supports this model \citep{posth2017deeply,petr2020evolutionary}.

Models for selection on introgressed alleles are often based on load arguments
(in particular, differences in the rate of accumulation of unconditionally
deleterious variation in populations of different sizes) or hybrid
incompatibilities \citep{mueller,etc}. These arguments are founded primarily in
population genetics theory, eschewing any notion of genetic variation
contributing to phenotypic variation and selection operating on those
phenotypes. However, most phenotypic traits are thought to be under stabilizing
selection \citep{sanjak2018evidence}, including gene regulation
\citep{gilad2006natural,hodgins2015gene,price2022detecting}. This is especially
relevant for selection on Neanderthal-introduced alleles in modern humans, as
the strongest signals of selection against such alleles occur in regulatory
regions \citep{sankararaman2014genomic}.

Stabilizing selection on a trait acts to maintain the phenotypic distribution
of that trait near some trait optimum, which is achieved by reducing phenotypic
variation (Figure~\ref{fig:one-pop}A). When the mean phenotype of the
population is close to the phenotypic optimum, classical models predict that
alleles contributing to such a trait are subject to underdominant selection,
i.e., selection against the minor allele at a locus
\citep{robertson1956effect}. This has proven to be a useful model for
understanding allele frequencies of variants contributing to a trait under
stabilizing selection in the single-population setting
\citep[e.g.,][]{keightley1988quantitative, simons2018population,
hayward2022polygenic}.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/one_pop.pdf}
    \caption{
        \textbf{Stabilizing selection and additive genetic variance in one population.}
        \aprcomment{See comparisons to simulations assuming free recombination in
        Figures~\ref{fig:one-popA}--\ref{fig:one-popC}.}
    }
    \label{fig:one-pop}
\end{figure}

In two diverged populations, the genomic architecture (i.e., mutations
underlying the trait) of a trait under stabilizing selection has a higher rate
of turnover compared to neutrally evolving loci \citep{yair2022population}. We
might therefore expect an accumulation of fixed differences between the
\emph{H. sapiens} and Neanderthal branches for alleles contributing to a trait,
even when the mean phenotype in each population is close to the same trait
optimum. When alleles that are fixed in one population are introduced to
another population in which they were previously absent, they will be at low
frequency and subsequently selected against. Likewise, if the ancestral allele
is reintroduced to a population fixed for a derived allele, the ancestral
allele will be at low frequency and will be selected against. In either case we
should expect the introgressed allele, whether ancestral or derived, to be
selected against, regardless the historical relative sizes of the populations
involved. This prediction contrasts with the population-genetics ``load''
model, in which haplotypes with fewer deleterious variants (such as those from
the population with larger historical size) are favored after introgression in
either direction.

\subsection*{Theory and model}

We consider a polygenic trait such that an individual's (additive) genetic
value is the sum over all effects of alleles in their genome, so that for
individual $i$, \(G_i = \sum_l g_{l, i} a_l\), where $a_l$ is the effect size
of the derived allele at locus $l$, and \(g_{l,i}\in\{0,1,2\}\) is their
genotype at that locus (i.e., the number of alleles they carry at that locus).
For unlinked loci, the expected additive genetic variance is \(V_A = \sum_l
2p_l(1-p_l)a_l^2\), where $p_l$ is the allele frequency at locus $l$. We ignore
dominance and epistatic effects, so that \(V_G = V_A\), assuming no linkage.
We can further ingore environmental effects \citep{simons2018population}, so
that the phenotypic variance \(V_P=V_A\).

Stabilizing selection acts to reduce phenotypic variation around the optimum
value $o$, and we assume a Gaussian fitness function so that relative fitness
is given by \(f(G_i | O, V_S) = \exp{(-(G_i - O) / 2 V_S)}\). $V_S$ can be
interpreted as the strength of selection on the trait, where larger $V_S$
implies weaker selection. For a population with mean phenotype at (or very
close to) the optimum, the mean fitness of the population (assuming a roughly
normal distribution of phenotypic values in the population) is
\[\bar{w} \approx \int_{-\infty}^\infty f(G | O, V_S) \mathcal{N}(0, V_G) dG
\approx \left(\frac{V_S}{V_S+V_G}\right)^{1/2}.\]
Thus, as the genetic variance increases, mean fitness among individuals in the
population decreases.

\subsubsection*{Mutation rates and effect sizes}

If all alleles contribute equally to the trait (as \(\pm a\)),
\citet{keightley1988quantitative} showed that the dynamics of $V_G$ can be
approximated with the recursion
\[V_{G,t+1} \approx V_{G,t}\left(1-a^2/2(V_S+V_G)\right)
\left(1-1/2N_e\right) + 2 \mu a^2,\]
where $\mu$ is the per-haploid, per-generation rate of mutation. In the
large-population-size limit, this gives the well-known results
\[\tilde{V_G} \approx 4\mu V_S,\] provided \(V_G \ll V_S\).
Mutations are not expected to each have the same effect size $a$, but
rather will be drawn from some distribution. Here, we will assume mutation
effect sizes are drawn from a normal distribution with mean 0 and given
variance $V_M$.

\subsubsection*{Approximating allelic dynamics via underdominance}

We model the distribution of allele frequencies underlying the trait using
the approximation that alleles are subject to symmetric underdominant selection.
For an allele with effect size $a$ (in the heterozygous state -- homozygotes
contribute $2a$ to the trait), the selection coefficient
\[s\approx \frac{a^2}{2(V_S + V_G)} \approx \frac{a^2}{2V_S},\]
if $V_G \ll V_S$ \citep[e.g.,][]{simons2018population}. Allelic dynamics can
be modeled using the diffusion approximation, where the expected change
in mean allele frequency per generation is
\[\E[M_{\delta_p}] \approx -s p(1-p)(1-2p),\]
and the expected change in variance of the allele frequency per generation is
\[\E[V_{\delta_p}] \approx \frac{1}{4N_e}p(1-p).\]
From this, because $s$ is always positive for any $a\not=0$, we see that
selection pushes allele frequencies to zero if $p<1/2$ and to one if $p>1/2$.
$p=1/2$ is an unstable equilibrium.

We extend the moments-based solution for the sample site-frequency spectrum
(SFS) \citep{jouganous2017} to include underdominance with given selection
coefficient $s$ (\(=a^2/2(V_S+V_G)\)). The contribution of alleles with effect
size $a$ to the total genetic variance $V_G$ is then found by computing the
expected pairwise diversity from the SFS (with sample size $n$, $\Phi_n$), as
\(V_{G,a} = 2a^2\sum_{j=1}^{n-1}\frac{j(n-j)}{n(n-1)}\Phi_n(j|a,\mu_a)da\).
Here, $\mu_a$ is the mutation rate of alleles with effect size $a$. The total
genetic variance is then \(V_G=\int_{-\infty}^\infty V_{G,a}
\mathcal{N}(0,V_M).\)

Figure~\ref{fig:one-pop}C shows that if $V_G$ is non-negligible compared to
$V_S$, ignoring $V_G$ and using $s=a^2/2V_S$ leads to lower estimates of
additive genetic variance. When mutation rates are large so that $V_G$ is not
small compared to $V_S$, using $s=a^2/2(V_S+V_G)$ provides estimates of $V_G$
that closely match simulations assuming free recombination between loci
(Figure~\ref{fig:toy-admixture}C,D). Because $V_G$ can change over time, this
means that $s$ is no longer constant, but can change due to factors such as
non-constant demography that increase or reduce $V_G$.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/reciprocal_admixture.pdf}
    \caption{
        \textbf{}
    }
    \label{fig:toy-admixture}
\end{figure}

\subsubsection*{Demographic history}

Our numerical solution for the SFS allows for non-constant population size
histories, population splits, continuous migration, and admixture. In this
study, we consider relativey simple scenarios involving population splits with
subsequent introgression events. We focus on parameter regimes relevant to
human-Neanderthal history. In the simplest model -- which is not meant to
perfectly match any particular known or inferred history, but instead
demonstrate the effect of reciprocal introgression following population
divergence -- a population of size $N_e=10\,000$ splits into two, on remaining
size $10\,000$ and the other shrinking to size $1\,000$. They remain isolated
for $2N_e$ generations (or 500 thousand years, assuming an average generation
time of 25 years) and then introgression occurs from one branch to the other,
contributing $5\%$ ancestry to the recipient population
(Figure~\ref{toy-admixture}A,B).

The second model is meant to more closely resemble inferred human-Neanderthal
history, in which the ancestral population of size $N_e=10\,000$ splits into
the human and Neanderthal branches. At 250ka, an early human-to-Neanderthal
introgression contributes $5\%$ ancestry to Neanderthals. The human branch
shrinks to size $1\,000$ 60ka, followed by exponential growth to $20\,000$ at
present time. Neanderthals contribute $2\%$ ancestry to this bottlenecked and
expanding human population at 50ka, after which their branch ceases
(Figure~\ref{fig:neand-to-human}A).

In each scenario, we track genetic diversity and phenotypic variance before and
after admixture to understand allele frequency and haplotype dynamics. The
trait optimum is $0$ in all branches, and the strength of selection $V_S=1$
remains constant.

\section*{Results}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/neanderthal_admixture.pdf}
    \caption{
        \textbf{}
    }
    \label{fig:neand-to-human}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/human_admixture.pdf}
    \caption{
        \textbf{}
    }
    \label{fig:human-to-neand}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/linkage_predictions.pdf}
    \caption{
        \textbf{}
    }
    \label{fig:linkage-pred}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/linkage_simulation.pdf}
    \caption{
        \textbf{}
    }
    \label{fig:linkage-sim}
\end{figure}

\clearpage

\bibliographystyle{genetics}
\bibliography{manuscript}

\section{Appendix}

\subsection{Additive genetic variance after admixture}

Suppose two pouplation (labeled 0 and 1) diverged some time in the past and
then admix in proportions $f$ and $1-f$.

Assuming no linkage, dominance or epistasis (so \(V_G=V_A\)),
\[V_G = \sum_l 2p_l(1-p_l)a_l^2 = \sum_l \pi_l a_l^2,\]
where $\pi$ denotes expected pairwise diverisity.
At a given locus (dropping the $l$), after admixture the allele frequency is
\[p=f p_0 + (1-f) p_1,\]
so that
\begin{align*}
    2p(1-p) & = 2(f p_0 + (1-f) p_1)(1 - f p_0 - (1-f) p_1) \\
    & = f^2 2p_0(1-p0) + (1-f)^2 p_1(1-p1) + 2f(1-f) (p_0(1-p_1) + p_1(1-p_0)) \\
    & = f^2 \pi_{0,0} + (1-f)^2 \pi_{1,1} + 2f(1-f)\pi_{0,1}.
\end{align*}

Plugging in to the definition for $V_G$, we get after admixture
\[V_G = f^2 V_{G,0} + (1-f)^2 V_{G,1} + 2f(1-f)\sum_l \pi_{0,1,l}a_l^2.\]
We can write $\pi_{0,1}$ at a given locus in terms of $\pi_{0,0}$, $\pi_{1,1}$,
and $F_2(0,1)=(p_0-p_1)^2$ as \citep{peter2016admixture}
\[\pi_{0,1} = F_2(0,1) + \frac{1}{2}\pi_{0,0} + \frac{1}{2}\pi_{1,1}.\]
Then
\begin{align*}
    V_G & = f^2 V_{G,0} + (1-f)^2 V_{G,1} + 2f(1-f)\sum_l \left[F_{2,l}(0,1)
    + \frac{1}{2}\pi_{0,0} + \frac{1}{2}\pi_{1,1}\right] a_l^2 \\
    & = f^2 V_{G,0} + (1-f)^2 V_{G,1} + 2f(1-f)\left[\frac{1}{2}V_{G,0} 
    + \frac{1}{2}V_{G,1} + \sum_l F_{2,l}(0,1)a_l^2\right] \\
    & = \left(f^2 + f(1-f)\right) V_{G,0} + \left((1-f) + f(1-f)\right)V_{G,1}
    + 2f(1-f) \sum_l F_{2,i}(0,1)a_l^2 \\
    & = f V_{G,0} + (1-f)V_{G,1} + 2f(1-f)\sum_i F_{2,l}(0,1)a_l^2.
\end{align*}

This shows that additive genetic variance after admixture is equal to the
additive variances in the source populations, weighted by their contributes,
plus a term that depends on the divergence at loci contributing to the trait
between the two populations, measured as $F_2$, weighted by the quadratic
factor $2f(1-f)$.

In general, $F_2$ at a given locus will depend on the demography connecting the
two populations, as well as the effect size at the locus due to selection on
the trait. In the infinitesimal limit, involving many loci each of vanishingly
small effect, dynamics at a given locus will be approximately neutral, so that
$F_2$ will depend only on the demography. In this case
\begin{align*}
    V_G & \approx f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] \sum_l a_l^2 \\
    & = f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] n \sigma_m^2
\end{align*}
Here, $\sigma_m^2$ is the variance of effect sizes of new mutations (assuming mean is
zero).

\begin{figure}[ht!]
    \centering
    %\includegraphics{}
    \caption{
        \textbf{Dynamics of additive genetic variance in two populations with admixture.}
        Coupled with the equations for the decay of $V_G$ due to drift and selection,
        and the increase in $V_G$ through mutation, show that the expected dynamics
        are matched by simulations without linkage.
    }
    \label{fig:VG_dynamics}
\end{figure}

\subsection{Dynamics of introgressed alleles}

\subsection{Reciprocal depletion of functional variation}

While incompatibilities require two or more loci (Morgan, 1942), introgression
with stabilizing selection can create the appearance of a single-locus
incompatibility via underdominance\dots

\subsection{The role of linkage}

\section{Discussion}

\section{Methods}

\end{document}

### 2023-06-27 doc/supplement.tex
add manuscript to repo

\documentclass[]{article}
\usepackage[round]{natbib}

\usepackage{fullpage}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}

% cross-reference with main text
\usepackage{xr}
\externaldocument{manuscript}

% local definitions
\newcommand{\aprcomment}[1]{{\textcolor{blue}{Comment: #1}}}

\title{Supporting Information}
\author{APR}
\date{\today}

\begin{document}
\maketitle

\section{Supplemental Figures}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/one_pop.a_0.01.pdf}
    \caption{
        \textbf{Simulations assuming no linkage with weak effects.}
        All mutations have $a=0.01$.
    }
    \label{fig:one-popA}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/one_pop.a_0.04.pdf}
    \caption{
        \textbf{Simulations assuming no linkage with moderate effects.}
        All mutations have $a=0.01$.
    }
    \label{fig:one-popB}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/one_pop.a_0.1.pdf}
    \caption{
        \textbf{Simulations assuming no linkage with strong effects.}
        All mutations have $a=0.1$.
    }
    \label{fig:one-popC}
\end{figure}

\end{document}

### 2023-06-27 figures/dist_introgressed_variants.py

@@ -1,0 +1,318 @@
+import numpy as np
+import matplotlib.pylab as plt, matplotlib
+import moments
+import demes, demesdraw
+import scipy.stats
+
+from single_population_stabilizing_selection import *
+
+# set font sizes
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=8)
+matplotlib.rc("axes", titlesize=8)
+matplotlib.rc("legend", fontsize=7)
+
+
+colors = plt.rcParams["axes.prop_cycle"].by_key()["color"]
+
+Ne = 1e4
+N1 = 1e4
+N2 = 1e3
+T1 = 500000 / 25
+T2 = 100000 / 25
+f = 0.05
+b = demes.Builder(time_units="generations")
+b.add_deme(
+    "Ancestral",
+    epochs=[dict(end_time=T1 + T2, start_size=Ne)],
+)
+b.add_deme("Deme1", ancestors=["Ancestral"], epochs=[dict(end_time=0, start_size=N1)])
+b.add_deme("Deme2", ancestors=["Ancestral"], epochs=[dict(end_time=0, start_size=N2)])
+b.add_pulse(sources=["Deme1"], dest="Deme2", time=T2, proportions=[f])
+g1 = b.resolve()
+
+b.data["pulses"][0]["sources"] = ["Deme2"]
+b.data["pulses"][0]["dest"] = "Deme1"
+g2 = b.resolve()
+
+
+# assumes symmetry +/- effect sizes, and population mean is at optimum of zero
+def gaussian_des(a_list, VG_list, SD, hill=False):
+    """
+    a_list: vector of effect size points
+    VG_list: computed VG for given theta from SFS
+    SD: standard deviation of the distribution of effect sizes of new mutations
+    """
+    assert len(a_list) == len(VG_list)
+    S = 0
+    aa = np.concatenate(([0], a_list))
+    dxs = (aa - np.concatenate(([aa[0]], aa))[:-1]) / 2 + (
+        np.concatenate((aa, [aa[-1]]))[1:] - aa
+    ) / 2
+    weights = scipy.stats.norm.pdf(a_list, scale=SD)
+    for v, dx, w in zip(VG_list, dxs[1:], weights):
+        S += 2 * v * dx * w
[…]

### 2023-06-27 figures/linkage_reduction_model.py

@@ -1,0 +1,136 @@
+import numpy as np
+import matplotlib.pylab as plt, matplotlib
+from mpl_toolkits.axes_grid1 import make_axes_locatable
+
+
+# set font sizes
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+
+colors = plt.rcParams["axes.prop_cycle"].by_key()["color"]
+
+
+def evolve_generation(p, q, D, r, s):
+    # with underdominance (s > 0)
+    p2 = p - s * p * (1 - p) * (1 - 2 * p)
+    q2 = q - s * D * (1 - 2 * p)
+    D2 = D - s * D * (1 - 2 * p) ** 2 - r * D
+    return p2, q2, D2
+
+
+def trajectories(p0, q0, D0, generations=1, r=0, s=0):
+    p = [p0]
+    q = [q0]
+    D = [D0]
+    for g in range(generations):
+        p2, q2, D2 = evolve_generation(p[-1], q[-1], D[-1], r, s)
+        p.append(p2)
+        q.append(q2)
+        D.append(D2)
+    return p, q, D
+
+
+def plot_split_trajectories(a, r, gens, p0, fig, ax, VS=1):
+    s = a ** 2 / 2 / VS
+    # divide subplot to show top and bottom
+    divider = make_axes_locatable(ax)
+    axb = divider.new_vertical(size="100%", pad=0.1)
+    fig.add_axes(axb)
+
+    q0 = p0
+    # initially fully linked
+    D0 = p0 - p0 * q0
+    p, q, D = trajectories(p0, q0, D0, generations=gens, r=r, s=a ** 2 / 2 / VS)
+
+    ax.plot(p, label="Selected allele frequency")
+    ax.plot(q, label="Linked allele frequency")
+    ax.plot(D, label="LD ($D=Cov(p,q)$)")
+    ax.set_ylim(0, p0 * 1.5)
+    ax.set_xlabel("Generations")
+
+    q0 = 1 - p0
+    # initially fully linked
+    D0 = q0 - q0 ** 2
[…]

### 2023-06-27 figures/linkage_reduction_simulation.py

@@ -1,0 +1,97 @@
+import numpy as np
+import matplotlib.pylab as plt, matplotlib
+from mpl_toolkits.axes_grid1 import make_axes_locatable
+
+
+# set font sizes
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+
+colors = plt.rcParams["axes.prop_cycle"].by_key()["color"]
+
+
+def evolve_generation(p, q, D, r, s):
+    # with underdominance (s > 0)
+    p2 = p - s * p * (1 - p) * (1 - 2 * p)
+    q2 = q - s * D * (1 - 2 * p)
+    D2 = D - s * D * (1 - 2 * p) ** 2 - r * D
+    return p2, q2, D2
+
+
+def trajectories(p0, q0, D0, generations=1, r=0, s=0):
+    p = [p0]
+    q = [q0]
+    D = [D0]
+    for g in range(generations):
+        p2, q2, D2 = evolve_generation(p[-1], q[-1], D[-1], r, s)
+        p.append(p2)
+        q.append(q2)
+        D.append(D2)
+    return p, q, D
+
+
+def get_expected_frequency_dip(a, r_max, gens, p0, VS=1):
+    rs = np.linspace(0, r_max, 101)
+    rs_plot = 100 * np.concatenate((-rs[:0:-1], rs))
+    qs = np.zeros(len(rs))
+    s = a ** 2 / 2 / VS
+    for i, r_dist in enumerate(rs):
+        p, q, D = trajectories(p0, p0, p0 - p0 ** 2, generations=gens, r=r_dist, s=s)
+        qs[i] = q[-1]
+    qs = np.concatenate((qs[:0:-1], qs))
+    return rs_plot, qs
+
+
+if __name__ == "__main__":
+    p0 = 0.05
+    q0 = p0
+    # initially fully linked
+    D0 = p0 - p0 * q0
+
+    VS = 1
+    gens = 2000
+
+    fig = plt.figure(6, figsize=(6.5, 2.5))
+    fig.clf()
+
+    ax1 = plt.subplot(1, 2, 1)
+    a = 0.05
+
[…]

### 2023-06-27 doc/manuscript.tex
some results

@@
 
+\subsection*{Additive genetic variance after admixture}
+
+We expect genetic variance to increase after introgression. The amount that
+genetic variance increases depends on the allelic differences accumulated
+between populations and the effects of those alleles. Assuming no linkage,
+dominance or epistasis, \(V_G=\sum_l 2p_l(1-p_l)a_l^2\). After admixture, with
+proportion $f$ contributed by the source population (labeled 0) into the focal
+population (labeled 1), \(p_l = fp_{l,0} + (1-f)p_{l,1}\). Plugging into the
+expression for $V_G$, and after some simple algebra (Appendix~AX), we can
+express the expected genetic variance directly after admixture as \[V_G =
+fV_{G,0} + (1-f) V_{G,1} + 2f(1-f)\sum_l F_{2,l} a_l^2,\] where $F_2 = (p_0 -
+p_1)^2$ is the squared difference in allele frequencies at a locus
+\citep[e.g.,][]{peter2016admixture}. This result in known \citep[e.g.,][]{},
+showing that additive variance is equal to that in the source populations
+weighted by their contributions, plus a term that depends on the divergence at
+trait-affecting loci between the populations weighted by the quadratic factor
+$2f(1-f)$.
+
+In general, $F_2$ at a given locus will depend on the demographic history
+relating the two populations and the effect size at the locus due to selection
+on the trait. In the infinitesimal limit, involving many loci each of
+vanishingly small effect, dynamics at a given locus will be approximately
+neutral, so that $F_2$ depends only on the demography. In this case
+\begin{align*}
+    V_G & \approx f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] \sum_l a_l^2 \\
+    & = f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] n_m V_M,
+\end{align*}
+where $V_M$ is the variance of effect sizes of new mutations (assuming the mean
+is zero).
+
+For $a \not\approx 0$, assuming neutral evolution will tend to overestimate
+$F_2$ compared to underdominant selection. We can compute expected $V_G$ both
+before and after admixture using the diffusion approximation for the joint SFS
+with underdominance. Comparing to simulations with free recombination between
+loci (Appendix~AY), we find that this provides an excellent fit to average
+observed $V_G$ over time (Figure~\ref{fig:toy-admixture}C,D). Following an
+initial rapid increase in $V_G$ due to admixture, it decays back to the
+steady-state expectation relatively rapidly.
+
+\begin{itemize}
+    \item Benefit of SFS -- partition by frequency classes (prev. seg vs introgressed)
+    \item VG from all both previously segregating and introgressed alleles decays,
+        replaced by VG due to new mutations
+    \item Dividing VG from introgressed alleles into introduced derived vs reintroduced
+        ancestral alleles shows the effect of differences in population sizes
+\end{itemize}
+
+\subsection*{Recurrent introgression and non-constant population size}
+
 \begin{figure}[ht!]
@@
 
-Suppose two pouplation (labeled 0 and 1) diverged some time in the past and
+Suppose two population (labeled 0 and 1) diverged some time in the past and
 then admix in proportions $f$ and $1-f$.
@@
 
-This shows that additive genetic variance after admixture is equal to the
-additive variances in the source populations, weighted by their contributes,
-plus a term that depends on the divergence at loci contributing to the trait
-between the two populations, measured as $F_2$, weighted by the quadratic
-factor $2f(1-f)$.
-
-In general, $F_2$ at a given locus will depend on the demography connecting the
-two populations, as well as the effect size at the locus due to selection on
-the trait. In the infinitesimal limit, involving many loci each of vanishingly
-small effect, dynamics at a given locus will be approximately neutral, so that
-$F_2$ will depend only on the demography. In this case
-\begin{align*}
-    V_G & \approx f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] \sum_l a_l^2 \\
-    & = f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] n \sigma_m^2
-\end{align*}
-Here, $\sigma_m^2$ is the variance of effect sizes of new mutations (assuming mean is
-zero).
-
 \begin{figure}[ht!]

### 2023-06-27 doc/manuscript.tex
outline some results and discussion

@@
 
-\begin{figure}[ht!]
+\begin{figure}[t!]
     \centering
@@
 
-\begin{figure}[ht!]
+\begin{itemize}
+    \item In the demographic model more closely resembling inferred history
+        connecting humans and Neanderthals (Figure~\ref{fig:neand-to-human}A),
+        we see that demographic history and the distribution of effect sizes
+        of new mutations impact $V_G$
+    \item Additive genetic variance increases after introgression in both directions,
+        which is expected to result in a period of decreased mean fitness
+    \item Population contractions decrease $V_G$, as the increased rate of drift
+        reduces diversity at loci underlying the trait
+    \item TODO: compare to linked simulations
+\end{itemize}
+
+\begin{figure}[t!]
     \centering
@@
 
-\begin{figure}[ht!]
+\begin{itemize}
+    \item Because we know if a mutation was previously segregating in the population
+        receiving migrants, we can partition variation by contributions from
+        segregating vs introgressed variants (whether introduced derived mutations
+        or reintroduced ancestral mutations)
+    \item At low introgression proportions, as here, the majority of $V_G$ is still
+        contributed by previously segregating mutations, which is then expected to
+        be steadily replaced by new mutations (Figure 2). $V_G$ due to introgressed
+        variation decays monotonically over time (Figure~\ref{fig:neand-to-human}C,D)
+    \item While the total $V_G$ due to introgressed variants decays, we can also
+        consider the average contribution of introgressed vs non-introgressed SNPs
+        to $V_G$ ($h^2$ per SNP) -- at least for this demographic model and
+        DES, the contribution per SNP of introgressed variants is initially lower
+        than those of previously segregating variants. These contributions change
+        over time, in a way that depends on both the genomic architecture underlying
+        the trait and demography
+    \item This dependence on both architecture and demography is apparent when
+        comparing to $V_G$ contributed by introgressed and non-introgressed variants
+        in the early human-to-Neanderthal admixture (Figure~\ref{fig:human-to-neand})
+        -- here, again the contribution
+        from introgressed variants to $V_G$ decays monotonically, but depending on
+        the distribution of effect sizes, relative per-SNP $h^2$ changes over time
+        (though in these cases introgressed $h^2$ per SNP remains lower than
+        non-introgressed, when considering mutations at matched allele frequencies)
+\end{itemize}
+
+\subsection*{The effects of linkage}
+
+\begin{itemize}
+    \item So far, we have considered model and simulation without linkage -- in this
+        case, the underdominance model provides an excellent approximation
+    \item Including linkage between sites dramatically complicates things, making
+        theorical or analytical progress difficult
+    \item To explore the effects of linkage, we used individual-based simulations
+        with a single large chromosome of size 1 Morgan, and varried the total
+        mutation rate per haploid and the variance of effect sizes of new mutations
+    \item We'll highlight two effects of linkage
+        \begin{itemize}
+            \item Deviations from expected $V_G$, as well as more complicated
+                short-term dynamics after admixture
+            \item The reduction in introgressed ancestry in the regions surrounding
+                trait-affecting alleles
+        \end{itemize}
+\end{itemize}
+
+\subsubsection*{Effects of linkage on expected genetic variance}
+
+\begin{figure}[t!]
+    \centering
+    %\includegraphics{}
+    \caption{}
+    \label{fig:linkage-deviations}
+\end{figure}
+
+\begin{itemize}
+    \item Stabilizing selection reduces $V_G$ by selecting against minor alleles
+        (underdominance), but with linkage it can also reduce $V_G$ by selecting
+        for haplotypes that carry opposite effect alleles that cancel out -- this
+        creates LD between positive and negative effect alleles at linked loci,
+        known as the Bulmer effect \citep{}
+    \item Thus, $V_G$ in simulations with linkage deviate from predictions
+        assuming underdominance alone
+    \item The severity of this deviation depends on a number of factors: mean
+        recombination rate between trait-affecting loci, the total mutation rate,
+        the effect sizes (and distribution of effect sizes of new mutations)
+    \item In general, lower mutation rates imply lower total $V_G$, lower polygenicity,
+        and therefore higher recombination rates between trait-affecting alleles --
+        we therefore expect this deviation to be exacerbated with higher mutation
+        rates (all else being equal)
+\end{itemize}
+
+\subsubsection*{Reduction of introgressed ancestry surrounding an introgressed trait-affecting allele}
+
+\begin{itemize}
+    \item Because introgressed trait-affecting alleles are selected against as the
+        minor allele, linkage will also remove introgressed ancestry segments in the
+        regions surrounding the selected alleles
+    \item The expected reduction in introgressed ancestry proportions surrounding
+        functional alleles will depend on the effect size of the trait-affecting allele
+        and the probability of recombination surrounding that site
+    \item We can consider first a deterministic model of the dynamics of introgressed
+        allele frequencies at and around an introgressed functional allele -- this
+        model considers a single trait-affecting locus, ignoring any effects of
+        interference between selected alleles
+        \begin{itemize}
+            \item Consider a site with fixed differences between the two populations
+                (the derived allele is fixed in one, lost in the other)
+            \item At the functional site itself, with admixture proportion $f$ from the
+                minor parental ancestry, the initial frequency of the derived allele
+                is $p_0=f$ or $p_0 = 1-f$, if the derived allele was fixed in the 
+                major parental ancestry population
+            \item In one generation, the expected allele frequency at the selected locus
+                is \[p_{t+1} \approx p_t - s p_t(1-p_t)(1-2p_t),\] where $s=a^2/2V_S$
+            \item As discussed earlier, if $p_0<1/2$, $p_t\rightarrow0$, and if $p_0>1/2$,
+                $p_t\rightarrow1$ as $t\rightarrow\infty$
+            \item Consider a neutral locus separated from the selected locus by
+                recombination probability $r$. Initially, the expected frequency
+                of introgressed ancestry at that locus is $q_0=f$
+            \item The linked ancestry frequency will change due to selection acting on
+                the trait-affecting locus, depending on the strength of selection
+                at that locus and the level of linkage disequilibrium between
+                the two loci
+            \item $D=Cov(p,q)$ is the standard covariance measure of LD between loci,
+                and $q$ is expected to change as \[q_{t+1} \approx q_t - s D_t(1-2p_t)\]
+            \item $D$ also changes over time, due to both selection and recombination,
+                so that \[D_{t+1} \approx D_t - r D_t - s D_t (1-2p_t)^2\]
+            \item Initially, $D_0 = \pm f(1-f)$, with $D$ being positive if $p_0=f$
+                and negative if $p_0=1-f$
+            \item Together, this forms a nonlinear system of equations for the
+                deterministic change in allele frequencies at the two loci (one
+                selected, one neutral) and LD between them -- note that this is
+                deterministic in the infinite population size limit
+        \end{itemize}
+    \item Using this model, we can predict how allele frequencies and LD are expected
+        to change over time after admixture, for a given effect size $a$ and 
+        recombination rate $r$ (Figure~\ref{fig:linkage-pred}A,B)
+    \item As expected, lower effect sizes result in a slower reduction in introgressed
+        ancestry frequency, while higher recombination rates more quickly decouple
+        the linked ancestry from the selected allele
+    \item Thus, the expected reduction in introgressed ancestry is largest for
+        larger effect sizes (Figure~\ref{fig:linkage-pred}C), and LD between the
+        selected and linked neutral alleles is largest for neutral sites closest
+        to the selected allele (Figure~\ref{fig:linkage-pred}D)
+\end{itemize}
+
+\begin{figure}[t!]
     \centering
@@
         \textbf{}
+        \aprcomment{Linked allele frequency should always be minor, and labeled
+        introgressed ancestry frequency}
     }
@@
 
-\begin{figure}[ht!]
+\begin{itemize}
+    \item To explore the accuracy of this two-locus model, we compared to
+        simulations using the demographic model shown in
+        Figure~\ref{fig:toy-admixture}A with introgression fraction $f=0.05$.
+        We simulated a single 1 Morgan
+        chromosome, with all mutations having effect sizes $\pm a$ ($a=0.02$ or
+        $0.05$, with equal probability of being trait-increasing or decreasing)
+        and varying the total mutation rate ($\mu=0.001$, $0.0025$, and $0.01$)
+    \item For each mutation that was fixed in one population and absent in the
+        other (either fixed in the minor or major ancestry population), we
+        determined the average introgressed ancestry surrounding those introgressed
+        alleles (Figure~\ref{fig:linkage-sim})
+    \item The deterministic model provides a very good approximation for linked
+        introgressed ancestry when the mutation rate is small
+    \item As the mutation rate increases, polygenicity increases so that
+        trait-affecting alleles are more densely distributed along the chromosome --
+        deviations from the deterministic model are due to interference between
+        tightly linked trait-affecting alleles. For the highest mutation rate shown
+        here, there will be many trait-affecting alleles within the 1 cM window
+        around any focal SNP
+\end{itemize}
+
+\begin{figure}[t!]
     \centering
@@
 
-\clearpage
+\section*{Discussion}
+
+\begin{itemize}
+    \item Stabilizing selection with a constant optimum is an appropriate
+        null model for the dynamics of trait variation and selection on
+        alleles underlying those traits
+    \item Stabilizing selection results in population genetic predictions
+        for patterns of introgressed variation in and around functional
+        regions that differ from both load- and incompatibility-based arguments
+        \begin{itemize}
+            \item Load-based models predict that haplotypes from the less ``loaded''
+                population will be favored under introgression in either direction
+                -- thus, regions of depleted ancestry after introgression in one
+                direction will not be expected to also be introgressed ancestry
+                deserts after admixture in the opposite direction
+            \item \citet{mueller1942} originally pointed out that hybrid
+                incompatibilities should not form at single loci -- classical
+                incompatibilities require two or more loci, which are not
+                expected to be tightly linked to one another. Thus, under a
+                model of hybrid incompatibilities, the minor parent ancestry
+                is again selected against, but those loci will be in different
+                regions. Again, deserts are therefore not expected to overlap
+                if selection is largely due to incompatibilities
+            \item Instead, stabilizing selection causes introgressed DNA to be
+                depleted at the same loci under bi-directional gene flow
+                (could be thought of as single-locus incompatibility effects --
+                but these are not incompatibilities in the classical sense,
+                as the same effects occur even without admixture)
+            \item A prediction we could make is that, if current models are correct
+                that there has been introgression in both directions between
+                \emph{H. sapiens} and Neanderthals, introgressed ancestry deserts
+                should overlap with each other more often than expected by chance
+        \end{itemize}
+    \item Some conjectures about hybrid zones
+\end{itemize}
+
+\section*{Methods}
+
+\subsection*{Computing expected genetic variance using the diffusion approximation}
+
+\subsection*{Simulations with free recombination}
+
+\subsection*{Individual-based simulations with linkage}
 
@@
 
-\begin{figure}[ht!]
+\begin{figure}[tb!]
     \centering
@@
 
-\subsection{The role of linkage}
-
-\section{Discussion}
-
-\section{Methods}
-
 \end{document}

### 2023-06-30 background_selection/bgs_functional_regions.py
bgs simulation script

@@ -1,0 +1,417 @@
+"""
+Run the fwdpy11 simulation and get ancestry proportions around introgressed
+alleles that were fixed differences between populations.
+"""
+
+
+import fwdpy11
+import tskit
+import numpy as np
+from dataclasses import dataclass
+from typing import List
+from collections import defaultdict
+import pickle
+import gzip
+import time
+import demes
+import sys
+
+
+def get_samples_dict(ts):
+    samples_dict = defaultdict(list)
+    for s in ts.samples():
+        samples_dict[(ts.node(s).time, ts.node(s).population)].append(s)
+    return samples_dict
+
+
+def get_ancestry_counts(ts):
+    # preprocess and set up storage arrays
+    samples_dict = get_samples_dict(ts)
+
+    max_t = max([k[0] for k in samples_dict.keys()])
+    ts2 = ts.decapitate(max_t)
+
+    num_samples = len(samples_dict[min(samples_dict.keys())])
+
+    sample_counts = np.zeros(ts2.num_nodes, dtype=int)
+    parents = -np.ones(ts2.num_nodes, dtype=int)
+
+    for s in samples_dict[min(samples_dict.keys())]:
+        sample_counts[s] += 1
+
+    num_seg = np.zeros(ts2.num_trees)
+    num_intro = np.zeros(ts2.num_trees)
+
+    for i, ed in enumerate(ts2.edge_diffs()):
+        interval, edges_out, edges_in = ed
+        # take care of edges out
+        for e_out in edges_out:
+            c = e_out.child
+            p = e_out.parent
+            assert parents[c] == p
+            parents[c] = -1
+            sample_counts[p] -= sample_counts[c]
+            while parents[p] != -1:
+                c = p
+                p = parents[c]
+                sample_counts[p] -= sample_counts[e_out.child]
+
+        # update storage with edges in
+        for e_in in edges_in:
+            c = e_in.child
+            p = e_in.parent
+            assert parents[c] == -1
+            parents[c] = p
+            sample_counts[p] += sample_counts[c]
+            while parents[p] != -1:
+                c = p
+                p = parents[c]
+                sample_counts[p] += sample_counts[e_in.child]
+        # count up ancestry proportions in this segment
[…]

### 2023-06-30 background_selection/bgs_functional_regions.py
fix up bgs script

@@ -109,9 +109,7 @@ def set_up_demographic_model(Ne=10000):
 class SimData:
     generation: int
     demes_ids: List[int]
-    mean_phenotype: List[float]
     mean_fitness: List[float]
-    var_phenotype: List[float]
 
 
 @dataclass
@@ -125,11 +123,9 @@ class Recorder:
             md = np.array(pop.diploid_metadata)
             # store lists of mean phenotypes, fitnesses, etc
             deme_ids = sorted(list(set(md["deme"])))
-            mean_pheno = [md[md["deme"] == i]["g"].mean() for i in deme_ids]
             mean_fitness = [md[md["deme"] == i]["w"].mean() for i in deme_ids]
-            var_pheno = [md[md["deme"] == i]["g"].var() for i in deme_ids]
             self.data.append(
-                SimData(pop.generation, deme_ids, mean_pheno, mean_fitness, var_pheno)
+                SimData(pop.generation, deme_ids, mean_fitness)
             )
         if pop.generation % 1000 == 0:
             print(f"{time.ctime()}, at generation {pop.generation}")
@@ -170,7 +166,7 @@ def run_sim(
         "gvalue": fwdpy11.Multiplicative(scaling=2),
         "simlen": simlen,
         "demography": model,
-        "prune_selected": False,
+        "prune_selected": True,
     }
     params = fwdpy11.ModelParams(**pdict)
 
@@ -196,29 +192,21 @@ def run_sim(
 
 def recorder_to_dict(recorder, g):
     gens0 = len(recorder.data)
-    mp0 = np.zeros(gens0, dtype="float32")
     mf0 = np.zeros(gens0, dtype="float32")
-    vp0 = np.zeros(gens0, dtype="float32")
     gens1 = int(
         (g.demes[1].start_time - g.demes[1].epochs[0].end_time) / g.generation_time
     )
-    mp1 = np.zeros(gens1, dtype="float32")
     mf1 = np.zeros(gens1, dtype="float32")
-    vp1 = np.zeros(gens1, dtype="float32")
     j = 0
     for i, d in enumerate(recorder.data):
-        mp0[i] = d.mean_phenotype[0]
         mf0[i] = d.mean_fitness[0]
-        vp0[i] = d.var_phenotype[0]
         if 1 in d.demes_ids:
-            mp1[j] = d.mean_phenotype[1]
             mf1[j] = d.mean_fitness[1]
[…]

### 2023-07-05 doc/manuscript.tex
update some figures

@@
 Genomic studies have shown that admixture is a common occurrance between
-diverged populations and closely related taxa \citep{}, and potentially has
-been widespread in primate \citep{} and hominin \citep{} evolution.
-Understanding the effects of admixture on phenotypic and molecular variation is
-therefore relevant to many natural systems, including to studies of the genetic
-basis of complex traits and diseases. In particular, archaic introgression in
-\emph{Homo sapiens} involving Neanderthals and Denisovans has recently
-attracted considerable attention, both in understanding the historical
-processes leading to observed distributions of introgressed DNA in present-day
-populations \citep{prufer2014complete,villanea20??,chen2020identifying} as well
-as its contribution to quantitative traits \citep{sankararaman,wei,others}.
+diverged populations and closely related taxa \citep{brandvain2014speciation,
+skoglund2015ancient,suvorov2022widespread}, and potentially has been widespread
+in primate \citep{tung2017contribution,sorensen2023genome} and hominin
+\citep{wolf2018outstanding,peter2020100} evolution. Understanding the effects
+of admixture on phenotypic and molecular variation is therefore relevant to
+many natural systems, including to studies of the genetic basis of complex
+traits and diseases. In particular, archaic introgression in \emph{Homo
+sapiens} involving Neanderthals and Denisovans has recently attracted
+considerable attention, both in understanding the historical processes leading
+to observed distributions of introgressed DNA in present-day populations
+\citep{prufer2014complete,villanea2019multiple,chen2020identifying} as well as
+its contribution to quantitative traits
+\citep{sankararaman2016combined,wei2023lingering}.
 
 After their introduction through admixture, introgressed alleles may be
-selected either for or against. Some introgressed haplotypes appear to have
-been positively selected in modern humans
-\citep{huerta2014altitude,gower2021detecting,Enard}, possibly because such
-alleles are locally adaptive and provide a fitness advantage when encountering
-a novel environment. Despite these cases of adaptive introgression,
-introgressed alleles were more likely to have been negatively selected in
-modern humans \citep{harris2016genetic,juric2016strength}. As Neanderthal and
-Denisovan population sizes are thought to have been relatively small for
-hundreds of thousands of years, population genetics theory predicts that they
-would accumulate deleterious variation at a faster rate than larger
-populations.  Introgressed haplotypes would then carry more deleterious
-variants, which would quickly be selected against after admixture. In mapping
-the distribution of haplotypes identified as introgressed from Neanderthals,
-there is a reduction of Neanderthal-related ancestry in enhancers and
-regulatory regions
+selected for or against. Some introgressed haplotypes appear to have been
+positively selected in modern humans
+\citep{huerta2014altitude,racimo2017signatures,enard2018evidence,gower2021detecting},
+possibly because such alleles are locally adaptive and provide a fitness
+advantage when encountering a novel environment. Despite these cases of
+adaptive introgression, introgressed alleles were more likely to have been
+negatively selected in modern humans
+\citep{harris2016genetic,juric2016strength}. As Neanderthal and Denisovan
+population sizes are thought to have been relatively small for hundreds of
+thousands of years, theory predicts that they would accumulate deleterious
+variation at a faster rate than larger populations. Introgressed haplotypes
+would then carry more deleterious variants, which would quickly be selected
+against after admixture. \aprcomment{Double check the following statements.} In
+mapping the distribution of haplotypes identified as introgressed from
+Neanderthals, there is a reduction of Neanderthal-related ancestry in enhancers
+and regulatory regions
 \citep{petr2019limits,telis2020selection,yermakovich2023long}. This
@@
 introgression in the reverse direction occurring tens or hundreds of thousands
-of years before the Neanderthal-to-human event surrounding the out-of-Africa
-expansion around 60 thousand years ago (ka). This is supported by early ``near
-modern'' \emph{H.  sapiens} outside of African around 120--100ka or earlier
+of years before the Neanderthal-to-human event during the global dispersal of
+later humans. This is supported by ``near modern'' \emph{H.  sapiens} outside
+of African around 120--100ka or earlier
 \citep{schwarcz1988esr,grun2005u,beyer2021climatic}, potentially overlapping
@@
 estimates of the genomic contribution of early \emph{H. sapiens} to
-Neanderthals vary \citep{}, up to XX\% of later Neanderthal genomes may have
-been contributed through introgression from humans to Neanderthals. Under the
-same argument that \emph{H. sapiens}-related haplotypes carry fewer deleterious
-alleles, human-introgressed DNA may have been favored in Neanderthal genomes.
-The replacement of Neanderthal mitochondrial and Y chromosomes by early human
+Neanderthals vary \citep{kuhlwilm2016ancient,hubisz2020mapping}, up to 3--5\%
+of later Neanderthal genomes may have been contributed through introgression
+from humans to Neanderthals. Under the reasoning that \emph{H.
+sapiens}-related haplotypes carried fewer deleterious alleles,
+human-introgressed DNA would have been favored in Neanderthal genomes. The
+replacement of Neanderthal mitochondrial and Y chromosomes by early human
 haplotypes supports this model \citep{posth2017deeply,petr2020evolutionary}.
 
-Models for selection on introgressed alleles are often based on load arguments
-(in particular, differences in the rate of accumulation of unconditionally
-deleterious variation in populations of different sizes) or hybrid
-incompatibilities \citep{mueller,etc}. These arguments are founded primarily in
-population genetics theory, eschewing any notion of genetic variation
-contributing to phenotypic variation and selection operating on those
-phenotypes. However, most phenotypic traits are thought to be under stabilizing
-selection \citep{sanjak2018evidence}, including gene regulation
-\citep{gilad2006natural,hodgins2015gene,price2022detecting}. This is especially
-relevant for selection on Neanderthal-introduced alleles in modern humans, as
+Models for selection against introgressed alleles are often based on load
+arguments (in particular, differences in the rate of accumulation of
+unconditionally deleterious variation in populations of different sizes) or
+hybrid incompatibilities \citep{muller1942isolating}. These arguments are
+founded primarily in population genetics theory, with little notion of genetic
+variation contributing to phenotypic variation that selection operates on.
+Most phenotypic traits are thought to be under stabilizing selection
+\citep{sanjak2018evidence,sella2019thinking}, including gene regulation
+\citep{gilad2006natural,hodgins2015gene,price2022detecting}. This is relevant
+to selection on Neanderthal-introduced alleles in modern humans, as the some of
 the strongest signals of selection against such alleles occur in regulatory
@@
 
-In two diverged populations, the genomic architecture (i.e., mutations
-underlying the trait) of a trait under stabilizing selection has a higher rate
-of turnover compared to neutrally evolving loci \citep{yair2022population}. We
-might therefore expect an accumulation of fixed differences between the
-\emph{H. sapiens} and Neanderthal branches for alleles contributing to a trait,
-even when the mean phenotype in each population is close to the same trait
-optimum. When alleles that are fixed in one population are introduced to
-another population in which they were previously absent, they will be at low
-frequency and subsequently selected against. Likewise, if the ancestral allele
-is reintroduced to a population fixed for a derived allele, the ancestral
-allele will be at low frequency and will be selected against. In either case we
-should expect the introgressed allele, whether ancestral or derived, to be
-selected against, regardless the historical relative sizes of the populations
-involved. This prediction contrasts with the population-genetics ``load''
-model, in which haplotypes with fewer deleterious variants (such as those from
-the population with larger historical size) are favored after introgression in
-either direction.
+In two diverged populations, the genomic architecture of a trait under
+stabilizing selection has a higher rate of turnover compared to neutrally
+evolving loci \citep{yair2022population}. We might therefore expect an
+accumulation of fixed differences between the \emph{H. sapiens} and Neanderthal
+branches for alleles contributing to a trait, even when the mean phenotype in
+each population is close to the same trait optimum. When alleles that are fixed
+in one population are introduced to another population in which they were
+previously absent, they will be at low frequency and subsequently selected
+against. Likewise, if the ancestral allele is reintroduced to a population
+fixed for a derived allele, the ancestral allele will be at low frequency and
+will be selected against. In either case we should expect the introgressed
+allele, whether ancestral or derived, to be selected against, regardless the
+historical relative sizes of the populations involved. This prediction
+contrasts with the population-genetics ``load'' model, in which haplotypes with
+fewer deleterious variants (such as those from the population with larger
+historical size) are favored after introgression in either direction.
 
@@
 approximated with the recursion
-\[V_{G,t+1} \approx V_{G,t}\left(1-a^2/2(V_S+V_G)\right)
+\[V_{G,t+1} \approx V_{G,t}\left(1-a^2/2(V_S+V_{G,t})\right)
 \left(1-1/2N_e\right) + 2 \mu a^2,\]
 where $\mu$ is the per-haploid, per-generation rate of mutation. In the
-large-population-size limit, this gives the well-known results
+large-population-size limit, this gives the well-known result
 \[\tilde{V_G} \approx 4\mu V_S,\] provided \(V_G \ll V_S\).
@@
 We model the distribution of allele frequencies underlying the trait using
-the approximation that alleles are subject to symmetric underdominant selection.
-For an allele with effect size $a$ (in the heterozygous state -- homozygotes
-contribute $2a$ to the trait), the selection coefficient
+the approximation that alleles are subject to symmetric underdominant
+selection \citep{robertson1956effect,keightley1988quantitative}.
+For an allele with (heterozygote) additive effect size $a$,
+the selection coefficient
 \[s\approx \frac{a^2}{2(V_S + V_G)} \approx \frac{a^2}{2V_S},\]
@@
 \[\E[V_{\delta_p}] \approx \frac{1}{4N_e}p(1-p).\]
-From this, because $s$ is always positive for any $a\not=0$, we see that
-selection pushes allele frequencies to zero if $p<1/2$ and to one if $p>1/2$.
-$p=1/2$ is an unstable equilibrium.
+%From this, because $s$ is always positive for any $a\not=0$, we see that
+%selection pushes allele frequencies to zero if $p<1/2$ and to one if $p>1/2$.
+%$p=1/2$ is an unstable equilibrium.
 
@@
 expected pairwise diversity from the SFS (with sample size $n$, $\Phi_n$), as
-\(V_{G,a} = 2a^2\sum_{j=1}^{n-1}\frac{j(n-j)}{n(n-1)}\Phi_n(j|a,\mu_a)da\).
+\[V_{G,a} = 2a^2\sum_{j=1}^{n-1}\frac{j(n-j)}{n(n-1)}\Phi_n(j|a,\mu_a)da.\]
 Here, $\mu_a$ is the mutation rate of alleles with effect size $a$. The total
-genetic variance is then \(V_G=\int_{-\infty}^\infty V_{G,a}
-\mathcal{N}(0,V_M).\)
+genetic variance is then 
+\[V_G=\int_{-\infty}^\infty V_{G,a}\mathcal{N}(0,V_M).\]
 
@@
 histories, population splits, continuous migration, and admixture. In this
-study, we consider relativey simple scenarios involving population splits with
+study, we consider relatively simple scenarios involving population splits with
 subsequent introgression events. We focus on parameter regimes relevant to
@@
 trait optimum is $0$ in all branches, and the strength of selection $V_S=1$
-remains constant.
+remains constant. \aprcomment{Justified by equivalence under rescaling
+\citep{simons2018population}.}
 
@@
 p_1)^2$ is the squared difference in allele frequencies at a locus
-\citep[e.g.,][]{peter2016admixture}. This result in known \citep[e.g.,][]{},
-showing that additive variance is equal to that in the source populations
-weighted by their contributions, plus a term that depends on the divergence at
-trait-affecting loci between the populations weighted by the quadratic factor
-$2f(1-f)$.
+\citep[e.g.,][]{peter2016admixture}. This result is known
+\citep[e.g.,][]{tufto2000quantitative}, showing that additive variance is equal
+to that in the source populations weighted by their contributions, plus a term
+that depends on the divergence at trait-affecting loci between the populations
+weighted by the quadratic factor $2f(1-f)$.
 
@@
     \centering
-    %\includegraphics{}
+    \includegraphics{../figures/model_comparison.SD_0.01.mu_0.001_0.025}
     \caption{}
@@
 
+\begin{figure}[t!]
+    \centering
+    \includegraphics{../figures/introgression_deserts.SD_0.02.pdf}
+    \caption{
+        \textbf{}
+    }
+    \label{fig:deserts}
+\end{figure}
+
 \section*{Discussion}
@@
                 should overlap with each other more often than expected by chance
+            \item See Figure~\ref{fig:ancestry-deserts}
         \end{itemize}
+    \item Shared deserts are most likely to form where moderate- or large-effect
+        alleles have fixed in one population or the other -- this could provide
+        an approach to map\dots?
+    \item Selection in structured populations with ongoing migration or
+        meta-populations -- SFS approach can handle this -- prospects for the
+        future
     \item Some conjectures about hybrid zones
+    \item Pleiotropy\dots!!!
 \end{itemize}

### 2023-07-05 figures/linkage_reduction_simulation.py

@@ -88,7 +88,7 @@ if __name__ == "__main__":
         y = fin["arr_0"]
         ax2.plot(rs, y, "-", lw=1, label=f"$\mu={mu}$")
 
-    #ax2.legend(frameon=False)
+    ax2.legend(frameon=False)
     ax2.set_xlabel("cM")
     ax2.set_ylabel("Introgressed ancestry proportion")
     ax2.set_ylim(0, 0.055)

### 2023-07-05 unlinked/turelli-to-lande.py

@@ -1,0 +1,113 @@
+import numpy as np
+import pickle
+import sys
+
+
+def get_fprime(freqs, es, optimum, VS):
+    two_f_a = 2 * freqs * es
+    VG_contributions = 2 * freqs * (1 - freqs) * es * es
+    EG = two_f_a.sum()
+    VG = VG_contributions.sum()
+    VGprime = VG - VG_contributions
+    EGprime = EG - two_f_a
+
+    delta_p = (
+        freqs
+        * (1 - freqs)
+        * np.sqrt((VG + VS) / (VGprime + VS))
+        * np.exp(((EG - optimum) ** 2) / (2 * (VG + VS)))  # nearly 1
+        * (  # also nearly 1
+            freqs * np.exp(-((EGprime + 2 * es - optimum) ** 2) / (2 * (VGprime + VS)))
+            - (1 - freqs) * np.exp(-((EGprime - optimum) ** 2) / (2 * (VGprime + VS)))
+            + (1 - 2 * freqs)
+            * np.exp(-((EGprime + es - optimum) ** 2) / (2 * (VGprime + VS)))
+        )
+    )
+    return freqs + delta_p
+
+
+def sample_generation(Ne, fprime):
+    return np.random.binomial(2 * Ne, p=fprime) / 2 / Ne
+
+
+def new_mutations(freqs, es, ss, mu, Ne):
+    num_muts = np.random.poisson(2 * Ne * mu)
+    extension = num_muts - (ss == 0).sum()
+    if extension > 0:
+        freqs = np.concatenate((freqs, np.zeros(extension)))
+        es = np.concatenate((es, np.zeros(extension)))
+        ss = np.concatenate((ss, np.zeros(extension, dtype=int)))
+    free_sites = np.where(ss == 0)[0]
+    for i in range(num_muts):
+        site_idx = free_sites[i]
+        # add to G
+        freqs[site_idx] = 1 / 2 / Ne
+        # add to seg sites
+        ss[site_idx] = 1
+        # add effect size
+        es[site_idx] = np.random.randn() * SD
+    return freqs, es, ss
+
+
+def evolve(freqs, es, ss, SD, optimum, VS, mu, Ne):
+    # get marginal fitnesses of alleles at all segregating sites
+    fprime = get_fprime(freqs, es, optimum, VS)
+
+    # binomial sampling of alleles to create offspring
[…]

### 2023-09-21 figures/r35-summary.py
r35 figures

@@ -1,0 +1,171 @@
+import numpy as np
+import matplotlib.pylab as plt, matplotlib
+import moments
+import demes, demesdraw
+import scipy.stats
+import os, pickle
+import sys
+
+
+from single_population_stabilizing_selection import *
+
+
+def evolve_generation(p, q, D, r, s):
+    # with underdominance (s > 0)
+    p2 = p - s * p * (1 - p) * (1 - 2 * p)
+    q2 = q - s * D * (1 - 2 * p)
+    D2 = D - s * D * (1 - 2 * p) ** 2 - r * D
+    return p2, q2, D2
+
+
+def trajectories(p0, q0, D0, generations=1, r=0, s=0):
+    p = [p0]
+    q = [q0]
+    D = [D0]
+    for g in range(generations):
+        p2, q2, D2 = evolve_generation(p[-1], q[-1], D[-1], r, s)
+        p.append(p2)
+        q.append(q2)
+        D.append(D2)
+    return p, q, D
+
+
+def frequency_dip(a_vals, r_max, gens, p0, ax_q, VS=1, legend=True):
+    rs = np.linspace(0, r_max, 101)
+    rs_plot = 100 * np.concatenate((-rs[:0:-1], rs))
+    for a in a_vals:
+        qs = np.zeros(len(rs))
+        s = a**2 / 2 / VS
+        for i, r_dist in enumerate(rs):
+            p, q, D = trajectories(p0, p0, p0 - p0**2, generations=gens, r=r_dist, s=s)
+            qs[i] = q[-1]
+        qs = np.concatenate((qs[:0:-1], qs))
+        ax_q.plot(rs_plot, qs, label=f"$a={a}$")
+    if legend:
+        ax_q.legend(frameon=False, title="Effect size")
+    ax_q.set_ylim(bottom=0)
+    ax_q.set_xlabel("Distance from selected locus (cM)")
+    ax_q.set_ylabel("Introgressed ancestry")
+
+
+
+
+# set font sizes
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
[…]

### 2023-09-21 figures/r35-summary.py
labels

@@ -54,8 +54,8 @@ def frequency_dip(a_vals, r_max, gens, p0, ax_q, VS=1, legend=True):
 plt.rcParams["legend.title_fontsize"] = "xx-small"
 matplotlib.rc("xtick", labelsize=6)
 matplotlib.rc("ytick", labelsize=6)
-matplotlib.rc("axes", labelsize=7)
-matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("axes", labelsize=6)
+matplotlib.rc("axes", titlesize=6)
 matplotlib.rc("legend", fontsize=6)
 
 
@@ -167,5 +167,9 @@ frequency_dip([0.01, 0.02, 0.05], 1e-2, 2000, 0.05, ax3)
 
 plt.tight_layout()
 plt.subplots_adjust(hspace=0.4, top=0.98, bottom=0.12, left=0.1, right=0.99)
+
+fig.text(0.02, 0.97, "A", fontsize=7, va="center", ha="center")
+fig.text(0.33, 0.97, "B", fontsize=7, va="center", ha="center")
+fig.text(0.33, 0.49, "C", fontsize=7, va="center", ha="center")
 plt.savefig("r35-summary.pdf")
 

### 2023-09-22 figures/r35-summary.py
update figure

@@ -35,9 +35,11 @@ def frequency_dip(a_vals, r_max, gens, p0, ax_q, VS=1, legend=True):
     rs_plot = 100 * np.concatenate((-rs[:0:-1], rs))
     for a in a_vals:
         qs = np.zeros(len(rs))
-        s = a**2 / 2 / VS
+        s = a ** 2 / 2 / VS
         for i, r_dist in enumerate(rs):
-            p, q, D = trajectories(p0, p0, p0 - p0**2, generations=gens, r=r_dist, s=s)
+            p, q, D = trajectories(
+                p0, p0, p0 - p0 ** 2, generations=gens, r=r_dist, s=s
+            )
             qs[i] = q[-1]
         qs = np.concatenate((qs[:0:-1], qs))
         ax_q.plot(rs_plot, qs, label=f"$a={a}$")
@@ -46,8 +48,6 @@ def frequency_dip(a_vals, r_max, gens, p0, ax_q, VS=1, legend=True):
     ax_q.set_ylim(bottom=0)
     ax_q.set_xlabel("Distance from selected locus (cM)")
     ax_q.set_ylabel("Introgressed ancestry")
-
-
 
 
 # set font sizes
@@ -165,6 +165,15 @@ ax2.text(700, 0.041, "$\sigma_M=0.01$", va="center", ha="center", fontsize=6)
 ax3 = plt.subplot2grid(grid, (1, 2), colspan=4)
 frequency_dip([0.01, 0.02, 0.05], 1e-2, 2000, 0.05, ax3)
 
+ax3.text(
+    0.55,
+    0.01,
+    "Introgressed alleles are\nselected against",
+    fontsize=6,
+    va="center",
+    ha="center",
+)
+
 plt.tight_layout()
 plt.subplots_adjust(hspace=0.4, top=0.98, bottom=0.12, left=0.1, right=0.99)
 
@@ -172,4 +181,3 @@ fig.text(0.02, 0.97, "A", fontsize=7, va="center", ha="center")
 fig.text(0.33, 0.97, "B", fontsize=7, va="center", ha="center")
 fig.text(0.33, 0.49, "C", fontsize=7, va="center", ha="center")
 plt.savefig("r35-summary.pdf")
-

### 2023-09-27 doc/manuscript.tex
update figure
edits to intro

@@
 
+\usepackage{setspace}
+
 \usepackage{xr}
 \externaldocument{supplement}
+
+\DeclareMathOperator{\Var}{Var}
 
@@
 
-\title{Stabilizing selection, archaic introgression, and the distribution
-of shared functional variation}
+\title{Archaic introgression and the distribution
+of shared functional variation under stabilizing selection}
 \author{APR}
@@
 
+\onehalfspacing
+
 \section*{Introduction}
 
-Genomic studies have shown that admixture is a common occurrance between
-diverged populations and closely related taxa \citep{brandvain2014speciation,
-skoglund2015ancient,suvorov2022widespread}, and potentially has been widespread
-in primate \citep{tung2017contribution,sorensen2023genome} and hominin
-\citep{wolf2018outstanding,peter2020100} evolution. Understanding the effects
-of admixture on phenotypic and molecular variation is therefore relevant to
-many natural systems, including to studies of the genetic basis of complex
-traits and diseases. In particular, archaic introgression in \emph{Homo
-sapiens} involving Neanderthals and Denisovans has recently attracted
-considerable attention, both in understanding the historical processes leading
-to observed distributions of introgressed DNA in present-day populations
-\citep{prufer2014complete,villanea2019multiple,chen2020identifying} as well as
-its contribution to quantitative traits
+Genomic surveys of natural systems suggest that historical admixture among
+diverged populations or closely related taxa commonly occors
+\citep{brandvain2014speciation,skoglund2015ancient,suvorov2022widespread} and
+is potentially widespread in primate
+\citep{tung2017contribution,sorensen2023genome} and hominin
+\citep{wolf2018outstanding,peter2020100} evolution. Admixture may therefore be
+an important driver of phenotypic and molecular variation and shape the genetic
+basis of complex traits. In humans, archaic introgression involving
+Neanderthals and Denisovans has attracted considerable attention, with efforts
+to describe the historical processes leading to observed distributions of
+introgressed DNA in present-day populations
+\citep{prufer2014complete,villanea2019multiple,chen2020identifying} and its
+contribution to quantitative traits
 \citep{sankararaman2016combined,wei2023lingering}.
 
-After their introduction through admixture, introgressed alleles may be
-selected for or against. Some introgressed haplotypes appear to have been
-positively selected in modern humans
+Once introduced through admixture, introgressed alleles may be selected for or
+against. Some introgressed haplotypes were positively selected in modern humans
 \citep{huerta2014altitude,racimo2017signatures,enard2018evidence,gower2021detecting},
-possibly because such alleles are locally adaptive and provide a fitness
-advantage when encountering a novel environment. Despite these cases of
-adaptive introgression, introgressed alleles were more likely to have been
-negatively selected in modern humans
-\citep{harris2016genetic,juric2016strength}. As Neanderthal and Denisovan
-population sizes are thought to have been relatively small for hundreds of
-thousands of years, theory predicts that they would accumulate deleterious
-variation at a faster rate than larger populations. Introgressed haplotypes
-would then carry more deleterious variants, which would quickly be selected
-against after admixture. \aprcomment{Double check the following statements.} In
-mapping the distribution of haplotypes identified as introgressed from
-Neanderthals, there is a reduction of Neanderthal-related ancestry in enhancers
-and regulatory regions
-\citep{petr2019limits,telis2020selection,yermakovich2023long}. This
-Neanderthal-allelic depletion, or ``deserts,'' support the hypothesis that
-introgressed functional alleles were selected against
+possibly due to locally adaptive variation that provided fitness advantages as
+humans encountered novel environments. Despite some cases of adaptive
+introgression, introduced alleles were more likely to have been selected
+against in modern humans \citep{harris2016genetic,juric2016strength}. Since
+Neanderthal and Denisovan population sizes were relatively small for hundreds
+of thousands of years, theory predicts they would have accumulated deleterious
+variants at an increased rate. Introgressed haplotypes loaded with more
+deleterious mutations would then have quickly been selected against after
+admixture. Mapping the distribution of Neanderthal-introgressed haplotypes in
+humans shows a reduction of Neanderthal-related ancestry in enhancers and
+regulatory regions
+\citep{petr2019limits,telis2020selection,yermakovich2023long}. These
+``deserts'' of Neanderthal ancestry support the hypothesis that introgressed
+functional alleles were selected against
 \citep{sankararaman2014genomic,sankararaman2016combined}.
 
-There is genetic evidence that early \emph{H. sapiens} reciprocally contributed
-to Neanderthal genomes \citep{kuhlwilm2016ancient,hubisz2020mapping}, with
-introgression in the reverse direction occurring tens or hundreds of thousands
-of years before the Neanderthal-to-human event during the global dispersal of
-later humans. This is supported by ``near modern'' \emph{H.  sapiens} outside
-of African around 120--100ka or earlier
+There is growing genetic evidence that early \emph{H. sapiens} reciprocally
+contributed to Neanderthal genomes
+\citep{kuhlwilm2016ancient,hubisz2020mapping,harris2023diverse}. This
+human-to-Neanderthal introgression would have occured tens or hundreds of
+thousands of years prior to Neanderthal introgression in humans during modern
+human global dispersal around 60 ka. This is supported by ``near modern''
+\emph{H. sapiens} outside of African around 120--100ka or earlier
 \citep{schwarcz1988esr,grun2005u,beyer2021climatic}, potentially overlapping
@@
 estimates of the genomic contribution of early \emph{H. sapiens} to
-Neanderthals vary \citep{kuhlwilm2016ancient,hubisz2020mapping}, up to 3--5\%
-of later Neanderthal genomes may have been contributed through introgression
-from humans to Neanderthals. Under the reasoning that \emph{H.
-sapiens}-related haplotypes carried fewer deleterious alleles,
-human-introgressed DNA would have been favored in Neanderthal genomes. The
-replacement of Neanderthal mitochondrial and Y chromosomes by early human
-haplotypes supports this model \citep{posth2017deeply,petr2020evolutionary}.
+Neanderthals vary, up to $\sim5\%$ of later Neanderthal genomes may have been
+contributed through introgression from humans to Neanderthals
+\citep{harris2023diverse}. If \emph{H. sapiens}-related haplotypes carried
+fewer deleterious alleles due to their larger long-term effective population
+size, human-introgressed DNA would have been favored in Neanderthal genomes.
+The replacement of Neanderthal mitochondrial and Y chromosomes by early human
+haplotypes supports such a model of post-admixture selection in the Neanderthal
+lineage \citep{posth2017deeply,petr2020evolutionary}.
 
@@
 unconditionally deleterious variation in populations of different sizes) or
-hybrid incompatibilities \citep{muller1942isolating}. These arguments are
-founded primarily in population genetics theory, with little notion of genetic
-variation contributing to phenotypic variation that selection operates on.
-Most phenotypic traits are thought to be under stabilizing selection
+hybrid incompatibilities \citep{muller1942isolating}. Such arguments, founded
+in population genetics theory, rarely take into account selection operating on
+phenotypic variation or the relationship between genetic and phenotypic
+variation. Many phenotypic traits are thought to be under stabilizing selection
 \citep{sanjak2018evidence,sella2019thinking}, including gene regulation
-\citep{gilad2006natural,hodgins2015gene,price2022detecting}. This is relevant
-to selection on Neanderthal-introduced alleles in modern humans, as the some of
-the strongest signals of selection against such alleles occur in regulatory
-regions \citep{sankararaman2014genomic}.
+\citep{gilad2006natural,hodgins2015gene,price2022detecting}. Because some of
+the strongest signals of selection against Neanderthal-introduced alleles are
+in regulatory regions \citep{sankararaman2014genomic}, stabilizing selection on
+quantitative traits is particularly relevant to the dynamics of functional
+genetic variation after introgression.
 
-Stabilizing selection on a trait acts to maintain the phenotypic distribution
-of that trait near some trait optimum, which is achieved by reducing phenotypic
-variation (Figure~\ref{fig:one-pop}A). When the mean phenotype of the
-population is close to the phenotypic optimum, classical models predict that
-alleles contributing to such a trait are subject to underdominant selection,
-i.e., selection against the minor allele at a locus
-\citep{robertson1956effect}. This has proven to be a useful model for
-understanding allele frequencies of variants contributing to a trait under
-stabilizing selection in the single-population setting
+Stabilizing selection acts to maintain the phenotypic distribution of a trait
+near some optimum, which is achieved by reducing phenotypic variation
+(Figure~\ref{fig:one-pop}A). When the mean phenotype of the population is close
+to the phenotypic optimum, classical models predict that alleles contributing
+to such a trait are subject to underdominant selection, i.e., selection against
+the minor allele at a locus \citep{robertson1956effect}. This has proven to be
+a useful model for understanding allele frequencies of variants contributing to
+a trait under stabilizing selection in single-population settings
 \citep[e.g.,][]{keightley1988quantitative, simons2018population,
@@
         Figures~\ref{fig:one-popA}--\ref{fig:one-popC}.}
+        \aprcomment{$w$ and $f$ need to match text}
     }
@@
 
-In two diverged populations, the genomic architecture of a trait under
-stabilizing selection has a higher rate of turnover compared to neutrally
-evolving loci \citep{yair2022population}. We might therefore expect an
-accumulation of fixed differences between the \emph{H. sapiens} and Neanderthal
-branches for alleles contributing to a trait, even when the mean phenotype in
-each population is close to the same trait optimum. When alleles that are fixed
-in one population are introduced to another population in which they were
-previously absent, they will be at low frequency and subsequently selected
-against. Likewise, if the ancestral allele is reintroduced to a population
-fixed for a derived allele, the ancestral allele will be at low frequency and
-will be selected against. In either case we should expect the introgressed
-allele, whether ancestral or derived, to be selected against, regardless the
-historical relative sizes of the populations involved. This prediction
-contrasts with the population-genetics ``load'' model, in which haplotypes with
-fewer deleterious variants (such as those from the population with larger
-historical size) are favored after introgression in either direction.
+In diverged populations, the genomic architecture of a trait under stabilizing
+selection has a higher rate of turnover compared to neutrally evolving loci
+\citep{yair2022population}. We might therefore expect an accumulation of fixed
+differences in the \emph{H. sapiens} and Neanderthal lineages at loci
+contributing to a trait, even when the mean phenotype in each population is
+close to the same trait optimum. When a derived allele at high frequency in one
+population is introduced to another population in which is was previously
+absent, it will be at low frequency and subsequently selected against.
+Likewise, if the ancestral allele is reintroduced to a population fixed for a
+derived allele, the ancestral allele will be at low frequency and will be
+selected against. In either case we should expect selection against the
+introgressed allele, whether ancestral or derived and regardless the historical
+relative sizes of the populations involved. This prediction contrasts with the
+population-genetics ``load'' model, in which haplotypes with fewer deleterious
+variants (such as those from the population with larger historical size) are
+favored after introgression in either direction.
 
@@
 dominance and epistatic effects, so that \(V_G = V_A\), assuming no linkage.
-We can further ingore environmental effects \citep{simons2018population}, so
-that the phenotypic variance \(V_P=V_A\).
+We further ingore environmental effects \citep{simons2018population}, so that
+the phenotypic variance \(V_P=V_A\).
 
 Stabilizing selection acts to reduce phenotypic variation around the optimum
-value $o$, and we assume a Gaussian fitness function so that relative fitness
-is given by \(f(G_i | O, V_S) = \exp{(-(G_i - O) / 2 V_S)}\). $V_S$ can be
-interpreted as the strength of selection on the trait, where larger $V_S$
-implies weaker selection. For a population with mean phenotype at (or very
-close to) the optimum, the mean fitness of the population (assuming a roughly
-normal distribution of phenotypic values in the population) is
-\[\bar{w} \approx \int_{-\infty}^\infty f(G | O, V_S) \mathcal{N}(0, V_G) dG
-\approx \left(\frac{V_S}{V_S+V_G}\right)^{1/2}.\]
-Thus, as the genetic variance increases, mean fitness among individuals in the
-population decreases.
+value $O$ (typically set to zero), and we assume a Gaussian fitness function so
+that relative fitness is given by \(f(G_i | O, V_S) = \exp{(-(G_i - O)^2 / 2
+V_S)}\). $V_S$ can be interpreted as the strength of selection on the trait,
+where larger $V_S$ implies weaker selection. For a population with mean
+phenotype at (or very close to) the optimum, the mean fitness of the population
+(assuming a roughly normal distribution of phenotypic values in the population)
+is \[\bar{w} \approx \int_{-\infty}^\infty f(G | 0, V_S) \mathcal{N}(0, V_G) dG
+= \left(\frac{V_S}{V_S+V_G}\right)^{1/2}.\] Thus, as the genetic variance
+increases, mean fitness among individuals in the population decreases.
 
@@
 
-If all alleles contribute equally to the trait (as \(\pm a\)),
+If all alleles contribute equally to the trait
+(with effect sizes \(\pm a\) occurring in equal proportion),
 \citet{keightley1988quantitative} showed that the dynamics of $V_G$ can be
@@
 Mutations are not expected to each have the same effect size $a$, but
-rather will be drawn from some distribution. Here, we will assume mutation
+rather will be drawn from some distribution. Throughout, we assume mutation
 effect sizes are drawn from a normal distribution with mean 0 and given
 variance $V_M$.
+
+\aprcomment{where does the $4\mu Ne V_M$ (Lande) result come from? point to fig 1}
 
@@
 We extend the moments-based solution for the sample site-frequency spectrum
-(SFS) \citep{jouganous2017} to include underdominance with given selection
-coefficient $s$ (\(=a^2/2(V_S+V_G)\)). The contribution of alleles with effect
-size $a$ to the total genetic variance $V_G$ is then found by computing the
-expected pairwise diversity from the SFS (with sample size $n$, $\Phi_n$), as
+(SFS) \citep{jouganous2017inferring} to include underdominance with given
+selection coefficient $s$ (\(=a^2/2(V_S+V_G)\)). The contribution of alleles
+with effect size $a$ to the total genetic variance $V_G$ is then found by
+computing the expected pairwise diversity from the SFS (with sample size $n$,
+$\Phi_n$), as
 \[V_{G,a} = 2a^2\sum_{j=1}^{n-1}\frac{j(n-j)}{n(n-1)}\Phi_n(j|a,\mu_a)da.\]
-Here, $\mu_a$ is the mutation rate of alleles with effect size $a$. The total
-genetic variance is then 
+Here, $\mu_a$ is the mutation rate of alleles with effect size $a$. Then
+assuming a normal distribution of effect sizes for new mutations, the total
+genetic variance is
 \[V_G=\int_{-\infty}^\infty V_{G,a}\mathcal{N}(0,V_M).\]
@@
 Our numerical solution for the SFS allows for non-constant population size
-histories, population splits, continuous migration, and admixture. In this
-study, we consider relatively simple scenarios involving population splits with
+histories, population splits, continuous migration, and admixture. Here, we
+consider relatively simple scenarios involving population splits with
 subsequent introgression events. We focus on parameter regimes relevant to
-human-Neanderthal history. In the simplest model -- which is not meant to
+human-Neanderthal history. In a simple toy model -- which is not meant to
 perfectly match any particular known or inferred history, but instead
 demonstrate the effect of reciprocal introgression following population
-divergence -- a population of size $N_e=10\,000$ splits into two, on remaining
+divergence -- a population of size $N_e=10\,000$ splits into two, one remaining
 size $10\,000$ and the other shrinking to size $1\,000$. They remain isolated
@@
 
+\aprcomment{Start with ``Here we show that\dots''}
+
 Suppose two population (labeled 0 and 1) diverged some time in the past and

### 2023-09-28 doc/manuscript.tex
edits

@@
         \aprcomment{$w$ and $f$ need to match text}
+        \aprcomment{Add the SHC model, and the ``vary SD'' fig as a panel here}
     }
@@
 
-\subsection*{Theory and model}
+\subsection*{Model}
 
@@
 Our numerical solution for the SFS allows for non-constant population size
-histories, population splits, continuous migration, and admixture. Here, we
+histories, population splits, continuous migration and admixture. Here, we
 consider relatively simple scenarios involving population splits with
@@
 contributing $5\%$ ancestry to the recipient population
-(Figure~\ref{toy-admixture}A,B).
+(Figure~\ref{fig:toy-admixture}A,B).
 
@@
 present time. Neanderthals contribute $2\%$ ancestry to this bottlenecked and
-expanding human population at 50ka, after which their branch ceases
+expanding human population at 50ka, after which they go extinct
 (Figure~\ref{fig:neand-to-human}A).
@@
 after admixture to understand allele frequency and haplotype dynamics. The
-trait optimum is $0$ in all branches, and the strength of selection $V_S=1$
-remains constant. \aprcomment{Justified by equivalence under rescaling
-\citep{simons2018population}.}
+trait optimum is kept at $0$ in all branches, and the strength of selection
+$V_S=1$ remains constant.
 
@@
 p_1)^2$ is the squared difference in allele frequencies at a locus
-\citep[e.g.,][]{peter2016admixture}. This result is known
+\citep{peter2016admixture}. This result is known
 \citep[e.g.,][]{tufto2000quantitative}, showing that additive variance is equal
@@
     V_G & \approx f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] \sum_l a_l^2 \\
-    & = f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] n_m V_M,
+    & = f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] L V_M,
 \end{align*}
 where $V_M$ is the variance of effect sizes of new mutations (assuming the mean
-is zero).
+is zero), and $L$ is the number of trait affecting loci.
 
-For $a \not\approx 0$, assuming neutral evolution will tend to overestimate
-$F_2$ compared to underdominant selection. We can compute expected $V_G$ both
-before and after admixture using the diffusion approximation for the joint SFS
-with underdominance. Comparing to simulations with free recombination between
-loci (Appendix~AY), we find that this provides an excellent fit to average
-observed $V_G$ over time (Figure~\ref{fig:toy-admixture}C,D). Following an
-initial rapid increase in $V_G$ due to admixture, it decays back to the
-steady-state expectation relatively rapidly.
+For $a \not\approx 0$, the assumption of neutral evolution will lead to
+overestimates of $F_2$ compared to underdominant selection. We can compute
+expected $V_G$ both before and after admixture using the diffusion
+approximation for the joint SFS with underdominance. Comparing to simulations
+with free recombination between loci (Appendix~AY), we find that this provides
+an excellent fit to average observed $V_G$ over time
+(Figure~\ref{fig:toy-admixture}C,D). In general, admixture causes a sudden
+increase in $V_G$ followed by a fairly rapid return to pre-admixture levels.
 
+Modeling the dynamics of genetic variance using the SFS lets us examine
+contributions to $V_G$ from different classes of mutations, such as those at
+different frequencies or arising at different times, and how those
+contributions change over time. Of interest is the contribution to $V_G$ from
+mutations that were already segregating in the recipient population, those that
+were introduced through introgression, and new mutations since the time of
+admixture.
+
+Paragraph on the following (point to appendix and supplement):
 \begin{itemize}

### 2023-09-28 figures/single_population_stabilizing_selection.py

@@ -2,6 +2,8 @@ import numpy as np
 import matplotlib.pylab as plt, matplotlib
 import moments
 import demes, demesdraw
+import scipy.stats
+import pickle
 
 # set font sizes
 plt.rcParams["legend.title_fontsize"] = "xx-small"
@@ -33,17 +35,18 @@ g = b.resolve()
 def get_VG(fs, a):
     return fs.pi() * a ** 2
 
+
 def get_gamma(a, VG, Ne, VS=1):
     s = a ** 2 / 2 / (VS + VG)
     gamma = -2 * Ne * s
     return gamma
 
+
 def init_fs(n, gamma, theta):
     return moments.Spectrum(
-        moments.LinearSystem_1D.steady_state_1D(
-            n, overdominance=gamma, theta=theta
-        )
+        moments.LinearSystem_1D.steady_state_1D(n, overdominance=gamma, theta=theta)
     )
+
 
 def get_time_series_VA(a, VS=1, mu=0.025, n=200, update_sel=False):
     # initial sfs
@@ -84,17 +87,54 @@ def get_time_series_VA(a, VS=1, mu=0.025, n=200, update_sel=False):
     VGs.insert(0, VGs[0])
     return t, VGs
 
-if __name__ == "__main__":
-    grid = (2, 2)
-    fig = plt.figure(1, figsize=(6.5, 3.5))
 
-    ax1 = plt.subplot2grid(grid, (0, 0), rowspan=2)
+def gaussian_des(a_list, VG_list, SD):
+    assert len(a_list) == len(VG_list)
+    S = 0
+    aa = np.concatenate(([0], a_list))
+    dxs = (aa - np.concatenate(([aa[0]], aa))[:-1]) / 2 + (
+        np.concatenate((aa, [aa[-1]]))[1:] - aa
+    ) / 2
+    weights = scipy.stats.norm.pdf(a_list, scale=SD)
+    for v, dx, w in zip(VG_list, dxs[1:], weights):
+        S += 2 * v * dx * w
+    c = 2 * scipy.stats.norm.cdf(-a_list[-1], scale=SD)
+    S += c * VG_list[-1]
+    return S
+
+
+def equilibrium_VG(SD):
+    EVG = 4 * mu * VS
+    gammas = [get_gamma(a, EVG, Ne, VS=VS) for a in a_list]
+
+    n = 500
[…]

### 2023-11-09 doc/manuscript.tex
updates to model and results

@@
 Genomic surveys of natural systems suggest that historical admixture among
-diverged populations or closely related taxa commonly occors
+diverged populations or closely related taxa commonly occurs
 \citep{brandvain2014speciation,skoglund2015ancient,suvorov2022widespread} and
@@
 \citep{kuhlwilm2016ancient,hubisz2020mapping,harris2023diverse}. This
-human-to-Neanderthal introgression would have occured tens or hundreds of
-thousands of years prior to Neanderthal introgression in humans during modern
-human global dispersal around 60 ka. This is supported by ``near modern''
-\emph{H. sapiens} outside of African around 120--100ka or earlier
+human-to-Neanderthal introgression occured tens or hundreds of thousands of
+years prior to Neanderthal introgression in humans during modern human global
+dispersal around 60 ka. This is supported by ``near modern'' \emph{H. sapiens}
+outside of African around 120--100ka or earlier
 \citep{schwarcz1988esr,grun2005u,beyer2021climatic}, potentially overlapping
@@
 estimates of the genomic contribution of early \emph{H. sapiens} to
-Neanderthals vary, up to $\sim5\%$ of later Neanderthal genomes may have been
+Neanderthals vary, around $6\%$ of later Neanderthal genomes may have been
 contributed through introgression from humans to Neanderthals
@@
 near some optimum, which is achieved by reducing phenotypic variation
-(Figure~\ref{fig:one-pop}A). When the mean phenotype of the population is close
+(Figure~\ref{fig:stab-sel}A). When the mean phenotype of the population is close
 to the phenotypic optimum, classical models predict that alleles contributing
@@
 the minor allele at a locus \citep{robertson1956effect}. This has proven to be
-a useful model for understanding allele frequencies of variants contributing to
-a trait under stabilizing selection in single-population settings
+a useful model for understanding genetic architectures of traits under
+stabilizing selection in single-population settings
 \citep[e.g.,][]{keightley1988quantitative, simons2018population,
@@
         \aprcomment{Add the SHC model, and the ``vary SD'' fig as a panel here}
+        \aprcomment{First figure should just be two left panels. Combine two
+        right panels to figure 2.}
     }
-    \label{fig:one-pop}
+    \label{fig:stab-sel}
 \end{figure}
 
-In diverged populations, the genomic architecture of a trait under stabilizing
-selection has a higher rate of turnover compared to neutrally evolving loci
-\citep{yair2022population}. We might therefore expect an accumulation of fixed
-differences in the \emph{H. sapiens} and Neanderthal lineages at loci
-contributing to a trait, even when the mean phenotype in each population is
-close to the same trait optimum. When a derived allele at high frequency in one
-population is introduced to another population in which is was previously
-absent, it will be at low frequency and subsequently selected against.
-Likewise, if the ancestral allele is reintroduced to a population fixed for a
-derived allele, the ancestral allele will be at low frequency and will be
-selected against. In either case we should expect selection against the
-introgressed allele, whether ancestral or derived and regardless the historical
-relative sizes of the populations involved. This prediction contrasts with the
-population-genetics ``load'' model, in which haplotypes with fewer deleterious
-variants (such as those from the population with larger historical size) are
-favored after introgression in either direction.
+In diverged populations, genetic variation contributing to a trait under
+stabilizing selection has a higher rate of turnover compared to neutrally
+evolving loci \citep{yair2022population}. We might therefore expect an
+accumulation of fixed differences in the \emph{H. sapiens} and Neanderthal
+lineages at loci contributing to such a trait, even when the mean phenotype in
+each population remains close to the same trait optimum. When a derived allele
+at high frequency in one population is introduced to another population in
+which is was previously absent, it will be at low frequency and subsequently
+selected against. Likewise, if the ancestral allele is reintroduced to a
+population fixed for a derived allele, the ancestral allele will be at low
+frequency and will be selected against. In either case we should expect
+selection against the introgressed allele, whether ancestral or derived and
+regardless the historical relative sizes of the populations involved. This
+prediction contrasts with the population-genetics ``load'' model, in which
+haplotypes with fewer deleterious variants (such as those from the population
+with larger historical size) are favored after introgression in either
+direction.
 
@@
 
-We consider a polygenic trait such that an individual's (additive) genetic
+We consider a polygenic trait for which an individual's (additive) genetic
 value is the sum over all effects of alleles in their genome, so that for
@@
 genotype at that locus (i.e., the number of alleles they carry at that locus).
-For unlinked loci, the expected additive genetic variance is \(V_A = \sum_l
-2p_l(1-p_l)a_l^2\), where $p_l$ is the allele frequency at locus $l$. We ignore
-dominance and epistatic effects, so that \(V_G = V_A\), assuming no linkage.
-We further ingore environmental effects \citep{simons2018population}, so that
-the phenotypic variance \(V_P=V_A\).
+With no linkage between trait-affecting loci, the expected additive genetic
+variance is \(V_A = \sum_l 2p_l(1-p_l)a_l^2\), where $p_l$ is the allele
+frequency at locus $l$. We ignore dominance and epistatic effects, so that
+\(V_G = V_A\).  We further ingore environmental effects
+\citep{simons2018population}, so that the phenotypic variance \(V_P=V_A\).
 
 Stabilizing selection acts to reduce phenotypic variation around the optimum
-value $O$ (typically set to zero), and we assume a Gaussian fitness function so
-that relative fitness is given by \(f(G_i | O, V_S) = \exp{(-(G_i - O)^2 / 2
-V_S)}\). $V_S$ can be interpreted as the strength of selection on the trait,
-where larger $V_S$ implies weaker selection. For a population with mean
-phenotype at (or very close to) the optimum, the mean fitness of the population
-(assuming a roughly normal distribution of phenotypic values in the population)
-is \[\bar{w} \approx \int_{-\infty}^\infty f(G | 0, V_S) \mathcal{N}(0, V_G) dG
-= \left(\frac{V_S}{V_S+V_G}\right)^{1/2}.\] Thus, as the genetic variance
-increases, mean fitness among individuals in the population decreases.
+value $O$ (typally set to zero), and we assume a Gaussian fitness function
+(Figure~\ref{fig:stab-sel}A) so that relative fitness is given by \(f(G_i | O,
+V_S) = \exp{(-(G_i - O)^2 / 2 V_S)}\). $V_S$ can be interpreted as the strength
+of selection on the trait, where larger $V_S$ implies weaker selection. For a
+population with mean phenotype at (or very close to) the optimum, the mean
+fitness of the population (assuming a roughly normal distribution of phenotypic
+values in the population) is \[\bar{w} \approx \int_{-\infty}^\infty f(G | 0,
+V_S) \mathcal{N}(0, V_G) dG = \left(\frac{V_S}{V_S+V_G}\right)^{1/2}.\] Thus,
+as the genetic variance increases, mean fitness among individuals in the
+population decreases.
 
-\subsubsection*{Mutation rates and effect sizes}
+\subsubsection*{Mutation rates, effect sizes and genetic variance}
 
@@
 where $\mu$ is the per-haploid, per-generation rate of mutation. In the
-large-population-size limit, this gives the well-known result
-\[\tilde{V_G} \approx 4\mu V_S,\] provided \(V_G \ll V_S\).
-Mutations are not expected to each have the same effect size $a$, but
-rather will be drawn from some distribution. Throughout, we assume mutation
-effect sizes are drawn from a normal distribution with mean 0 and given
-variance $V_M$.
+large-population-size limit, this gives the well-known result for
+steady-state additive genetic variance,
+\[V_G \approx 4\mu V_S,\] provided \(V_G \ll V_S\).
 
-\aprcomment{where does the $4\mu Ne V_M$ (Lande) result come from? point to fig 1}
+Mutations are not expected to each have the same effect size $a$, but rather
+follow some distribution. Throughout, we assume effect sizes are drawn from a
+normal distribution with mean 0 and given variance $V_M$. When population sizes
+or effect sizes are small, drift dominates the dynamics of $V_G$, and at steady
+state \citet{lande1976natural} found \[V_G \approx 4 N_e \mu V_M.\]
+Interpolating between the selection- and drift-dominated regimes,
+\begin{align}\label{eq:SHC}
+    V_G & \approx \frac{4 \mu V_S}{1 + \frac{VS}{N_e V_M}}.
+\end{align}
+This, the ``stochastic
+house-of-Cards'' (SHC) approximation (Figure~\ref{fig:stab-sel}B), was given by
+\citet{burger1989much} and is discussed in detail in \citet[][Ch.
+28]{walsh2018evolution}.
 
@@
 
-Figure~\ref{fig:one-pop}C shows that if $V_G$ is non-negligible compared to
-$V_S$, ignoring $V_G$ and using $s=a^2/2V_S$ leads to lower estimates of
-additive genetic variance. When mutation rates are large so that $V_G$ is not
-small compared to $V_S$, using $s=a^2/2(V_S+V_G)$ provides estimates of $V_G$
-that closely match simulations assuming free recombination between loci
-(Figure~\ref{fig:toy-admixture}C,D). Because $V_G$ can change over time, this
+If $V_G$ is non-negligible compared to $V_S$, ignoring $V_G$ and using
+$s=a^2/2V_S$ leads to lower estimates of additive genetic variance
+(Figure~\ref{fig:toy-admixture}B). When mutation rates are large so that $V_G$
+is not small compared to $V_S$, using $s=a^2/2(V_S+V_G)$ provides estimates of
+$V_G$ that closely match simulations assuming free recombination between loci
+(Figure~\ref{fig:toy-admixture}X,Y). Because $V_G$ can change over time, this
 means that $s$ is no longer constant, but can change due to factors such as
@@
     \caption{
-        \textbf{}
+        \textbf{Predicting the genetic variance of a trait under stabilizing
+        seletion and non-steady state demography.}
+        \aprcomment{Make it in generations?}
+        \aprcomment{Right two panels should go to supplement, move single-pop
+        with size change from figure 1 to here.}
     }
@@
 contributing $5\%$ ancestry to the recipient population
-(Figure~\ref{fig:toy-admixture}A,B).
+(Figure~\ref{fig:toy-admixture}X,Y).
 
@@
 after admixture to understand allele frequency and haplotype dynamics. The
-trait optimum is kept at $0$ in all branches, and the strength of selection
-$V_S=1$ remains constant.
+trait optimum is kept at $0$ in all branches, so that no optimum shift occurs,
+and the strength of selection $V_S=1$ remains constant.
+
+\begin{figure}[t!]
+    \centering
+    \includegraphics{../figures/neanderthal_admixture.pdf}
+    \caption{
+        \textbf{}
+        \aprcomment{Show total/introgressed/non-introgressed in both directions
+        with both $\sigma_M$.}
+    }
+    \label{fig:neand-to-human}
+\end{figure}
 
@@
 population (labeled 1), \(p_l = fp_{l,0} + (1-f)p_{l,1}\). Plugging into the
-expression for $V_G$, and after some simple algebra (Appendix~AX), we can
-express the expected genetic variance directly after admixture as \[V_G =
-fV_{G,0} + (1-f) V_{G,1} + 2f(1-f)\sum_l F_{2,l} a_l^2,\] where $F_2 = (p_0 -
-p_1)^2$ is the squared difference in allele frequencies at a locus
-\citep{peter2016admixture}. This result is known
+expression for $V_G$ and after some simple algebra (Appendix,
+section~\ref{sec:VG-admixture}), we can express the expected genetic variance
+directly after admixture as \[V_G = fV_{G,0} + (1-f) V_{G,1} + 2f(1-f)\sum_l
+F_{2,l} a_l^2,\] where $F_2 = (p_0 - p_1)^2$ is the squared difference in
+allele frequencies at a locus \citep{peter2016admixture}. This result is known
 \citep[e.g.,][]{tufto2000quantitative}, showing that additive variance is equal
@@
 
-In general, $F_2$ at a given locus will depend on the demographic history
+$F_2$ at a given locus depends on the demographic history
 relating the two populations and the effect size at the locus due to selection
@@
 vanishingly small effect, dynamics at a given locus will be approximately
-neutral, so that $F_2$ depends only on the demography. In this case
-\begin{align*}
+neutral, so that $F_2$ depends only on the demography. In this case,
+\begin{align}
     V_G & \approx f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] \sum_l a_l^2 \\
+    \nonumber
     & = f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] L V_M,
-\end{align*}
-where $V_M$ is the variance of effect sizes of new mutations (assuming the mean
-is zero), and $L$ is the number of trait affecting loci.
+\end{align}
+where $L$ is the number of trait affecting loci.
+
+\subsubsection*{Predicted $V_G$ from the SFS}
 
@@
 approximation for the joint SFS with underdominance. Comparing to simulations
-with free recombination between loci (Appendix~AY), we find that this provides
-an excellent fit to average observed $V_G$ over time
-(Figure~\ref{fig:toy-admixture}C,D). In general, admixture causes a sudden
-increase in $V_G$ followed by a fairly rapid return to pre-admixture levels.
+with free recombination between loci (Appendix,
+section~\ref{sec:VG-admixture}), we find that this provides an excellent fit to
+average observed $V_G$ over time (Figure~\ref{fig:toy-admixture}C,D). In
+general, admixture causes a sudden increase in $V_G$ followed by a fairly rapid
+return to pre-admixture levels.
 
@@
 contributions change over time. Of interest is the contribution to $V_G$ from
-mutations that were already segregating in the recipient population, those that
+alleles that were already segregating in the recipient population, those that
 were introduced through introgression, and new mutations since the time of
-admixture.
+admixture (Figure~\ref{fig:toy-admixture}). In many scenarios, the initial
+divergence of the two populations will have occurred long enough before the
+admixture event, so that the mutations underlying $V_G$ are largely unique in
+each population. After mixing, previously segregating and introgressed alleles
+go to fixation or loss, and the variance of the trait is increasingly due to
+new mutations.
 
-Paragraph on the following (point to appendix and supplement):
-\begin{itemize}
-    \item Benefit of SFS -- partition by frequency classes (prev. seg vs introgressed)
-    \item VG from all both previously segregating and introgressed alleles decays,
-        replaced by VG due to new mutations
-    \item Dividing VG from introgressed alleles into introduced derived vs reintroduced
-        ancestral alleles shows the effect of differences in population sizes
-\end{itemize}
+Introgressed variation can initially make up a considerable portion of $V_G$,
+with those alleles being either newly introduced derived alleles or
+reintroduced ancestral alleles. The relative sizes of the two populations
+impacts the number of each, as derived alleles will accumulate more readily in
+a population with smaller effective size. Nonetheless, the increase in $V_G$
+can be fairly similar in either direction of introgrssion, as the term
+\(2f(1-f)\sum_l F_{2,l} a_l^2\) contributes in either case and can be much
+larger than $f V_G$ from the source population.
 
-\subsection*{Recurrent introgression and non-constant population size}
+\subsubsection*{Complex demography and partitioning heritability by origin of alleles}
 
-\begin{figure}[t!]
-    \centering
-    \includegraphics{../figures/neanderthal_admixture.pdf}
-    \caption{
-        \textbf{}
-    }
-    \label{fig:neand-to-human}
-\end{figure}
+We used a historical model more closely resembling inferred human-Neanderthal
+history (Figure~\ref{fig:neand-to-human}A) to explore the effects of population
+size changes and reciprocal admixture on the additive genetic architecture of
+traits under stabilizing selection. As expected
+(Figure~\ref{fig:toy-admixture}), population contractions decrease $V_G$ as the
+increased rate of drift reduces allelic diversity at trait-affecting loci, and
+introgression increases $V_G$ in both directions of gene flow.
 
-\begin{itemize}
-    \item In the demographic model more closely resembling inferred history
-        connecting humans and Neanderthals (Figure~\ref{fig:neand-to-human}A),
-        we see that demographic history and the distribution of effect sizes
-        of new mutations impact $V_G$
-    \item Additive genetic variance increases after introgression in both directions,
-        which is expected to result in a period of decreased mean fitness
-    \item Population contractions decrease $V_G$, as the increased rate of drift
-        reduces diversity at loci underlying the trait
-    \item TODO: compare to linked simulations
-\end{itemize}
+Because the genetic architectures considered here are purely additive, we can
+track mutations in an admixed populations by whether they were previously
+segregating, fixed or lost in either parental populations or if they arose as
+new mutations since the time of admixture (Methods Section X). This allows us
+to partition $V_G$ by contributions from these different sets of mutations. At
+low introgression proportions, as modeled here, the majority of $V_G$ is still
+contributed by previously segregating, non-introgressed mutations (Figure XAB).
+$V_G$ due to existing mutations decays monotonically over time and is fairly
+rapidly replace by new mutations (Figure XCD).
+
+The average contribution of introgressed vs. non-introgressed SNPs to $V_G$
+(i.e., $h^2$ per SNP) can similarly by tracked over time. For the
+human-Neanderthal-like demographic model and genetic architectures considered
+here, the contribution per SNP of introgressed variants is initally lower than
+that of non-introgressed variants. These contributions change over time,
+depending on mutational variance (Figure YA vs B), as well as demography
+(comparing the two directions of introgression, Figure YC D). When weighting
+$h^2$ per SNP of introgressed SNPs by matching to allele frequencies of
+non-introgressed variants (Methods), relative contributions depend sensitively
+on evolutionary parameters and the time since admixture.
+
+\aprcomment{In discussion, point out that this makes interpreting results, such
+as \cite{wei2023lingering}, challenging.}
 
@@
         \textbf{}
+        \aprcomment{Put $h^2$ per SNP into its own figure.}
     }
@@
 
-\begin{itemize}
-    \item Because we know if a mutation was previously segregating in the population
-        receiving migrants, we can partition variation by contributions from
-        segregating vs introgressed variants (whether introduced derived mutations
-        or reintroduced ancestral mutations)
-    \item At low introgression proportions, as here, the majority of $V_G$ is still
-        contributed by previously segregating mutations, which is then expected to
-        be steadily replaced by new mutations (Figure 2). $V_G$ due to introgressed
-        variation decays monotonically over time (Figure~\ref{fig:neand-to-human}C,D)
-    \item While the total $V_G$ due to introgressed variants decays, we can also
-        consider the average contribution of introgressed vs non-introgressed SNPs
-        to $V_G$ ($h^2$ per SNP) -- at least for this demographic model and
-        DES, the contribution per SNP of introgressed variants is initially lower
-        than those of previously segregating variants. These contributions change
-        over time, in a way that depends on both the genomic architecture underlying
-        the trait and demography
-    \item This dependence on both architecture and demography is apparent when
-        comparing to $V_G$ contributed by introgressed and non-introgressed variants
-        in the early human-to-Neanderthal admixture (Figure~\ref{fig:human-to-neand})
-        -- here, again the contribution
-        from introgressed variants to $V_G$ decays monotonically, but depending on
-        the distribution of effect sizes, relative per-SNP $h^2$ changes over time
-        (though in these cases introgressed $h^2$ per SNP remains lower than
-        non-introgressed, when considering mutations at matched allele frequencies)
-\end{itemize}
+\subsection*{Expected genetic variance with linkage}
 
-\subsection*{The effects of linkage}
+In the preceding sections, we find that approximating the dynamics of
+trait-affecting alleles using an underdominance model
+\citep{robertson1956effect} provides an excellent approximation of $V_G$ in
+complex demographic scenarios. However, this relies on an absence of linkage
+between trait-affecting alleles. The inclusion of linkage can lead to
+noticeable effects on expected $V_G$, so that Equation~\ref{eq:SHC} slightly
+overestimates $V_G$ at steady state
+\citep{burger1989much,burger1994distribution,walsh2018evolution}.
 
-\begin{itemize}
-    \item So far, we have considered model and simulation without linkage -- in this
-        case, the underdominance model provides an excellent approximation
-    \item Including linkage between sites dramatically complicates things, making
-        theorical or analytical progress difficult
-    \item To explore the effects of linkage, we used individual-based simulations
-        with a single large chromosome of size 1 Morgan, and varried the total
-        mutation rate per haploid and the variance of effect sizes of new mutations
-    \item We'll highlight two effects of linkage
-        \begin{itemize}
-            \item Deviations from expected $V_G$, as well as more complicated
-                short-term dynamics after admixture
-            \item The reduction in introgressed ancestry in the regions surrounding
-                trait-affecting alleles
-        \end{itemize}
-\end{itemize}
+To investigate the effects of linkage, we used chromosome-scale
+individual-based simulations. By varying the mutation rate and the variance of
+effect sizes of new mutations, we include scenarios ranging from low to high
+polygenicity and from weak to strong selection on individual alleles. In this
+and the following sections, we will highlight two effects of linkage. First,
+we observe deviations of $V_G$ from expectations without linkage, which can be
+large for highly polygenic traits. Second, selection on introgressed
+trait-affecting alleles results in a reduction of introgressed ancestry in
+surrounding regions.
 
-\subsubsection*{Effects of linkage on expected genetic variance}
+With linkage between two or more alleles affecting a trait under stabilizing
+selection, linkage disequilibrium (LD) can develop between alleles. Classicaly,
+stabilizing selection will cause mutations of opposite-signed effects to be in
+coupling LD and those of same-signed effects in repulsion LD
+\citep{bulmer1971effect}. This has the effect of reducing $V_G$. For lower
+mutational variances ($2N_e V_M\approx1$, Figure~\ref{fig:low_VM}), we observe
+this phenomenon. With low mutational input, and thus low polygenicity, $V_G$ in
+individual simulations closely matches expectations without linkage. As we
+increase the mutation rate, $V_G$ is reduced relative to those expectations.
+However, when the mutational variance is much larger ($2Ne V_m \approx 50$), we
+see the reverse trend (Figure~\ref{fig:high_VM}). At low mutation rate, there
+is a close match between observed $V_G$ and unlinked expectations, although
+simulated values are slightly higher. As the mutation rate increases, $V_G$
+increases to be relatively much larger that expectations, rather than smaller.
 
-\begin{figure}[t!]
-    \centering
-    \includegraphics{../figures/model_comparison.SD_0.01.mu_0.001_0.025}
-    \caption{}
-    \label{fig:linkage-deviations}
-\end{figure}
+The strength of the deviation of $V_G$ between models with and without linkage
+depends on a number of factors. The total mutation rate affects not only the
+polygenicity, with higher mutation rates causing higher number of segregating
+alleles, but also the mean recombination rate between those alleles, as they
+will be more or less densely distributed in the genome. The distribution of
+effect sizes plays an important role, as seen by the opposite trends in $V_G$
+relative to unlinked predictions, which are most apparent with high mutation
+rates. There appear to be complex dynamics involving the Bulmer Effect
+\citep{bulmer1971effect} and interference between selected alleles
+\citep{hill1966effect}, which warrants careful investigation.
 
-\begin{itemize}
-    \item Stabilizing selection reduces $V_G$ by selecting against minor alleles
-        (underdominance), but with linkage it can also reduce $V_G$ by selecting
-        for haplotypes that carry opposite effect alleles that cancel out -- this
-        creates LD between positive and negative effect alleles at linked loci,
-        known as the Bulmer effect \citep{}
-    \item Thus, $V_G$ in simulations with linkage deviate from predictions
-        assuming underdominance alone
-    \item The severity of this deviation depends on a number of factors: mean
-        recombination rate between trait-affecting loci, the total mutation rate,
-        the effect sizes (and distribution of effect sizes of new mutations)
-    \item In general, lower mutation rates imply lower total $V_G$, lower polygenicity,
-        and therefore higher recombination rates between trait-affecting alleles --
-        we therefore expect this deviation to be exacerbated with higher mutation
-        rates (all else being equal)
-\end{itemize}
-
-\subsubsection*{Reduction of introgressed ancestry surrounding an introgressed trait-affecting allele}
+\subsubsection*{Introgressed ancestry is reduced around introduced trait-affecting alleles}
 
@@
         introgressed ancestry frequency}
+        \aprcomment{Combine with next figure}
     }
@@
         \textbf{}
+        \aprcomment{Combine with previous figure}
     }
@@
 \end{figure}
+
+\subsection*{Introgressed ancestry deserts are shared under reciprocal introgression}
 
@@
 
-\section{Appendix}
+\appendix
 
-\subsection{Additive genetic variance after admixture}
+\section{Additive genetic variance after admixture}\label{sec:VG-admixture}
 
@@
 
-\subsection{Dynamics of introgressed alleles}
+\section{Dynamics of introgressed alleles}
 
-\subsection{Reciprocal depletion of functional variation}
+\section{Reciprocal depletion of functional variation}
 

### 2023-11-21 doc/manuscript.tex
finished drafting results

@@
 
-\subsection*{Expected genetic variance with linkage}
+\subsection*{The effects of linkage}
 
@@
 effect sizes plays an important role, as seen by the opposite trends in $V_G$
-relative to unlinked predictions, which are most apparent with high mutation
-rates. There appear to be complex dynamics involving the Bulmer Effect
+relative to unlinked predictions (Figure~X vs. Figure~Y), which are most
+apparent with high mutation rates, and thus high polygenicity. In such cases,
+there appear to be complex dynamics involving the Bulmer Effect
 \citep{bulmer1971effect} and interference between selected alleles
-\citep{hill1966effect}, which warrants careful investigation.
+\citep{hill1966effect}.
 
-\subsubsection*{Introgressed ancestry is reduced around introduced trait-affecting alleles}
+\subsubsection*{Introgressed ancestry is reduced around introduced
+trait-affecting alleles}
 
-\begin{itemize}
-    \item Because introgressed trait-affecting alleles are selected against as the
-        minor allele, linkage will also remove introgressed ancestry segments in the
-        regions surrounding the selected alleles
-    \item The expected reduction in introgressed ancestry proportions surrounding
-        functional alleles will depend on the effect size of the trait-affecting allele
-        and the probability of recombination surrounding that site
-    \item We can consider first a deterministic model of the dynamics of introgressed
-        allele frequencies at and around an introgressed functional allele -- this
-        model considers a single trait-affecting locus, ignoring any effects of
-        interference between selected alleles
-        \begin{itemize}
-            \item Consider a site with fixed differences between the two populations
-                (the derived allele is fixed in one, lost in the other)
-            \item At the functional site itself, with admixture proportion $f$ from the
-                minor parental ancestry, the initial frequency of the derived allele
-                is $p_0=f$ or $p_0 = 1-f$, if the derived allele was fixed in the 
-                major parental ancestry population
-            \item In one generation, the expected allele frequency at the selected locus
-                is \[p_{t+1} \approx p_t - s p_t(1-p_t)(1-2p_t),\] where $s=a^2/2V_S$
-            \item As discussed earlier, if $p_0<1/2$, $p_t\rightarrow0$, and if $p_0>1/2$,
-                $p_t\rightarrow1$ as $t\rightarrow\infty$
-            \item Consider a neutral locus separated from the selected locus by
-                recombination probability $r$. Initially, the expected frequency
-                of introgressed ancestry at that locus is $q_0=f$
-            \item The linked ancestry frequency will change due to selection acting on
-                the trait-affecting locus, depending on the strength of selection
-                at that locus and the level of linkage disequilibrium between
-                the two loci
-            \item $D=Cov(p,q)$ is the standard covariance measure of LD between loci,
-                and $q$ is expected to change as \[q_{t+1} \approx q_t - s D_t(1-2p_t)\]
-            \item $D$ also changes over time, due to both selection and recombination,
-                so that \[D_{t+1} \approx D_t - r D_t - s D_t (1-2p_t)^2\]
-            \item Initially, $D_0 = \pm f(1-f)$, with $D$ being positive if $p_0=f$
-                and negative if $p_0=1-f$
-            \item Together, this forms a nonlinear system of equations for the
-                deterministic change in allele frequencies at the two loci (one
-                selected, one neutral) and LD between them -- note that this is
-                deterministic in the infinite population size limit
-        \end{itemize}
-    \item Using this model, we can predict how allele frequencies and LD are expected
-        to change over time after admixture, for a given effect size $a$ and 
-        recombination rate $r$ (Figure~\ref{fig:linkage-pred}A,B)
-    \item As expected, lower effect sizes result in a slower reduction in introgressed
-        ancestry frequency, while higher recombination rates more quickly decouple
-        the linked ancestry from the selected allele
-    \item Thus, the expected reduction in introgressed ancestry is largest for
-        larger effect sizes (Figure~\ref{fig:linkage-pred}C), and LD between the
-        selected and linked neutral alleles is largest for neutral sites closest
-        to the selected allele (Figure~\ref{fig:linkage-pred}D)
-\end{itemize}
+Because introgressed trait-affecting alleles are selection against as the minor
+allele, introgressed ancestry segments in the regions surrounding the selected
+alleles will also be removed due to linkage. The expected reduction in such
+introgressed ancestry will depend on the effect size of the trait-affecting
+allele and the probability of recombination surrounding that locus. We first
+consider a deterministic model of the dynamics of introgressed allele and
+linked ancestry frequencies with variable recombination between them. This
+simple model ignores the effects of drift and of interference between multiple
+selected alleles.
+
+We model a trait-affecting site with a fixed difference between the two
+parental populations, in which the derived allele may be fixed in either
+population. At the functional site, with admixture proportion $f$ from the
+minor parental source, the initial frequency of the derived allele is either
+$p_0=f$ or $p_0=1-f$. Over one generation, the expected allele frequency at the
+selected locus is, to leading order in $s$,
+\begin{align}\label{eq:system-p}
+    p_{t+1} & = p_t - s p_t(1-p_t)(1-2p_t),
+\end{align}
+with
+\(s=a^2/2V_S\). As discussed above, if \(p_0<1/2\), \(p_t\rightarrow0\), and if
+\(p_0>1/2\), \(p_t\rightarrow1\) as \(t\rightarrow\infty\).
+
+We consider a neutral locus separated from the selected locus by recombination
+probability $r$. Initially, the expected frequency of linked introgressed
+ancestry is \(q_0=f\), which changes over time due to linked selection on the
+trait-affecting allele. Letting \(D=Cov(p,q)\) be the standard covariance
+measure of LD between the two loci, $q$ is expected to change as
+\begin{align}\label{eq:system-q}
+    q_{t+1} & = q_t - s D_t(1-2p_t).
+\end{align}
+Initially, \(D_0=\pm f(1-f)\), with \(D\) being
+positive if \(p_0=f\) and negative if \(p_0=1-f\). LD between the loci changes
+deterministically over time, due to both selection and recombination, so that
+\begin{align}\label{eq:system-D}
+    D_{t+1} = D_t - r D_t - s D_t (1-2p_t)^2.
+\end{align}
+Together, this forms a
+simple nonlinear system of equations for the deterministic change in allele
+frequencies at the two loci and LD between them.
+
+Using this model, we predict the changes in introgressed allele frequences and
+LD after admixture for given effect size $a$ and recombination rate $r$
+(Figure~\ref{fig:linkage-pred}). As expected, smaller effect sizes result in a
+slower decay in introgressed ancestry frequency at both selected and linked
+loci, and larger recombination rates more quickly decouple the linked ancestry
+from the selected allele dynamics. Thus, the expected reduction in introgressed
+ancestry is largest for larger effect sizes (Figure~\ref{fig:linkage-pred}X),
+and LD between the selected and linked neutral alleles is largest for neutral
+sites closest to the selected allele (Figure~\ref{fig:linkage-pred}Y).
+
+We assessed the accuracy of the deterministic two-locus model using
+individual-based forward-in-time simulations \citep{thornton2019polygenic}
+under a simple demographic model (Figure~\ref{fig:toy-admixture}A), with
+introgression fraction \(f=0.05\). We simulated a single chromosome of length 1
+Morgan, with all mutations having effect sizes \(\pm a\) (\(a=0.02\) or
+\(0.05\)), and we varied the total per-chromosome mutation rate (\(\mu=0.001\),
+\(0.0025\) and \(0.01\)). For each fixed-difference mutation in the parental
+populations, we determined the average introgressed ancestry surrounding such
+loci (Methods).
+
+The deterministic model (Equations~\ref{eq:system-p}--\ref{eq:system-D})
+provides a very good approximation when mutation rates are small. As the
+mutation rate increases in these simulations, polygenicity also increases so
+that trait-affecting alleles are more densely distributed along the chromosome.
+Deviations from the deterministic model are due to multiple selected alleles
+affecting local introgressed ancestry. For the highest mutation rate shown
+here, there are expected to be multiple other trait-affecting alleles within a
+1 cM window around any focal SNP, which interact with each other and also
+affect local ancestry proportions.
 
@@
 
-\begin{itemize}
-    \item To explore the accuracy of this two-locus model, we compared to
-        simulations using the demographic model shown in
-        Figure~\ref{fig:toy-admixture}A with introgression fraction $f=0.05$.
-        We simulated a single 1 Morgan
-        chromosome, with all mutations having effect sizes $\pm a$ ($a=0.02$ or
-        $0.05$, with equal probability of being trait-increasing or decreasing)
-        and varying the total mutation rate ($\mu=0.001$, $0.0025$, and $0.01$)
-    \item For each mutation that was fixed in one population and absent in the
-        other (either fixed in the minor or major ancestry population), we
-        determined the average introgressed ancestry surrounding those introgressed
-        alleles (Figure~\ref{fig:linkage-sim})
-    \item The deterministic model provides a very good approximation for linked
-        introgressed ancestry when the mutation rate is small
-    \item As the mutation rate increases, polygenicity increases so that
-        trait-affecting alleles are more densely distributed along the chromosome --
-        deviations from the deterministic model are due to interference between
-        tightly linked trait-affecting alleles. For the highest mutation rate shown
-        here, there will be many trait-affecting alleles within the 1 cM window
-        around any focal SNP
-\end{itemize}
-
 \begin{figure}[t!]
@@
 
-\subsection*{Introgressed ancestry deserts are shared under reciprocal introgression}
+\subsection*{Introgressed ancestry deserts are shared under stabilizing
+selection and reciprocal introgression}
+
+As shown in the previous section, introgressed ancestry is reduced around loci
+with trait-affecting alleles regardless of the parental population the derived
+allele is present in. We should therefore expect to observe reductions in
+introgressed ancestry at the same trait-affecting loci after gene flow in
+either direction. If introgression occurs in both directions, regions of
+reduced introgressed ancestry will coincide around such loci, and introgression
+``deserts'' that appear due to selection against trait-affecting alleles will
+tend to be shared.
+
+To demonstrate this effect, we performed chromosome-scale simulations under a
+model of human-Neanderthal reciprocal introgression
+\citep[Figure~\ref{fig:neand-to-human}A,][]{harris2023diverse}. Mutations
+affecting a trait under stabilizing selection in each population could arise in
+100 kb ``functional'' regions, spaced 1 Mb apart (Methods). Sampling
+individuals from both the human and Neanderthal lineages, we observed the
+average introgressed ancestry proportions were lowest within functional regions
+and dipped in the surrounding regions (Figure~\ref{fig:deserts}A). This
+corresponded to an increased proportion of ancestry deserts (defined as
+    contiguous 50kb regions with no observed introgressed ancestry in the
+sample) within and immediately surrounding the functional regions. Functional
+regions also displayed an enrichment of \emph{shared} ancestry deserts, with
+the probability of observing ancestry deserts (either within a population or
+shared) decaying to background levels as the distance from the functional
+region increases.
 

### 2023-11-21 doc/manuscript.tex
start fleshig out discussion

@@
 neutral, so that $F_2$ depends only on the demography. In this case,
-\begin{align}
+\begin{align} \label{eq:VG-admixture}
     V_G & \approx f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] \sum_l a_l^2 \\
@@
 
+Many phenotypic traits are under stabilizing selection
+\citep{hodgins2015gene,sanjak2018evidence,sella2019thinking}. This motivates
+stabilizing selection around a shared optimum as an appropriate null models for
+the dynamics of allele frequencies affecting polygenic traits
+\citep{yair2022population}. Stabilizing selection results in distinct dynamics
+for trait-affecting alleles, characterized by symmetric underdominance
+\citep{robertson1956effect,keightley1988quantitative}, that is, selection
+against the minor allele. Because admixture and introgression events are common
+in natural populations, it will be important to understand the effects of gene
+flow on the architecture of complex traits.  \aprcomment{or\dots``studies of
+complex traits in admixed populations rely on this understanding, or
+something''}
+
+After admixture between diverged populations, the additive genetic variance of
+a trait rapidly increases. The genetic variance after admixture depends on both
+the existing genetic variances within the parental population and their
+divergence at trait-affecting loci, measured by $F_2$
+(Equation~\ref{eq:VG-admixture}). $V_G$ is then expected to fairly quickly
+decay to background levels. During this process, introgressed and pre-existing
+trait-affecting alleles are replaced by new mutations, turning over the genetic
+architecture of the trait (Figure\ref{fig:toy-admixture}). Selection occurs
+against the minor alleles at trait-affecting loci, so that introgressed alleles
+from the minor parental ancestry, whether derived or reintroduced ancestral
+alleles, are selected against (Figure~\ref{fig:linkage-pred}).
+
+Selection against introgressed alleles will also remove linked introgressed
+ancestry in the surrounding regions. Thus, introgressed ancestry deserts are
+more likely to form at and around loci contributing to selected traits
+(Figure~\ref{fig:deserts}). This process is symmetric, so that deserts tend to
+form in the same regions under reciprocal introgression. When comparing
+distributions of introgressed haplotypes in two diverged populations with more
+recent gene flow, we should expect to see an overlap in deserts of introgressed
+ancestry at loci affecting traits under stabilizing selection.
+
+The expected pattern of shared ancestry deserts under stabilizing selection
+differs from models of both deleterious load and incompatibilities. Load-based
+models predict that haplotypes that have accumulated more deterious mutations,
+e.g., from a population with small long-term effective population size, will be
+selected against under either direction of gene flow. Introgressed ancestry at
+a given selected locus will decrease in frequency in one introgression scenario
+and increase the other. This may explain the replacement of MT and Y chromosome
+DNA in Neanderthals by human haplotypes after early human-to-Neanderthal
+introgression \citep{posth2017deeply,petr2020evolutionary}.
+
+\citet{muller1942isolating} originally hypothesized that hybrid
+incompatibilities should not often form within a single locus, but instead the
+two or more loci involved will most often be found at distant or unlinked loci
+in genome. Theory predicts that hybrid incompatibilities are resolved via
+selection against the minor parental ancestry [cite]. Depending on the
+direction of introgression, different incompatibilities alleles will be
+selected against and ancestry deserts will form around those different loci.
+While such a model predicts selection against introgressed ancestry, similar to
+the stabilizing selection model, an important distinction is that deserts are
+not expected to overlap with bidirectional gene flow.
+
+\aprcomment{any data about distributions of incompatibility loci in the
+genome?}
+
+\aprcomment{Second-to-last paragraph}
+Turning to observations from human and Neanderthal genomes
+\citep{harris2023diverse}
+
+\aprcomment{last paragraph}
 \begin{itemize}
-    \item Stabilizing selection with a constant optimum is an appropriate
-        null model for the dynamics of trait variation and selection on
-        alleles underlying those traits
-    \item Stabilizing selection results in population genetic predictions
-        for patterns of introgressed variation in and around functional
-        regions that differ from both load- and incompatibility-based arguments
-        \begin{itemize}
-            \item Load-based models predict that haplotypes from the less ``loaded''
-                population will be favored under introgression in either direction
-                -- thus, regions of depleted ancestry after introgression in one
-                direction will not be expected to also be introgressed ancestry
-                deserts after admixture in the opposite direction
-            \item \citet{mueller1942} originally pointed out that hybrid
-                incompatibilities should not form at single loci -- classical
-                incompatibilities require two or more loci, which are not
-                expected to be tightly linked to one another. Thus, under a
-                model of hybrid incompatibilities, the minor parent ancestry
-                is again selected against, but those loci will be in different
-                regions. Again, deserts are therefore not expected to overlap
-                if selection is largely due to incompatibilities
-            \item Instead, stabilizing selection causes introgressed DNA to be
-                depleted at the same loci under bi-directional gene flow
-                (could be thought of as single-locus incompatibility effects --
-                but these are not incompatibilities in the classical sense,
-                as the same effects occur even without admixture)
-            \item A prediction we could make is that, if current models are correct
-                that there has been introgression in both directions between
-                \emph{H. sapiens} and Neanderthals, introgressed ancestry deserts
-                should overlap with each other more often than expected by chance
-            \item See Figure~\ref{fig:ancestry-deserts}
-        \end{itemize}
-    \item Shared deserts are most likely to form where moderate- or large-effect
-        alleles have fixed in one population or the other -- this could provide
-        an approach to map\dots?
     \item Selection in structured populations with ongoing migration or
@@
         future
-    \item Some conjectures about hybrid zones
+\end{itemize}
+\aprcomment{end on a comment about the usefulness of the SFS approach and the
+promise for future studies of dynamics of the architecture of complex traits
+with non-trivial demographic scenarios -- possible that this should be earlier,
+and we can include caveats there. Maybe second paragraph of discussion.}
+
+\aprcomment{don't end on this, or might not need to include it -- caveats can
+come earlier?}
+\begin{itemize}
+    \item Some conjectures about hybrid zones - single-locus ``incompatibilities''
     \item Pleiotropy\dots!!!
@@
 \subsection*{Individual-based simulations with linkage}
+
+\section*{Acknowledgements}
+\begin{itemize}
+    \item Lab members
+    \item Kevin Thornton
+    \item Bret Payseur
+\end{itemize}
 

### 2023-11-23 doc/manuscript.tex
discussion drafted

@@
 
-\newcommand{\aprcomment}[1]{{\textcolor{blue}{Comment: #1}}}
+\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
 \newcommand{\E}{\mathbb{E}}
@@
 \begin{abstract}
-    .
+
 \end{abstract}
@@
 Many phenotypic traits are under stabilizing selection
-\citep{hodgins2015gene,sanjak2018evidence,sella2019thinking}. This motivates
-stabilizing selection around a shared optimum as an appropriate null models for
-the dynamics of allele frequencies affecting polygenic traits
-\citep{yair2022population}. Stabilizing selection results in distinct dynamics
-for trait-affecting alleles, characterized by symmetric underdominance
+\citep{hodgins2015gene,sanjak2018evidence,sella2019thinking}. This has
+motivated using stabilizing selection around a shared optimum as a null model
+for the dynamics of alleles affecting polygenic traits
+\citep{yair2022population}. Stabilizing selection has a distinct effect on
+trait-affecting alleles, characterized by symmetric underdominance
 \citep{robertson1956effect,keightley1988quantitative}, that is, selection
-against the minor allele. Because admixture and introgression events are common
-in natural populations, it will be important to understand the effects of gene
-flow on the architecture of complex traits.  \aprcomment{or\dots``studies of
-complex traits in admixed populations rely on this understanding, or
-something''}
+against the minor allele. There has been some previous work on how migration
+shapes the genetic variance of a trait under stabilizing selection
+\aprcomment{[cite]}, but most studies consider single populations, often at
+steady state. Admixture and introgression events are common in natural
+populations, so understanding their effects on the architecture of complex
+traits is needed.
+
+Here, we have shown that multi-population, non-equilibrium approaches for the
+site frequency spectrum with underdominance can be used to model the additive
+genetic variance of a trait under stabilizing selection. While this approach
+ignores biologically relevant aspects such as linkage between sites and
+pleiotropy, it can still provide important insights into the dynamics of trait
+architectures. By modeling allelic contributions to the trait as additive, it
+allows for the decomposition of contributions from alleles of different
+origins, either by source population or mutation time. It may therefore be
+useful for understanding the expected contributions of introgressed variants to
+complex traits, for example from Neanderthal-introgressed mutations in humans
+\citep{wei2023lingering}. In the limited scenarios explored here, we see that
+the expected contributions of introgressed variants to heritability are
+complicated even in the purely additive case, depending on the distribution of
+effect sizes, demographic history, and the time since admixture
+(Figures~\ref{fig:neand-to-human}, \ref{fig:human-to-neand}).
+
+\aprcomment{While we have focused on the relatively simple scenario of two
+diverged populations recently exchanging migrants, SFS methods presented here
+can incorporate continuous migration, more than two populations, or changes in
+the trait optimum. We leave these for future work\dots}
 
@@
 divergence at trait-affecting loci, measured by $F_2$
-(Equation~\ref{eq:VG-admixture}). $V_G$ is then expected to fairly quickly
-decay to background levels. During this process, introgressed and pre-existing
-trait-affecting alleles are replaced by new mutations, turning over the genetic
-architecture of the trait (Figure\ref{fig:toy-admixture}). Selection occurs
-against the minor alleles at trait-affecting loci, so that introgressed alleles
-from the minor parental ancestry, whether derived or reintroduced ancestral
-alleles, are selected against (Figure~\ref{fig:linkage-pred}).
+(Equation~\ref{eq:VG-admixture}). $V_G$ is then expected to decay fairly
+quickly to background levels. During this process, introgressed and
+pre-existing trait-affecting alleles are replaced by new mutations, turning
+over the genetic architecture of the trait (Figure\ref{fig:toy-admixture}).
+Selection occurs against the minor alleles at trait-affecting loci, so that
+introgressed alleles from the minor parental ancestry, whether derived or
+reintroduced ancestral alleles, are selected against
+(Figure~\ref{fig:linkage-pred}).
 
@@
 introgression \citep{posth2017deeply,petr2020evolutionary}.
+%However, this is
+%not the general pattern observed when comparing introgressed ancestry in the
+%autosomal genomes of humans and Neanderthals \citep{harris2023diverse}.
 
-\citet{muller1942isolating} originally hypothesized that hybrid
-incompatibilities should not often form within a single locus, but instead the
-two or more loci involved will most often be found at distant or unlinked loci
-in genome. Theory predicts that hybrid incompatibilities are resolved via
-selection against the minor parental ancestry [cite]. Depending on the
-direction of introgression, different incompatibilities alleles will be
-selected against and ancestry deserts will form around those different loci.
-While such a model predicts selection against introgressed ancestry, similar to
-the stabilizing selection model, an important distinction is that deserts are
-not expected to overlap with bidirectional gene flow.
+The classical model of Bateson-Dobzhansky-Muller incompatibilities (BDMIs)
+\citep{bateson1909heredity,dobzhansky1936studies,muller1942isolating} explains
+the accumulation of reproductive isolation through negative epistatic
+interactions that are exposed in hybrids. \citet{muller1942isolating}
+hypothesized that such BDMIs should most often form between distant or unlinked
+loci, instead of within single or tightly linked loci. Because theory predicts
+that hybrid incompatibilities are resolved via selection against the minor
+parental ancestry \citep{moran2021genomic}, ancestry deserts should form in
+different genomic regions as the different incompatibility alleles are selected
+against, depending on the direction of introgression. While such a model
+predicts selection against introgressed ancestry, similar to that of
+stabilizing selection, an important distinction is that deserts due to
+selection against incompatibility loci are not expected to overlap under
+bidirectional gene flow. While there is little empirical data on the
+distribution of BDMIs, studies point to interacting BDMI alleles being unlinked
+\citep[][e.g.,]{presgraves2003fine} and an asymmetry in the alleles under
+selection in different introgression scenarios
+\citep{maheshwari2011genetics,moran2021genomic}.
 
-\aprcomment{any data about distributions of incompatibility loci in the
-genome?}
-
-\aprcomment{Second-to-last paragraph}
-Turning to observations from human and Neanderthal genomes
-\citep{harris2023diverse}
-
-\aprcomment{last paragraph}
-\begin{itemize}
-    \item Selection in structured populations with ongoing migration or
-        meta-populations -- SFS approach can handle this -- prospects for the
-        future
-\end{itemize}
-\aprcomment{end on a comment about the usefulness of the SFS approach and the
-promise for future studies of dynamics of the architecture of complex traits
-with non-trivial demographic scenarios -- possible that this should be earlier,
-and we can include caveats there. Maybe second paragraph of discussion.}
-
-\aprcomment{don't end on this, or might not need to include it -- caveats can
-come earlier?}
-\begin{itemize}
-    \item Some conjectures about hybrid zones - single-locus ``incompatibilities''
-    \item Pleiotropy\dots!!!
-\end{itemize}
+Finally, turning to the distribution of introgressed ancestry deserts in humans
+and Neanderthals, \citet{harris2023diverse} find that regions of depleted
+Neanderthal ancestry in humans overlap more than would be expected by chance
+with regions lacking human-introgressed alleles in the Altai Neanderthal.
+Human-introgressed ancestry in the Neanderthal genome is also depleted in
+functional regions, as observed in humans \citep{sankararaman2014genomic}.
+\citet{harris2023diverse} propose that epistatic interactions between
+introgressed alleles and the recipient backgrounds could drive these patterns,
+which they interpret as evidence for the inital process of speciation between
+humans and Neanderthals. However, this observation of shared ancestry deserts
+does not match expectations under a classic model of BDMIs, as argued above.
+Instead, at least some of the pattern may be due to stabilizing selection
+acting on complex traits, such as gene regulation.  Importantly, such
+overlapping ancestry deserts are expected even when a trait is under
+stabilizing selection for the same phenotypic optimal value. The selective
+causes of ancestry deserts in both humans and Neanderthals remain unknown.
+Stabilizing selection will prove a useful null model for testing scenarios of
+epistasis, incompatibilities and adaptive introgression.
 

### 2023-12-14 doc/manuscript.tex
update doc, directional deserts

@@
 
-\usepackage{fullpage}
-\usepackage{amsmath}
+\usepackage[letterpaper,margin=1in]{geometry}
+\usepackage{lineno}   %% <- no mathlines option
+\usepackage{amsmath}  %% <- after lineno
+\usepackage{etoolbox} %% <- for \cspreto, \csappto
 \usepackage{amsfonts}
@@
 \usepackage{color}
+\usepackage{url}
 
-\usepackage{setspace}
+\usepackage{setspace, caption}
+\captionsetup{font=onehalfspacing} % Spacing in float captions
 
@@
 \DeclareMathOperator{\Var}{Var}
-
 \newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
 \newcommand{\E}{\mathbb{E}}
+\usepackage{xspace}
+\newcommand{\moments}{\texttt{moments}\xspace}
+\newcommand{\fwdpy}{\texttt{fwdpy11}\xspace}
+\newcommand{\tskit}{\texttt{tskit}\xspace}
+
+%% Patch 'normal' math environments:
+\newcommand*\linenomathpatch[1]{%
+  \cspreto{#1}{\linenomath}%
+  \cspreto{#1*}{\linenomath}%
+  \csappto{end#1}{\endlinenomath}%
+  \csappto{end#1*}{\endlinenomath}%
+}
+
+\linenomathpatch{equation}
+\linenomathpatch{gather}
+\linenomathpatch{multline}
+\linenomathpatch{align}
+\linenomathpatch{alignat}
+\linenomathpatch{flalign}
+
+%\linenumbers
 
@@
 
+    Many phenotypic traits are under stabilizing selection, which maintains
+    individual phenotypic values near some optimum. Classical theory shows that
+    at loci contributing to the genetic variance of a trait under stabilizing
+    selection, the minor allele is selected against via symmetric
+    underdominance. Most studies of the dynamics of the genetic architecture
+    and variance of a trait under stabilizing selection have focused on single
+    populations. However, natural populations are typically structured in some
+    way, and admixture and introgression are common occurances, in particular
+    in human evolution. Because stabilizing selection results in selection
+    against the minor allele at a trait-affecting locus, the alleles from the
+    minor parental ancestry will be selected against after admixture. Here, we
+    show that the site frequency spectrum can be used to model the genetic
+    architecture of such traits, allowing for complex multi-population
+    histories including admixture. We develop a simple model with linkage to a
+    neutral locus to show that introgressed ancestry is expected to be depleted
+    around trait-affecting alleles. Using this model and individual-based
+    simulations, we show that introgressed ancestry deserts tend to form around
+    such loci. When introgression between two diverged populations occurs in
+    both directions, as has been inferred between humans and Neanderthals,
+    introgressed ancestry deserts will often be shared across populations.
+    Stabilizing selection for a shared phenotypic optimum may therefore explain
+    recent observations in which regions of depleted human-introgressed
+    ancestry in the Neanderthal genome overlap with Neanderthal-ancestry
+    deserts in humans.
+
 \end{abstract}
@@
 
-Genomic surveys of natural systems suggest that historical admixture among
+Genomic surveys of natural systems show that historical admixture among
 diverged populations or closely related taxa commonly occurs
-\citep{brandvain2014speciation,skoglund2015ancient,suvorov2022widespread} and
-is potentially widespread in primate
-\citep{tung2017contribution,sorensen2023genome} and hominin
-\citep{wolf2018outstanding,peter2020100} evolution. Admixture may therefore be
-an important driver of phenotypic and molecular variation and shape the genetic
-basis of complex traits. In humans, archaic introgression involving
-Neanderthals and Denisovans has attracted considerable attention, with efforts
-to describe the historical processes leading to observed distributions of
-introgressed DNA in present-day populations
-\citep{prufer2014complete,villanea2019multiple,chen2020identifying} and its
-contribution to quantitative traits
-\citep{sankararaman2016combined,wei2023lingering}.
+\citep{brandvain2014speciation, skoglund2015ancient, suvorov2022widespread} and
+is potentially widespread in primate \citep{tung2017contribution,
+sorensen2023genome} and hominin \citep{wolf2018outstanding, peter2020100}
+evolution. Admixture may therefore be an important driver of phenotypic and
+molecular variation and shapes the genetic basis of complex traits. In humans,
+archaic introgression involving Neanderthals and Denisovans has attracted
+considerable attention, including efforts to describe the historical processes
+leading to observed distributions of introgressed DNA in present-day
+populations \citep{prufer2014complete, villanea2019multiple,
+chen2020identifying} and the contribution of introgressed variation to
+quantitative traits \citep{sankararaman2016combined, wei2023lingering}.
 
 Once introduced through admixture, introgressed alleles may be selected for or
-against. Some introgressed haplotypes were positively selected in modern humans
-\citep{huerta2014altitude,racimo2017signatures,enard2018evidence,gower2021detecting},
-possibly due to locally adaptive variation that provided fitness advantages as
-humans encountered novel environments. Despite some cases of adaptive
-introgression, introduced alleles were more likely to have been selected
-against in modern humans \citep{harris2016genetic,juric2016strength}. Since
+against. Some introgressed haplotypes were likely positively selected in modern
+humans \citep{huerta2014altitude, racimo2017signatures, enard2018evidence,
+gower2021detecting}, possibly due to locally adaptive variation that provided
+fitness advantages as humans encountered novel environments. Despite some cases
+of adaptive introgression, mostly introduced alleles were likely to have been
+selected against in humans \citep{harris2016genetic, juric2016strength}. Since
 Neanderthal and Denisovan population sizes were relatively small for hundreds
-of thousands of years, theory predicts they would have accumulated deleterious
-variants at an increased rate. Introgressed haplotypes loaded with more
-deleterious mutations would then have quickly been selected against after
+of thousands of years, theory predicts they would have accumulated fixed
+deleterious variation at an increased rate. Introgressed haplotypes loaded with
+more deleterious mutations would then have quickly been selected against after
 admixture. Mapping the distribution of Neanderthal-introgressed haplotypes in
-humans shows a reduction of Neanderthal-related ancestry in enhancers and
-regulatory regions
-\citep{petr2019limits,telis2020selection,yermakovich2023long}. These
-``deserts'' of Neanderthal ancestry support the hypothesis that introgressed
-functional alleles were selected against
-\citep{sankararaman2014genomic,sankararaman2016combined}.
+humans shows a reduction of Neanderthal-related ancestry in coding and
+regulatory regions \citep{petr2019limits, telis2020selection,
+yermakovich2023long}. These ``deserts'' of Neanderthal ancestry support the
+hypothesis that introgressed functional alleles were selected against
+\citep{sankararaman2014genomic, sankararaman2016combined}.
 
 There is growing genetic evidence that early \emph{H. sapiens} reciprocally
-contributed to Neanderthal genomes
-\citep{kuhlwilm2016ancient,hubisz2020mapping,harris2023diverse}. This
-human-to-Neanderthal introgression occured tens or hundreds of thousands of
-years prior to Neanderthal introgression in humans during modern human global
-dispersal around 60 ka. This is supported by ``near modern'' \emph{H. sapiens}
-outside of African around 120--100ka or earlier
-\citep{schwarcz1988esr,grun2005u,beyer2021climatic}, potentially overlapping
-with Neanderthals and providing opportunities for early contacts. While
-estimates of the genomic contribution of early \emph{H. sapiens} to
-Neanderthals vary, around $6\%$ of later Neanderthal genomes may have been
-contributed through introgression from humans to Neanderthals
+contributed to Neanderthal genomes \citep{kuhlwilm2016ancient,
+hubisz2020mapping, harris2023diverse}. Such introgression would have occurred
+tens or hundreds of thousands of years prior to Neanderthal introgression in
+humans during the global dispersal of modern humans around 60 ka. This is
+supported by ``near modern'' \emph{H. sapiens} outside of African around
+120--100 ka or earlier \citep{schwarcz1988esr, grun2005u, beyer2021climatic},
+potentially overlapping with Neanderthals and providing opportunities for early
+contacts. While estimates of the genomic contribution of early \emph{H.
+sapiens} to Neanderthals vary, around $6\%$ of later Neanderthal genomes may
+have been contributed through introgression from humans to Neanderthals
 \citep{harris2023diverse}. If \emph{H. sapiens}-related haplotypes carried
@@
 The replacement of Neanderthal mitochondrial and Y chromosomes by early human
-haplotypes supports such a model of post-admixture selection in the Neanderthal
-lineage \citep{posth2017deeply,petr2020evolutionary}.
+haplotypes supports this model of post-admixture selection in the Neanderthal
+lineage \citep{posth2017deeply, petr2020evolutionary}.
 
@@
 unconditionally deleterious variation in populations of different sizes) or
-hybrid incompatibilities \citep{muller1942isolating}. Such arguments, founded
+hybrid incompatibilities \citep{muller1942isolating}. These arguments, founded
 in population genetics theory, rarely take into account selection operating on
@@
 variation. Many phenotypic traits are thought to be under stabilizing selection
-\citep{sanjak2018evidence,sella2019thinking}, including gene regulation
-\citep{gilad2006natural,hodgins2015gene,price2022detecting}. Because some of
+\citep{sanjak2018evidence, sella2019thinking}, including gene regulation
+\citep{gilad2006natural, hodgins2015gene, price2022detecting}. Because some of
 the strongest signals of selection against Neanderthal-introduced alleles are
@@
 near some optimum, which is achieved by reducing phenotypic variation
-(Figure~\ref{fig:stab-sel}A). When the mean phenotype of the population is close
-to the phenotypic optimum, classical models predict that alleles contributing
-to such a trait are subject to underdominant selection, i.e., selection against
-the minor allele at a locus \citep{robertson1956effect}. This has proven to be
-a useful model for understanding genetic architectures of traits under
+(Figure~\ref{fig:stab-sel}A). When the mean phenotype of the population is
+close to the phenotypic optimum, classical models predict that trait-affecting
+alleles are subject to underdominant selection, i.e., selection against the
+minor allele at the locus \citep{robertson1956effect}. This has proven to be a
+useful model for understanding genetic architectures of traits under
 stabilizing selection in single-population settings
@@
     \centering
-    \includegraphics{../figures/one_pop.pdf}
+    \includegraphics{../figures/stab_sel.pdf}
     \caption{
-        \textbf{Stabilizing selection and additive genetic variance in one population.}
-        \aprcomment{See comparisons to simulations assuming free recombination in
-        Figures~\ref{fig:one-popA}--\ref{fig:one-popC}.}
-        \aprcomment{$w$ and $f$ need to match text}
-        \aprcomment{Add the SHC model, and the ``vary SD'' fig as a panel here}
-        \aprcomment{First figure should just be two left panels. Combine two
-        right panels to figure 2.}
+        \textbf{Additive genetic variance under stabilizing selection.}
+        (A) Stabilizing selection acts to maintain phenotypic values of individuals
+        in the population near some optimum. Throughout, we assume a Gaussian fitness
+        function.
+        (B) With low mutational variance ($V_M$), the expected additive genetic
+        variance ($V_G$) is proportional to the population-scaled mutation rate.
+        When $V_M$ is large so that mutational effects can be strong, $V_G$ is
+        independent of the mutation rate. The stochastic house-of-cards model
+        (Eq.~\ref{eq:SHC}) interpolates these regimes, assuming steady-state
+        dynamics \citep{burger1989much}. Expected $V_G$ computed using an SFS
+        \citep[using \moments,][]{jouganous2017inferring} matches
+        simulations assuming no linkage between loci affecting the trait, which
+        align closely the stochastic house-of-cards approximation. Here, the
+        mutation rate $\mu=0.01$ (per haploid), $V_S=1$, and $N_e=10,000$.
+        Mutation effects are normally distributed with mean zero and variance
+        $V_M$.
     }
@@
 
-In diverged populations, genetic variation contributing to a trait under
+In diverged populations, the genetic variation contributing to a trait under
 stabilizing selection has a higher rate of turnover compared to neutrally
-evolving loci \citep{yair2022population}. We might therefore expect an
-accumulation of fixed differences in the \emph{H. sapiens} and Neanderthal
-lineages at loci contributing to such a trait, even when the mean phenotype in
-each population remains close to the same trait optimum. When a derived allele
-at high frequency in one population is introduced to another population in
-which is was previously absent, it will be at low frequency and subsequently
-selected against. Likewise, if the ancestral allele is reintroduced to a
-population fixed for a derived allele, the ancestral allele will be at low
-frequency and will be selected against. In either case we should expect
-selection against the introgressed allele, whether ancestral or derived and
-regardless the historical relative sizes of the populations involved. This
-prediction contrasts with the population-genetics ``load'' model, in which
-haplotypes with fewer deleterious variants (such as those from the population
-with larger historical size) are favored after introgression in either
-direction.
+evolving loci \citep{yair2022population}. We might therefore expect a rapid
+divergence of trait architecture and an accumulation of fixed differences in
+the \emph{H. sapiens} and Neanderthal lineages at trait-contributing loci, even
+when the mean phenotype in each population remains close to the same trait
+optimum. When a derived allele at high frequency in one population is
+introduced to another population in which is was previously absent, it will be
+at low frequency (if the admixture proportion is low) and subsequently selected
+against. Likewise, if the ancestral allele is reintroduced to a population
+fixed for a derived allele, the ancestral allele will be at low frequency and
+will be selected against. In either case we should expect selection against the
+introgressed allele, whether ancestral or derived and regardless the historical
+relative sizes of the populations involved. This prediction contrasts with the
+population-genetics ``load'' model, in which haplotypes with fewer deleterious
+variants (such as those from the population with larger historical size) are
+favored after introgression in either direction.
 
-\subsection*{Model}
+In this article, we show that admixture between diverged populations results in
+selection against the minor parental ancestry at trait-affecting loci. We
+develop a numerical approach based on the site frequency spectrum to predict
+genetic variation under complex demographic scenarios, and which we use to
+partition predicted trait heritability by introgressed vs. non-introgressed
+variation. Using simulations with linkage, we demonstrate that deserts of
+introgressed ancestry form around trait-contributing loci. When gene flow
+occurs bidirectionally, such deserts will tend to overlap in location in each
+population. As we argue in the Discussion, stabilizing selection on shared
+trait optima provides an explanation for the overlap of introgressed ancestry
+deserts in human and Neanderthal genomes after reciprocal introgression,
+recently observed in \citet{harris2023diverse}.
 
-We consider a polygenic trait for which an individual's (additive) genetic
-value is the sum over all effects of alleles in their genome, so that for
-individual $i$, \(G_i = \sum_l g_{l, i} a_l\), where $a_l$ is the effect size
-of the derived allele at locus $l$, and \(g_{l,i}\in\{0,1,2\}\) is their
-genotype at that locus (i.e., the number of alleles they carry at that locus).
-With no linkage between trait-affecting loci, the expected additive genetic
-variance is \(V_A = \sum_l 2p_l(1-p_l)a_l^2\), where $p_l$ is the allele
-frequency at locus $l$. We ignore dominance and epistatic effects, so that
-\(V_G = V_A\).  We further ingore environmental effects
-\citep{simons2018population}, so that the phenotypic variance \(V_P=V_A\).
+\section*{Model and Methods}
+
+We consider a polygenic trait for which an individual's additive genetic value
+is the sum over all effects of alleles in their genome. For individual $i$,
+\(G_i = \sum_l g_{l, i} a_l\), where $a_l$ is the effect size of the derived
+allele at locus $l$, and \(g_{l,i}\in\{0,1,2\}\) is their genotype at that
+locus (i.e., the number of alleles they carry at that locus). With no linkage
+between trait-affecting loci, the expected additive genetic variance is \(V_A =
+\sum_l 2p_l(1-p_l)a_l^2\), where $p_l$ is the allele frequency at locus $l$. We
+ignore dominance and epistatic effects, so that \(V_G = V_A\). We further
+ingore environmental effects \citep{simons2018population}, so that the
+phenotypic variance \(V_P=V_G\).
 
@@
 (Figure~\ref{fig:stab-sel}A) so that relative fitness is given by \(f(G_i | O,
-V_S) = \exp{(-(G_i - O)^2 / 2 V_S)}\). $V_S$ can be interpreted as the strength
-of selection on the trait, where larger $V_S$ implies weaker selection. For a
-population with mean phenotype at (or very close to) the optimum, the mean
-fitness of the population (assuming a roughly normal distribution of phenotypic
-values in the population) is \[\bar{w} \approx \int_{-\infty}^\infty f(G | 0,
-V_S) \mathcal{N}(0, V_G) dG = \left(\frac{V_S}{V_S+V_G}\right)^{1/2}.\] Thus,
-as the genetic variance increases, mean fitness among individuals in the
+V_S) = \exp{(-(G_i - O)^2 / 2 V_S)}\). $\frac{1}{V_S}$ is interpreted as the
+strength of selection on the trait, so larger $V_S$ implies weaker selection.
+For a population with mean phenotype at (or very close to) the optimum, the
+mean fitness of the population (assuming a roughly normal distribution of
+phenotypic values in the population) is
+\[\bar{w} \approx \int_{-\infty}^\infty
+f(G | 0, V_S) \mathcal{N}(0, V_G) dG =
+\left(\frac{V_S}{V_S+V_G}\right)^{1/2}.\]
+Thus, as the genetic variance increases, mean fitness among individuals in the
 population decreases.
 
-\subsubsection*{Mutation rates, effect sizes and genetic variance}
+\subsection*{Mutation rates, effect sizes and genetic variance}
 
@@
 
-Mutations are not expected to each have the same effect size $a$, but rather
-follow some distribution. Throughout, we assume effect sizes are drawn from a
-normal distribution with mean 0 and given variance $V_M$. When population sizes
-or effect sizes are small, drift dominates the dynamics of $V_G$, and at steady
-state \citet{lande1976natural} found \[V_G \approx 4 N_e \mu V_M.\]
-Interpolating between the selection- and drift-dominated regimes,
+Mutations are not expected to each have the same effect size $|a|$, but rather
+follow some distribution. Here, when mutation effects are not constant, we
+assume effect sizes are drawn from a normal distribution with mean 0 and given
+variance $V_M$. When population sizes or effect sizes are small, drift
+dominates the dynamics of $V_G$, and at steady state \citet{lande1976natural}
+found \[V_G \approx 4 N_e \mu V_M.\] Interpolating between the selection- and
+drift-dominated regimes,
 \begin{align}\label{eq:SHC}
-    V_G & \approx \frac{4 \mu V_S}{1 + \frac{VS}{N_e V_M}}.
+    V_G & \approx \frac{4 \mu V_S}{1 + \frac{V_S}{N_e V_M}}.
 \end{align}
@@
 
-\subsubsection*{Approximating allelic dynamics via underdominance}
+\subsection*{Approximating allelic dynamics via underdominance}
 
-We model the distribution of allele frequencies underlying the trait using
-the approximation that alleles are subject to symmetric underdominant
-selection \citep{robertson1956effect,keightley1988quantitative}.
-For an allele with (heterozygote) additive effect size $a$,
-the selection coefficient
-\[s\approx \frac{a^2}{2(V_S + V_G)} \approx \frac{a^2}{2V_S},\]
-if $V_G \ll V_S$ \citep[e.g.,][]{simons2018population}. Allelic dynamics can
-be modeled using the diffusion approximation, where the expected change
-in mean allele frequency per generation is
-\[\E[M_{\delta_p}] \approx -s p(1-p)(1-2p),\]
-and the expected change in variance of the allele frequency per generation is
-\[\E[V_{\delta_p}] \approx \frac{1}{4N_e}p(1-p).\]
-%From this, because $s$ is always positive for any $a\not=0$, we see that
-%selection pushes allele frequencies to zero if $p<1/2$ and to one if $p>1/2$.
-%$p=1/2$ is an unstable equilibrium.
+As initially shown by \citet{robertson1956effect} \citep[see
+also,][]{keightley1988quantitative, simons2018population}, when the mean
+phenotype of the population is close to the optimal phenotypic value,
+stabilizing selection results in symmetric underdominant selection on alleles
+contributing additively to the trait. Thus, for an allele at frequency $p$ with
+selection coefficient $s$, the expected change in $p$ over one generation is
+\begin{equation}
+    \E[\Delta p] = -sp(1-p)(1-2p).
+\end{equation}
+The selection coefficent $s$ depends on the strength of selection on the trait
+$V_S$ and the effect size $a$, as 
+\begin{equation}
+    s\approx \frac{a^2}{2(V_S+V_G)} \approx \frac{a^2}{2V_S}.
+\end{equation}
+Details are shown in Appendix~\ref{sec:underdominance}.
 
-We extend the moments-based solution for the sample site-frequency spectrum
+\subsubsection*{Computing expected genetic variance from the SFS}
+
+We model allelic dynamics using the diffusion approximation, where the expected
+change in mean allele frequency per generation is
+\[\E[M_{\delta_p}] = E[\Delta p] \approx -s p(1-p)(1-2p),\]
+as shown above, and the expected change in variance of the allele frequency per
+generation is
+\[\E[V_{\delta_p}] \approx \frac{1}{2N_e}p(1-p).\]
+Because $s$ is always positive for any $a\not=0$, we see that selection pushes
+allele frequencies to zero if $p<1/2$ and to one if $p>1/2$, resulting in
+symmetric underdominance.
+
+We extend the \moments-based solution for the sample site-frequency spectrum
 (SFS) \citep{jouganous2017inferring} to include underdominance with given
-selection coefficient $s$ (\(=a^2/2(V_S+V_G)\)). The contribution of alleles
-with effect size $a$ to the total genetic variance $V_G$ is then found by
-computing the expected pairwise diversity from the SFS (with sample size $n$,
-$\Phi_n$), as
-\[V_{G,a} = 2a^2\sum_{j=1}^{n-1}\frac{j(n-j)}{n(n-1)}\Phi_n(j|a,\mu_a)da.\]
-Here, $\mu_a$ is the mutation rate of alleles with effect size $a$. Then
-assuming a normal distribution of effect sizes for new mutations, the total
-genetic variance is
-\[V_G=\int_{-\infty}^\infty V_{G,a}\mathcal{N}(0,V_M).\]
+selection coefficient $s$ (Appendix~\ref{sec:moments-underdominance}). The
+contribution of alleles with effect size $a$ to the total genetic variance
+$V_G$ is found by computing the expected pairwise diversity from the SFS (with
+sample size $n$, denoted $\Phi_n$), as \[V_{G,a} \approx 2a^2\sum_{j=1}^{n-1}
+\frac{j(n-j)}{n(n-1)} \Phi_n(j|a,\mu_a) da.\] Here, $\mu_a$ is the mutation
+rate of alleles with effect size $a$. Then assuming a normal distribution of
+effect sizes for new mutations, the total genetic variance is \[V_G \approx
+\int_{-\infty}^\infty V_{G,a}\mathcal{N}(0,V_M).\]
 
 If $V_G$ is non-negligible compared to $V_S$, ignoring $V_G$ and using
-$s=a^2/2V_S$ leads to lower estimates of additive genetic variance
-(Figure~\ref{fig:toy-admixture}B). When mutation rates are large so that $V_G$
-is not small compared to $V_S$, using $s=a^2/2(V_S+V_G)$ provides estimates of
+$s=a^2/2V_S$ leads to deflated estimates of additive genetic variance
+(Figure~\ref{fig:supp-one-pop}). When mutation rates are large so that $V_G$ is
+not small compared to $V_S$, using $s=a^2/2(V_S+V_G)$ provides estimates of
 $V_G$ that closely match simulations assuming free recombination between loci
-(Figure~\ref{fig:toy-admixture}X,Y). Because $V_G$ can change over time, this
-means that $s$ is no longer constant, but can change due to factors such as
-non-constant demography that increase or reduce $V_G$.
+(Figures~\ref{fig:toy-admixture} and
+\ref{fig:one-pop-0.01}--\ref{fig:one-pop-0.1}). Because $V_G$ can vary over
+time, this means that $s$ is no longer constant and can change due to factors
+such as non-constant demography that increase or reduce $V_G$.
 
@@
     \centering
-    \includegraphics{../figures/reciprocal_admixture.pdf}
+    \includegraphics{../figures/one_pop.pdf}
     \caption{
-        \textbf{Predicting the genetic variance of a trait under stabilizing
-        seletion and non-steady state demography.}
-        \aprcomment{Make it in generations?}
-        \aprcomment{Right two panels should go to supplement, move single-pop
-        with size change from figure 1 to here.}
+        \textbf{Population size changes and additive genetic variance.}
+        We used \moments to track expected genetic variance of a trait
+        under stabilizing selection as the population chages size. (A) The
+        population goes through a 10-fold size reduction, followed by a recovery.
+        (B) All mutations had effect sizes $\pm0.01$, $0.04$, or $0.1$, with equal
+        probabilitiy of being trait increasing or decreasing. Depending on the
+        selection coefficient ($s\approx a^2/2V_S$) compared to $N_e$, $V_G$
+        could increase or decrease after a sudden population size change.
+        Predictions using \moments match simulations with free recombination
+        between trait-affeting loci
+        (Figures~\ref{fig:one-pop-0.01}--\ref{fig:one-pop-0.1}).
     }
-    \label{fig:toy-admixture}
+    \label{fig:one-pop}
 \end{figure}
@@
 subsequent introgression events. We focus on parameter regimes relevant to
-human-Neanderthal history. In a simple toy model -- which is not meant to
+human-Neanderthal history. In a simple toy model (which is not meant to
 perfectly match any particular known or inferred history, but instead
 demonstrate the effect of reciprocal introgression following population
-divergence -- a population of size $N_e=10\,000$ splits into two, one remaining
-size $10\,000$ and the other shrinking to size $1\,000$. They remain isolated
+divergence) a population of size $N_e=10{,}000$ splits into two, one remaining
+size $10{,}000$ and the other shrinking to size $1{,}000$. They remain isolated
 for $2N_e$ generations (or 500 thousand years, assuming an average generation
@@
 contributing $5\%$ ancestry to the recipient population
-(Figure~\ref{fig:toy-admixture}X,Y).
+(Figure~\ref{fig:toy-admixture}A,B).
 
 The second model is meant to more closely resemble inferred human-Neanderthal
-history, in which the ancestral population of size $N_e=10\,000$ splits into
-the human and Neanderthal branches. At 250ka, an early human-to-Neanderthal
+history, in which the ancestral population of size $N_e=10{,}000$ splits into the
+human and Neanderthal branches. At 250ka, an early human-to-Neanderthal
 introgression contributes $5\%$ ancestry to Neanderthals. The human branch
-shrinks to size $1\,000$ 60ka, followed by exponential growth to $20\,000$ at
-present time. Neanderthals contribute $2\%$ ancestry to this bottlenecked and
-expanding human population at 50ka, after which they go extinct
-(Figure~\ref{fig:neand-to-human}A).
+shrinks to size $1{,}000$ 60ka, followed by exponential growth to size $20,000$
+at present time. Neanderthals contribute $2\%$ ancestry to this bottlenecked
+and expanding human population at 50ka, after which they go extinct
+(Figure~\ref{fig:human-neand-h2}A).
 
@@
 
-\begin{figure}[t!]
+\begin{figure}[tb!]
     \centering
-    \includegraphics{../figures/neanderthal_admixture.pdf}
+    \includegraphics{../figures/reciprocal_admixture.pdf}
     \caption{
-        \textbf{}
-        \aprcomment{Show total/introgressed/non-introgressed in both directions
-        with both $\sigma_M$.}
+        \textbf{Genetic variance of a trait under stabilizing seletion with
+        multi-population demography.}
+        (A,B) Two simple demographic models, in which Deme 1 has size 10,000
+        and Deme 2 has size 1,000. In each scenario, we allow $5\%$ admixture
+        from one deme to the other after begin isolated for 20,000 ($2N_e$)
+        generations.
+        (C,D) We compared predicted (additive) $V_G$ from the site frequency
+        spectra (using \moments) to simulations
+        without linkage between trait-affecting loci. In Deme 2, $V_G$ decreases
+        after divergence and population size reduction. After admixture in both
+        directions, $V_G$ increases and then decays again toward levels at steady
+        state. Predictions from \moments closely match observed $V_G$ in
+        simulations.
+        Here, mutations were drawn from a normal distribution of fitnesses
+        with $V_M=0.0025$.
+        Other parameters: $\mu=0.025$, $V_S=1$, and the optimal phenotype remained
+        the same in each population.
+        (E,F) Using \moments, we partitioned additive contributions to $V_G$ by
+        mutations that were previously segregating in the focal population,
+        introgressed variants (either derived or reintroduced ancestral alleles),
+        and by new mutations since the time of admixture. In both demographic
+        scenarios, introgressed variants contribute a substantial proportion to
+        $V_G$, though it is primarily composed of introduced derived alleles when
+        admixture is from the small to the large population, while primarily
+        reintroduced ancestral alleles under the reverse direction of gene flow.
+        In both cases, $V_G$ is increasingly due to new mutations and the genetic
+        architecture of the trait turns over with time since admixture.
     }
-    \label{fig:neand-to-human}
+    \label{fig:toy-admixture}
 \end{figure}
+
+\subsubsection*{Simulations with free recombination}
+
+We compare our \moments-based predictions for $V_G$ in non-equilibrium settings
+to simulations assuming linkage equilibrium as well as individual-based
+simulations with recombination (next section). For both, we allow mutations
+to arise at rate $mu$ per haploid genome copy per generation. These are drawn
+from some distribution of mutation effects, which then contribute to individual
+trait values.
+
+To simulate allele frequency changes under free recombination (assuming linkage
+equilibrium between all trait-affecting alleles at all times), for each focal
+segregating allele we integrate over possible genetic backgrounds contributed
+by all other segregating alleles. We make the assumption that many alleles
+contribute to the trait, so the variance in genetic backgrounds is normally
+distributed around the mean genetic value $E_{\tilde{G}}=\sum_l 2p_l a_l$ with
+variance $V_{\tilde{G}}=\sum_l 2p_l(1-p_l)a_l^2$, where the sums omit the focal
+locus. The expected change in frequency is then computed using the approach
+outlined in Appendix~\ref{sec:underdominance} (without taking the first-order
+Taylor series approximation for the exponentials). Allele counts in the next
+generation are then binomially sampled with parameter $p+\E[\Delta p]$
+independently for each allele.
+
+\subsubsection*{Individual-based simulations with linkage}
+
+To include the effects of linkage between multiple selected alleles or selected
+and neutral alleles, we used \fwdpy \citep{thornton2019polygenic} to run
+Wright-Fisher simulations under the same demographic models considered above.
+In these simulations, we consider large (1 Morgan, or 100 Mb with a per-base
+recombination rate of $10^-8$) chromosomes with a uniform recombination
+landscape and specify per-haploid mutation rates. Mutations may be uniformly
+distributed across the chromosome, as in Figures~\ref{fig:linkage} and
+\ref{fig:supp-low-VM}--\ref{fig:supp-high-VM}, or they may fall within
+functional regions, as in Figures~\ref{fig:deserts} and
+\ref{fig:deserts-directional}. In these simulations, such regions are centered
+2 Mb apart and are 100kb in size, so that there are 50 evenly spaced regions
+across the chromosome. Mutation effect sizes are drawn either as constant
+values $\pm a$, or from a normal distribution with given mutational variance
+$V_M$.
+
+In these simulations, we track the empirical phenotypic variance (since there
+is no simulated environmental effect, this is equivalent to $V_G$) in each
+population each generation. We measure the effects of linked selection on
+(neutral) introgressed ancestry using \tskit to analyze genealogical
+information \citep{ralph2020efficiently}. To measure the reduction in
+introgressed ancestry around introgressed variants (as shown in
+Figure~\ref{fig:linkage}), we find each locus with a fixed difference between
+the parental populations at the time of admixture by preserving the generation
+immediately preceding admixture. For such a locus, for each sample we determine
+which (preserved) parental population its ancestry traces to at varying
+distances from the selected locus. Ancestry proportions are then averaged over
+each fixed difference. To measure proportions of introgressed ancestry or
+probabilities of observed ancestry deserts (as in Figure~\ref{fig:deserts}), we
+take the same approach to average ancestral source population proportions in 50
+kb windows. Ancestry deserts are defined as any such window with no ancestry
+inherited from the source population of introgression events.
+
+\subsection*{Data and code availability}
+
+All code to run analyses, create figures, and compile this manuscript are available
+at \url{https://github.com/apragsdale/neanderthal_stabilizing_selection}.
 
@@
 population (labeled 1), \(p_l = fp_{l,0} + (1-f)p_{l,1}\). Plugging into the
-expression for $V_G$ and after some simple algebra (Appendix,
-section~\ref{sec:VG-admixture}), we can express the expected genetic variance
+expression for $V_G$ and after some simple algebra
+(Appendix~\ref{sec:VG-admixture}), we can express the expected genetic variance
 directly after admixture as \[V_G = fV_{G,0} + (1-f) V_{G,1} + 2f(1-f)\sum_l
@@
 approximation for the joint SFS with underdominance. Comparing to simulations
-with free recombination between loci (Appendix,
-section~\ref{sec:VG-admixture}), we find that this provides an excellent fit to
-average observed $V_G$ over time (Figure~\ref{fig:toy-admixture}C,D). In
-general, admixture causes a sudden increase in $V_G$ followed by a fairly rapid
-return to pre-admixture levels.
+with free recombination between loci (Methods), we find that this provides an
+excellent fit to average observed $V_G$ over time
+(Figure~\ref{fig:toy-admixture}C,D). In general, admixture causes a sudden
+increase in $V_G$ followed by a fairly rapid return to pre-admixture levels.
 
@@
 were introduced through introgression, and new mutations since the time of
-admixture (Figure~\ref{fig:toy-admixture}). In many scenarios, the initial
+admixture (Figure~\ref{fig:toy-admixture}E,F). In many scenarios, the initial
 divergence of the two populations will have occurred long enough before the
@@
 
+\begin{figure}[t!]
+    \centering
+    \includegraphics{../figures/h2-per-SNP.pdf}
+    \caption{
+        \textbf{Allele contributions to heritability under human-Neanderthal
+        reciprocal introgression.}
+        (A) A simple model of divergence and admixture between humans and
+        Neanderthals. Using \moments, we computed predicted $V_G$ over time,
+        partitioned by variation that was introgressed vs. non-introgressed
+        (Figure~\ref{fig:human-neand-VG}).
+        (B-E) Predicted per-SNP contributions to genetic variance ($h^2$ per
+        SNP) is plotted over the 50 thousand years following introgression.
+        For non-introgressed variants, we also plot $h^2$ per SNP weighted by
+        allele frequencies matching those of introgressed variants. These
+        are shown for (B,C) Neanderthal-to-human introgression 50 kya,
+        (D,E) human-to-Neanderthal introgression 250 kya, (B,D) $V_M=0.0025$,
+        and (C,E) $V_M=0.0001$.
+    }
+    \label{fig:human-neand-h2}
+\end{figure}
+
 We used a historical model more closely resembling inferred human-Neanderthal
-history (Figure~\ref{fig:neand-to-human}A) to explore the effects of population
+history (Figure~\ref{fig:human-neand-h2}A) to explore the effects of population
 size changes and reciprocal admixture on the additive genetic architecture of
@@
 segregating, fixed or lost in either parental populations or if they arose as
-new mutations since the time of admixture (Methods Section X). This allows us
-to partition $V_G$ by contributions from these different sets of mutations. At
-low introgression proportions, as modeled here, the majority of $V_G$ is still
-contributed by previously segregating, non-introgressed mutations (Figure XAB).
-$V_G$ due to existing mutations decays monotonically over time and is fairly
-rapidly replace by new mutations (Figure XCD).
+new mutations since the time of admixture. This allows us to partition $V_G$ by
+contributions from these different sets of mutations
+(Figure~\ref{fig:human-neand-VG}). At low introgression proportions, as modeled
+here, the majority of $V_G$ is still contributed by previously segregating,
+non-introgressed mutations. $V_G$ due to existing mutations decays
+monotonically over time and is fairly rapidly replace by new mutations.
 
@@
 that of non-introgressed variants. These contributions change over time,
-depending on mutational variance (Figure YA vs B), as well as demography
-(comparing the two directions of introgression, Figure YC D). When weighting
-$h^2$ per SNP of introgressed SNPs by matching to allele frequencies of
-non-introgressed variants (Methods), relative contributions depend sensitively
-on evolutionary parameters and the time since admixture.
-
-\aprcomment{In discussion, point out that this makes interpreting results, such
-as \cite{wei2023lingering}, challenging.}
-
-\begin{figure}[t!]
-    \centering
-    \includegraphics{../figures/human_admixture.pdf}
-    \caption{
-        \textbf{}
-        \aprcomment{Put $h^2$ per SNP into its own figure.}
-    }
-    \label{fig:human-to-neand}
-\end{figure}
+depending on mutational variance, as well as demography
+(Figure~\ref{fig:human-neand-h2}). When weighting $h^2$-per-SNP of
+non-introgressed SNPs by matching to allele frequencies of introgressed
+variants, relative contributions depend sensitively on evolutionary parameters
+and the time since admixture.
 
@@
 between trait-affecting alleles. The inclusion of linkage can lead to
-noticeable effects on expected $V_G$, so that Equation~\ref{eq:SHC} slightly
-overestimates $V_G$ at steady state
-\citep{burger1989much,burger1994distribution,walsh2018evolution}.
+noticeable effects on expected $V_G$, so that Equation~\ref{eq:SHC} differs
+from observed $V_G$ at steady state
+\citep{burger1989much, burger1994distribution, walsh2018evolution}.
 
 To investigate the effects of linkage, we used chromosome-scale
-individual-based simulations. By varying the mutation rate and the variance of
-effect sizes of new mutations, we include scenarios ranging from low to high
-polygenicity and from weak to strong selection on individual alleles. In this
-and the following sections, we will highlight two effects of linkage. First,
-we observe deviations of $V_G$ from expectations without linkage, which can be
-large for highly polygenic traits. Second, selection on introgressed
-trait-affecting alleles results in a reduction of introgressed ancestry in
-surrounding regions.
+individual-based simulations \citep{thornton2019polygenic}. By varying the
+mutation rate and the variance of effect sizes of new mutations, we include
+scenarios ranging from low to high polygenicity and from weak to strong
+selection on individual alleles. In this and the following sections, we will
+highlight two effects of linkage. First, we observe deviations of $V_G$ from
+expectations without linkage, which can be large for highly polygenic traits.
+Second, selection on introgressed trait-affecting alleles results in a
+reduction of introgressed ancestry in surrounding regions.
 
@@
 \citep{bulmer1971effect}. This has the effect of reducing $V_G$. For lower
-mutational variances ($2N_e V_M\approx1$, Figure~\ref{fig:low_VM}), we observe
-this phenomenon. With low mutational input, and thus low polygenicity, $V_G$ in
-individual simulations closely matches expectations without linkage. As we
-increase the mutation rate, $V_G$ is reduced relative to those expectations.
-However, when the mutational variance is much larger ($2Ne V_m \approx 50$), we
-see the reverse trend (Figure~\ref{fig:high_VM}). At low mutation rate, there
-is a close match between observed $V_G$ and unlinked expectations, although
-simulated values are slightly higher. As the mutation rate increases, $V_G$
-increases to be relatively much larger that expectations, rather than smaller.
+mutational variances ($2N_e V_M\approx1$, Figure~\ref{fig:supp-low-VM}), we
+observe this phenomenon. With low mutational input, and thus low polygenicity,
+$V_G$ in individual simulations closely matches expectations without linkage.
+As we increase the mutation rate, $V_G$ is reduced relative to those
+expectations.  However, when the mutational variance is much larger
+($2N_eV_m\approx 50$), we see the reverse trend
+(Figure~\ref{fig:supp-high-VM}). At low mutation rate, there is a close match
+between observed $V_G$ and unlinked expectations, although simulated values are
+slightly higher. As the mutation rate increases, $V_G$ increases to be
+relatively much larger than expectations, rather than smaller.
 
@@
 \citep{hill1966effect}.
+
+\begin{figure}[t!]
+    \centering
+    \includegraphics{../figures/ancestry_reduction.pdf}
+    \caption{
+        \textbf{Reduced introgressed ancestry around alleles contributing to a
+        trait under stabilizing selection.}
+        Using a deterministic model
+        (Equations~\ref{eq:system-p}--\ref{eq:system-D}), we model the
+        frequencies of an introgressed trait-affecting allele and a linked
+        neutral allele, initially absent from the recipient population so that
+        their frequencies equal the introgression proportion ($f=0.05$).
+        The neutral allele frequency is reduced at a rate that depends on
+        the effect size of the selected allele and the probability of
+        recombination between them. LD (as measured by $D=Cov(p,q)$) decays
+        to zero over time.
+        (C) Alleles with strong effects are expected to result in a larger
+        depletion of introgressed ancestry around the selected locus.
+        (D,E) Compared to individual-based simulations, deterministic model
+        predicts the dip in introgressed ancestry around trait-affecting loci,
+        when the mutation rate is low, so that trait-affecting are sparsely
+        distributed. When mutation rates (and thus polygenicity) are high,
+        selected alleles are close together, so that selective interference
+        is more pronounced and local ancestry is affected by multiple selected
+        alleles.
+    }
+    \label{fig:linkage}
+\end{figure}
 
@@
 LD after admixture for given effect size $a$ and recombination rate $r$
-(Figure~\ref{fig:linkage-pred}). As expected, smaller effect sizes result in a
+(Figure~\ref{fig:linkage}A,B). As expected, smaller effect sizes result in a
 slower decay in introgressed ancestry frequency at both selected and linked
@@
 from the selected allele dynamics. Thus, the expected reduction in introgressed
-ancestry is largest for larger effect sizes (Figure~\ref{fig:linkage-pred}X),
+ancestry is largest for larger effect sizes (Figure~\ref{fig:linkage}C),
 and LD between the selected and linked neutral alleles is largest for neutral
-sites closest to the selected allele (Figure~\ref{fig:linkage-pred}Y).
+sites closest to the selected allele (Figure~\ref{fig:supp-LD}).
 
@@
 
-\begin{figure}[t!]
-    \centering
-    \includegraphics{../figures/linkage_predictions.pdf}
-    \caption{
-        \textbf{}
-        \aprcomment{Linked allele frequency should always be minor, and labeled
-        introgressed ancestry frequency}
-        \aprcomment{Combine with next figure}
-    }
-    \label{fig:linkage-pred}
-\end{figure}
+\subsection*{Introgressed ancestry deserts are shared under stabilizing
+selection and reciprocal introgression}
 
@@
     \centering
-    \includegraphics{../figures/linkage_simulation.pdf}
+    \includegraphics{../figures/introgression_deserts.SD_0.02.pdf}
     \caption{
-        \textbf{}
-        \aprcomment{Combine with previous figure}
+        \textbf{Stabilizing selection causes an enrichment of introgressed
+        ancestry deserts in functional regions.} (A) In chromosome-scale
+        simulations under a demography with reciprocal migration between humans
+        and Neanderthals (Figure~\ref{fig:human-neand-h2}A), stabilizing
+        selection causes a chromosome-wide reduction of introgressed ancestry
+        (below the $5\%$ and $2\%$ introgression proportions). This depletion
+        is most pronounced in ``functional'' regions that allowed for
+        trait-affecting mutations. (B) Introgressed ancestry deserts are
+        more likely to occur in such functional regions, as are
+        \emph{shared} deserts when compared across samples from humans and
+        Neanderthals.
     }
-    \label{fig:linkage-sim}
+    \label{fig:deserts}
 \end{figure}
-
-\subsection*{Introgressed ancestry deserts are shared under stabilizing
-selection and reciprocal introgression}
 
@@
 model of human-Neanderthal reciprocal introgression
-\citep[Figure~\ref{fig:neand-to-human}A,][]{harris2023diverse}. Mutations
+\citep[Figure~\ref{fig:human-neand-h2}A,][]{harris2023diverse}. Mutations
 affecting a trait under stabilizing selection in each population could arise in
@@
     contiguous 50kb regions with no observed introgressed ancestry in the
-sample) within and immediately surrounding the functional regions. Functional
-regions also displayed an enrichment of \emph{shared} ancestry deserts, with
-the probability of observing ancestry deserts (either within a population or
-shared) decaying to background levels as the distance from the functional
-region increases.
-
-\begin{figure}[t!]
-    \centering
-    \includegraphics{../figures/introgression_deserts.SD_0.02.pdf}
-    \caption{
-        \textbf{}
-    }
-    \label{fig:deserts}
-\end{figure}
+sample) within and immediately surrounding the functional regions
+(Figure~\ref{fig:deserts}B). Functional regions also displayed an enrichment of
+\emph{shared} ancestry deserts, with the probability of observing ancestry
+deserts (either within a population or shared) decaying to background levels as
+the distance from the functional region increases.
 
@@
 
-Many phenotypic traits are under stabilizing selection
-\citep{hodgins2015gene,sanjak2018evidence,sella2019thinking}. This has
-motivated using stabilizing selection around a shared optimum as a null model
-for the dynamics of alleles affecting polygenic traits
-\citep{yair2022population}. Stabilizing selection has a distinct effect on
-trait-affecting alleles, characterized by symmetric underdominance
-\citep{robertson1956effect,keightley1988quantitative}, that is, selection
-against the minor allele. There has been some previous work on how migration
-shapes the genetic variance of a trait under stabilizing selection
-\aprcomment{[cite]}, but most studies consider single populations, often at
-steady state. Admixture and introgression events are common in natural
-populations, so understanding their effects on the architecture of complex
-traits is needed.
+Many phenotypic traits are under stabilizing selection \citep{hodgins2015gene,
+sanjak2018evidence, sella2019thinking}. This has motivated using stabilizing
+selection around a shared optimum as a null model for the dynamics of alleles
+affecting polygenic traits \citep{yair2022population}. Stabilizing selection
+has a distinct effect on trait-affecting alleles, characterized by symmetric
+underdominance \citep{robertson1956effect, keightley1988quantitative}, that is,
+selection against the minor allele. Some theoretical and simulation studies
+consider the effect of population differentiation and migration on the genetic
+architecture of a trait under stabilizing selection
+\citep[e.g.,][]{tufto2000quantitative, yeaman2011genetic, yair2022population},
+but most previous work has focused on single population scenarios, often
+assuming steady state dynamics. Admixture and introgression events are common
+in natural populations, so understanding their effects on the architecture of
+complex traits is needed.
 
@@
 effect sizes, demographic history, and the time since admixture
-(Figures~\ref{fig:neand-to-human}, \ref{fig:human-to-neand}).
-
-\aprcomment{While we have focused on the relatively simple scenario of two
-diverged populations recently exchanging migrants, SFS methods presented here
-can incorporate continuous migration, more than two populations, or changes in
-the trait optimum. We leave these for future work\dots}
+(Figures~\ref{fig:human-neand-h2}, \ref{fig:human-neand-VG}).
 
@@
 pre-existing trait-affecting alleles are replaced by new mutations, turning
-over the genetic architecture of the trait (Figure\ref{fig:toy-admixture}).
+over the genetic architecture of the trait (Figure~\ref{fig:toy-admixture}).
 Selection occurs against the minor alleles at trait-affecting loci, so that
@@
 reintroduced ancestral alleles, are selected against
-(Figure~\ref{fig:linkage-pred}).
+(Figure~\ref{fig:linkage}).
 
@@
 DNA in Neanderthals by human haplotypes after early human-to-Neanderthal
-introgression \citep{posth2017deeply,petr2020evolutionary}.
-%However, this is
-%not the general pattern observed when comparing introgressed ancestry in the
-%autosomal genomes of humans and Neanderthals \citep{harris2023diverse}.
+introgression \citep{posth2017deeply, petr2020evolutionary}, but does not broadly
+match observations across the autosomal genome \citep{harris2023diverse}.
 
 The classical model of Bateson-Dobzhansky-Muller incompatibilities (BDMIs)
-\citep{bateson1909heredity,dobzhansky1936studies,muller1942isolating} explains
-the accumulation of reproductive isolation through negative epistatic
+\citep{bateson1909heredity, dobzhansky1936studies, muller1942isolating}
+explains the accumulation of reproductive isolation through negative epistatic
 interactions that are exposed in hybrids. \citet{muller1942isolating}
@@
 distribution of BDMIs, studies point to interacting BDMI alleles being unlinked
-\citep[][e.g.,]{presgraves2003fine} and an asymmetry in the alleles under
-selection in different introgression scenarios
-\citep{maheshwari2011genetics,moran2021genomic}.
+\citep[e.g.,][]{presgraves2003fine} and an asymmetry in the alleles under
+selection in different introgression scenarios \citep{maheshwari2011genetics,
+moran2021genomic}.
 
@@
 which they interpret as evidence for the inital process of speciation between
-humans and Neanderthals. However, this observation of shared ancestry deserts
-does not match expectations under a classic model of BDMIs, as argued above.
+humans and Neanderthals. However, the observation of shared ancestry deserts
+does not match expectations under a classic model of BDMIs as described above.
 Instead, at least some of the pattern may be due to stabilizing selection
@@
 
-\section*{Methods}
-
-\subsection*{Computing expected genetic variance using the diffusion approximation}
-
-\subsection*{Simulations with free recombination}
-
-\subsection*{Individual-based simulations with linkage}
-
-\section*{Acknowledgements}
-\begin{itemize}
-    \item Lab members
-    \item Kevin Thornton
-    \item Bret Payseur
-\end{itemize}
+%\section*{Acknowledgements}
+%\begin{itemize}
+%    \item Lab members
+%    \item Kevin Thornton
+%    \item Bret Payseur
+%\end{itemize}
 
@@
 
+\section{Stabilizing selection results in underdominant selection on
+trait-affecting alleles} \label{sec:underdominance}
+
+This result is well known and has been derived many times before. We include it
+here for completeness. We consider a polygenic trait under stabilizing
+selection, and we assume the mean phenotype of the population is close to the
+optimal value (which will be the case if the optimum has not shifted recently).
+Because many segregating loci are assumed to contribute to the phenotypic
+variance of the trait, the phenotypic distribution is well-approximated as
+normally distributed around the optimum: \(f(G)=\frac{1}{\sqrt{2\pi V_G}}
+e^{-G^2/2V_G}\). We assume a Gaussian fitness function, so that the fitness
+of an individual with genotypic value $G$ is given by \(w(G)=e^{-G^2/2V_S}\).
+
+For a derived allele (labeled 1) with frequency $p$, the expected change in
+frequency over one generation due to selection in general is
+\[\E[\Delta p]=p(1-p)\frac{w_{1\cdot}-w_{0\cdot}}{\bar{w}},\]
+where \(\bar{w}\) is the mean fitness of the population, and marginal fitnesses
+for the derived and ancestral alleles are
+\[w_{1\cdot}=p w_{11} + (1-p)w_{01},\]
+and
+\[w_{0\cdot}=p w_{01} + (1-p)w_{00},\]
+respectively. Here, $w_{11}$, $w_{01}$, and $w_{00}$ are the relative fitnesses
+of individuals who are homozygous for the derived allele, heterozygous, or
+homozygous for the ancestral allele, respectively.
+
+We partition an individual's genetic values into contributions from their
+genetic background and the focal locus: $G=\tilde{G}+xa$, where $x\in{0,1,2}$.
+We then integrate over genetic backgrounds to find the expected fitnesses of
+each genotype. As described in \citet[][SI section 2.1]{simons2018population},
+because each individual locus contributes a small amount to $V_G$,
+$V_{\tilde{G}}\approx V_G$ and the fitness function is well approximated as
+\(f(\tilde{G})=\frac{1}{\sqrt{2\pi V_G}}e^{-\tilde{G}^2/2V_G}\). Then,
+mean fitness
+\[
+    \bar{w} = \sqrt{\frac{V_S}{V_S+V_G}},
+\]
+as shown in the main text, and
+\[ 
+    w_{11} \approx f(\tilde{G}) w(\tilde{G} + 2a) d\tilde{G}
+    = \bar{w}\exp{\left(\frac{-4a^2(1-p)^2}{2(V_S+V_G)}\right)},
+\]
+\[
+    w_{01} \approx f(\tilde{G}) w(\tilde{G} + a) d\tilde{G}
+    = \bar{w}\exp{\left(\frac{-a^2(1-2p)^2}{2(V_S+V_G)}\right)},
+\]
+and
+\[
+    w_{00} \approx f(\tilde{G}) w(\tilde{G}) d\tilde{G}
+    = \bar{w}\exp{\left(\frac{-4a^2p^2}{2(V_S+V_G)}\right)}.
+\]
+
+Taking the first-order Taylor series expansion of the exponentials
+($e^{-x}\approx1-x$, for small $x$) and combining terms,
+\[
+    \frac{w_{1\cdot} - w_{0\cdot}}{\bar{w}} \approx \frac{-a^2(1-2p)}{2(V_S+V_G)},
+\]
+demonstrating the result.
+
+\section{Moment equations for over- and underdominance}
+\label{sec:moments-underdominance}
+
+For symmetric underdominance, the relative fitnesses of genotypes \(aa:Aa:AA\)
+are \(1:1-s:1\). We consider any value of $s$ (positive or negative,
+corresponding to either under- or overdominance); with stabilizing selection,
+\(s=\frac{a^2}{2V_S}\). We extended \moments \citep{jouganous2017inferring} to
+compute the sample site frequency spectrum \(\Phi_{\mathbf{n}}\) for one or
+more (up to five) populations with sample sizes \(\mathbf{n}\). This provides a
+good approximation for the distribution of trait-affecting allele frequencies
+across multiple populations if the trait optimum is shared across populations
+and each population's mean phenotype remains close to that optimum. This is
+expected to be the case if there are no optimum shifts in any lineage.
+Hardy-Weinberg equilibrium is assumed at all loci. Accounting for optimum
+shifts in one or more lineages would require a combination of direct and
+underdominant selection \citep[e.g.,][]{hayward2022polygenic}, which we leave
+for future work.
+
+We refer readers to \citet{jouganous2017inferring} for a detailed introduction
+to the general moments-based framework for the dynamics of \(\Phi_\mathbf{n}\).
+Here, we describe how underdominant selection changes \(\Phi\) over a single
+generation. With symmetric underdominance, selection acts heterozygotes. This
+can be formulated as some proportion of heterozygotes failing to reproduce in a
+given generation, with those selected lineages replaced by copies drawn from
+the full population.
+
+In a single population, we consider $n$ tracked lineages of which $i$ of those
+carry the derived allele (\(\Phi_n(i)\) is thus the count of loci with $i$
+observed derived alleles in a haploid sample of size of $n$). $i$ can increase
+or decrease due to selection ``events''. Here, we assume $s$ is small enough so
+that at most a single selection event occurs among the $n$ lineages in any
+given generation. This is a reasonable approximation as long as $s$ is not
+extremely large \citep{jouganous2017inferring} -- for selection coefficients
+induced by stabilizing selection, this is typically a safe assumption.
+
+Two selective events can change $i$ to $i+1$ or $i-1$: (a) a tracked copy
+carrying a derived allele is heterozygous (paired with an ancestral
+allele-carrying copy), selected against, and then replaced by an ancestral
+allele drawn from the rest of the population, so that \(i\rightarrow i-1\), or
+(b) a tracked copy carrying the ancestral allele is paired with a derived
+allele-carrying copy, selected against, and then replaced by a derived allele,
+so that \(i\rightarrow i+1\). In both cases, we require drawing two additional
+lineages to find the \(\Phi_n\) in the next generation, one for the diploid
+pair and one for the replacement allele (i.e., \(\Phi_n^{t+1}(i)\) requires
+\(\Phi_{n+2}^t\)). This results in an unclosed system of equations, and we use
+a quadratic jackknife approximation to approximate \(\Phi_{n+2}\) from
+\(\Phi_n\), as described in \citet{jouganous2017inferring}.
+
+For case (a), \(i\rightarrow i-1\) (i.e., \(\Phi_n(i)\) is reduced) with
+probability \[s\frac{i(n-i+2)(n-i+1)}{(n+2)(n+1)}\Phi_{n+2}(i),\] and
+\(i+1\rightarrow i\) (\(\Phi_n(i)\) increases) with probability
+\[s\frac{(i+1)(n-i+1)(n-i)}{(n+2)(n+1)}\Phi_{n+2}(i+1).\] For case (b),
+\(i\rightarrow i+1\) with probability
+\[s\frac{(n-i)(i+2)(i+1)}{(n+2)(n+1)}\Phi_{n+2}(i+2),\] and \(i-1 \rightarrow
+i\) with probability \[s\frac{(n-i+1)(i+1)i}{(n+2)(n+1)}\Phi_{n+2}(i+1).\]
+These can be combined (with negative rates for the reduction of \(\Phi_n(i)\))
+to describe the change of \(\Phi_n(i)\) for all \(0\leq i \leq n\).
+Figures~\ref{fig:underdominance-validation-small-s} and
+\ref{fig:underdominance-validation-large-s} show that this approach,
+implemented in \moments, is accurate compared to discrete Wright-Fisher
+simulations.
+
 \section{Additive genetic variance after admixture}\label{sec:VG-admixture}
 
-\aprcomment{Start with ``Here we show that\dots''}
-
-Suppose two population (labeled 0 and 1) diverged some time in the past and
-then admix in proportions $f$ and $1-f$.
+Here, we derive the expected genetic variance after admixture between two
+source populations (Equation~\ref{eq:VG-admixture}). Suppose two population
+(labeled 0 and 1) diverged some time in the past and then admix in proportions
+$f$ and $1-f$.
 
@@
     + \frac{1}{2}V_{G,1} + \sum_l F_{2,l}(0,1)a_l^2\right] \\
-    & = \left(f^2 + f(1-f)\right) V_{G,0} + \left((1-f) + f(1-f)\right)V_{G,1}
+    & = \left(f^2 + f(1-f)\right) V_{G,0} + \left((1-f)^2 + f(1-f)\right)V_{G,1}
     + 2f(1-f) \sum_l F_{2,i}(0,1)a_l^2 \\
@@
 
-\begin{figure}[tb!]
-    \centering
-    %\includegraphics{}
-    \caption{
-        \textbf{Dynamics of additive genetic variance in two populations with admixture.}
-        Coupled with the equations for the decay of $V_G$ due to drift and selection,
-        and the increase in $V_G$ through mutation, show that the expected dynamics
-        are matched by simulations without linkage.
-    }
-    \label{fig:VG_dynamics}
-\end{figure}
-
-\section{Dynamics of introgressed alleles}
-
-\section{Reciprocal depletion of functional variation}
-
-While incompatibilities require two or more loci (Morgan, 1942), introgression
-with stabilizing selection can create the appearance of a single-locus
-incompatibility via underdominance\dots
-
 \end{document}

### 2023-12-14 doc/supplement.tex
update some figures
edits to intro
updates to model and results
update doc, directional deserts

@@
 \usepackage{booktabs}
+\usepackage{bm}
 
@@
 
+\usepackage{xspace}
+\newcommand{\moments}{\texttt{moments}\xspace}
+\newcommand{\fwdpy}{\texttt{fwdpy11}\xspace}
+
 \title{Supporting Information}
@@
 
+\renewcommand{\thesection}{S\arabic{section}}
+\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\thefigure}{S\arabic{figure}}
+
 \section{Supplemental Figures}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics{../figures/one_pop_supp.pdf}
+    \caption{
+        Single-population dynamics when $V_G$ is not small.
+        Solid lines: expected $V_G$ assuming $s=a^2/2V_S$. Dashed lines:
+        expected $V_G$ assuming $s=a^2/2(V_S+V_G)$. When mutation rates
+        are large (here, $\mu=0.01$, $V_S=1$), $V_G$ is non-negligible
+        compared to $V_S$. By having to account for $V_G$ in the
+        translation of effect size to selection coefficient, it makes
+        $s$ non-constant if $V_G$ changes over time (due to non-constant
+        demography, for example). In this case, $s$ must be updated
+        regularly, given the current state of the population.
+        Here, the demographic history is a bottleneck followed by a
+        recovery, as depicted in Figure~\ref{fig:one-pop}A.
+    }
+    \label{fig:supp-one-pop}
+\end{figure}
 
@@
     \caption{
-        \textbf{Simulations assuming no linkage with weak effects.}
-        All mutations have $a=0.01$.
+        Simulations with no linkage and weak effects.
+        Total mutation rate was $\mu=0.025$ and all mutations had effect
+        sizes $a=\pm 0.01$. The demographic history (bottleneck and
+        recovery) is depicted in Figure~\ref{fig:one-pop}A. The predicted
+        trajectory of additive genetic variance using \moments was found
+        assuming symmetric underdominant selection on trait-affecting
+        alleles, with selection coefficient $s=a^2/2(V_S+V_G)$.
     }
-    \label{fig:one-popA}
+    \label{fig:one-pop-0.01}
 \end{figure}
@@
     \caption{
-        \textbf{Simulations assuming no linkage with moderate effects.}
-        All mutations have $a=0.01$.
+        Simulations with no linkage and moderate effects.
+        All parameters were consistent with Figure~\ref{fig:one-pop-0.01},
+        but with effect sizes $a=\pm0.04$.
     }
-    \label{fig:one-popB}
+    \label{fig:one-pop-0.04}
 \end{figure}
@@
     \caption{
-        \textbf{Simulations assuming no linkage with strong effects.}
-        All mutations have $a=0.1$.
+        Simulations with no linkage and strong effects.
+        All parameters were consistent with Figure~\ref{fig:one-pop-0.01}, 
+        but with effect sizes $a=\pm0.1$. In each comparison (weak, moderate,
+        and strong effect sizes), predictions from \moments closely match
+        observed average additive genetic variance from simulations.
     }
-    \label{fig:one-popC}
+    \label{fig:one-pop-0.1}
+\end{figure}
+
+\begin{figure}[t!]
+    \centering
+    \includegraphics{../figures/human-neand-VG.pdf}
+    \caption{
+        Genetic variance of a trait under stabilizing selection in a model with
+        human (blue) and Neanderthal (orange) demography. We consider two
+        mutational variances ($\sigma_M=0.05$ and $0.01$). The demographic model
+        (Figure~\ref{fig:human-neand-h2}A) includes $5\%$ admixture from humans
+        to Neanderthals 250 ka, and $2\%$ admixture from Neanderthals to humans
+        50 ka.
+    }
+    \label{fig:human-neand-VG}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics{../figures/model_comparison.SD_0.01.mu_0.001_0.025.pdf}
+    \caption{
+        With low mutational variance ($V_M=0.0001$), individual-based simulations
+        using \fwdpy closely match our predictions using \moments, when the
+        mutation rate is small (Left). When the mutation rate is increased,
+        polygenicity increases and unlinked expectations from \moments deviate
+        from observed $V_G$ in individual-based simulations. In this case, with
+        relatively small mutational variance, the observed $V_G$ is reduced in
+        comparison, consistent with the Bulmber (1971) effect.
+    }
+    \label{fig:supp-low-VM}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering 
+    \includegraphics{../figures/model_comparison.SD_0.05.mu_0.001_0.025.pdf}
+    \caption{
+        With high mutational variance ($V_M=0.0025$), strong deviations are
+        observed between individual-based simulations and expectations without
+        linkage. When the mutational input is large, observed $V_G$ can be
+        much larger than expectations without linkage, possibly consistent with
+        widespread interference between alleles.
+    }
+    \label{fig:supp-high-VM}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics{../figures/supp-LD.pdf}
+    \caption{
+        Linkage disequilibrium between the a trait-affecting and neutral allele, 
+        2000 generations after admixture. Inititally, the introgression proportion
+        was $0.05$. Mutations with stronger effect sizes, and thus stronger selection
+        against them, decrease in frequency more rapidly, leading to reduced LD
+        as measured by $D=Cov(p,q)$ (Figure~\ref{fig:linkage}).
+    }
+    \label{fig:supp-LD}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics[width=0.6\textwidth]{../figures/underdominance.Ne_10000.s_-5e-05.pdf}
+    \caption{
+        Comparison of predicted (\moments) and simulated SFS with underdominant
+        selection. In this comparison, $N_e=10^4$ and $s=-5\times10^{-5}$, so that
+        the population-size scale selection coefficient $\gamma=-1$.
+        Simulations were performed under a Wright-Fisher model without linkage.
+    }
+    \label{fig:underdominance-validation-small-s}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics[width=0.6\textwidth]{../figures/underdominance.Ne_10000.s_-0.001.pdf}
+    \caption{
+        Comparison of predicted (\moments) and simulated SFS with underdominant
+        selection. In this comparison, $N_e=10^4$ and $s=-1\times10^{-3}$, so that
+        the population-size scale selection coefficient $\gamma=-20$.
+        Simulations were performed under a Wright-Fisher model without linkage.
+    }
+    \label{fig:underdominance-validation-large-s}
 \end{figure}

### 2023-12-14 desert_overlap/neanderthal_deserts.directional.py

@@ -1,0 +1,403 @@
+"""
+Run the fwdpy11 simulation and get ancestry proportions around introgressed
+alleles that were fixed differences between populations.
+"""
+
+
+import fwdpy11
+import tskit
+import numpy as np
+from dataclasses import dataclass
+from typing import List
+from collections import defaultdict
+import pickle
+import gzip
+import time
+import demes
+import sys, os
+
+
+def get_samples_dict(ts):
+    samples_dict = defaultdict(list)
+    for s in ts.samples():
+        samples_dict[(ts.node(s).time, ts.node(s).population)].append(s)
+    return samples_dict
+
+
+def get_ancestry_counts(ts):
+    # preprocess and set up storage arrays
+    samples_dict = get_samples_dict(ts)
+
+    max_t = max([k[0] for k in samples_dict.keys()])
+    ts2 = ts.decapitate(max_t)
+
+    num_samples = len(samples_dict[min(samples_dict.keys())])
+
+    sample_counts = np.zeros(ts2.num_nodes, dtype=int)
+    parents = -np.ones(ts2.num_nodes, dtype=int)
+
+    for s in samples_dict[min(samples_dict.keys())]:
+        sample_counts[s] += 1
+
+    num_seg = np.zeros(ts2.num_trees)
+    num_intro = np.zeros(ts2.num_trees)
+
+    for i, ed in enumerate(ts2.edge_diffs()):
+        interval, edges_out, edges_in = ed
+        # take care of edges out
+        for e_out in edges_out:
+            c = e_out.child
+            p = e_out.parent
+            assert parents[c] == p
+            parents[c] = -1
+            sample_counts[p] -= sample_counts[c]
+            while parents[p] != -1:
+                c = p
+                p = parents[c]
+                sample_counts[p] -= sample_counts[e_out.child]
+
+        # update storage with edges in
+        for e_in in edges_in:
+            c = e_in.child
+            p = e_in.parent
+            assert parents[c] == -1
+            parents[c] = p
+            sample_counts[p] += sample_counts[c]
+            while parents[p] != -1:
+                c = p
+                p = parents[c]
+                sample_counts[p] += sample_counts[e_in.child]
+        # count up ancestry proportions in this segment
[…]

### 2023-12-14 figures/linkage_reduction_model.py

@@ -47,17 +47,17 @@ def plot_split_trajectories(a, r, gens, f, fig, ax, VS=1, legend=True):
     D0 = f * (1 - f)
     p, q, D = trajectories(f, q0, D0, generations=gens, r=r, s=a ** 2 / 2 / VS)
 
-    ax.plot(p, label="Selected allele frequency")
-    ax.plot(q, label="Introgressed ancestry frequency")
-    ax.plot(D, label="$|D|$ ($D=Cov(p,q)$)")
+    ax.plot(p)
+    ax.plot(q)
+    ax.plot(D)
     ax.set_ylim(0, f * 1.5)
-    ax.set_xlabel("Generations")
+    ax.set_xlabel("Generations since admixture")
 
     # initially fully linked
     p, q, D = trajectories(1 - f, q0, -D0, generations=gens, r=r, s=a ** 2 / 2 / VS)
-    axb.plot(p, label="Selected allele frequency")
-    axb.plot(q, label="Introgressed ancestry frequency")
-    axb.plot(D, label="$|D|$ ($D=Cov(p,q)$)")
+    axb.plot(p, label="Selected\nallele freq.")
+    axb.plot(q, label="Introgressed\nallele freq.")
+    axb.plot(D, label="LD $(|D|)$")
     axb.set_ylim(1 - f * 1.5, 1)
     if legend:
         axb.legend(frameon=False)
@@ -84,9 +84,11 @@ def frequency_dip(a_vals, r_max, gens, p0, ax_q, ax_D, VS=1, legend=True):
     for a in a_vals:
         qs = np.zeros(len(rs))
         Ds = np.zeros(len(rs))
-        s = a**2 / 2 / VS
+        s = a ** 2 / 2 / VS
         for i, r_dist in enumerate(rs):
-            p, q, D = trajectories(p0, p0, p0 - p0**2, generations=gens, r=r_dist, s=s)
+            p, q, D = trajectories(
+                p0, p0, p0 - p0 ** 2, generations=gens, r=r_dist, s=s
+            )
             qs[i] = q[-1]
             Ds[i] = D[-1]
         qs = np.concatenate((qs[:0:-1], qs))
[…]

### 2023-12-14 figures/neanderthal_admixure_figure.py

@@ -1,304 +1,0 @@
-import numpy as np
-import matplotlib.pylab as plt, matplotlib
-import moments
-import demes, demesdraw
-import scipy.stats
-import os, pickle
-import sys
-
-
-from single_population_stabilizing_selection import *
-
-# set font sizes
-plt.rcParams["legend.title_fontsize"] = "xx-small"
-matplotlib.rc("xtick", labelsize=6)
-matplotlib.rc("ytick", labelsize=6)
-matplotlib.rc("axes", labelsize=7)
-matplotlib.rc("axes", titlesize=7)
-matplotlib.rc("legend", fontsize=6)
-
-
-colors = plt.rcParams["axes.prop_cycle"].by_key()["color"]
-
-gen = 25
-Ne = 1e4
-N_N = 1e3
-T_split = 600000
-
-T_H_to_N = 250000
-T_N_to_H = 50000
-f_H_to_N = 0.05
-f_N_to_H = 0.02
-
-T_bottle = 60000
-N_bottle = 1000
-N_Eur = 20000
-
-b = demes.Builder(time_units="years", generation_time=25)
-b.add_deme(
-    "Human",
-    epochs=[
-        dict(end_time=T_bottle, start_size=Ne),
-        dict(end_time=0, start_size=N_bottle, end_size=N_Eur),
-    ],
-)
-b.add_deme(
-    "Neanderthal",
-    start_time=T_split,
-    ancestors=["Human"],
-    epochs=[dict(end_time=45000, start_size=N_N)],
-)
-b.add_pulse(
-    sources=["Human"], dest="Neanderthal", time=T_H_to_N, proportions=[f_H_to_N]
-)
-b.add_pulse(
-    sources=["Neanderthal"], dest="Human", time=T_N_to_H, proportions=[f_N_to_H]
-)
-g = b.resolve()
-
-grid = (3, 8)
-
-fig = plt.figure(3, figsize=(6.5, 4.5))
-fig.clf()
-
-# plot the model
-ax1 = plt.subplot2grid(grid, (0, 0), colspan=2, rowspan=2)
-demesdraw.tubes(g, ax=ax1, labels=None)
-ax1.set_ylabel("Time ago (ka)")
-ax1.set_xlabel("       Human         Neand")
[…]

### 2023-12-14 figures/single_population_stabilizing_selection.py

@@ -16,21 +16,6 @@ matplotlib.rc("legend", fontsize=6)
 
 colors = plt.rcParams["axes.prop_cycle"].by_key()["color"]
 
-Ne = 1e4
-NB = 1e3
-T1 = 2 * Ne
-T2 = 2 * Ne
-b = demes.Builder(time_units="generations")
-b.add_deme(
-    "Deme",
-    epochs=[
-        dict(end_time=T1 + T2, start_size=Ne),
-        dict(end_time=T2, start_size=NB),
-        dict(end_time=0, start_size=Ne),
-    ],
-)
-g = b.resolve()
-
 
 def get_VG(fs, a):
     return fs.pi() * a ** 2
@@ -48,7 +33,7 @@ def init_fs(n, gamma, theta):
     )
 
 
-def get_time_series_VA(a, VS=1, mu=0.025, n=200, update_sel=False):
+def get_time_series_VA(a, Ne, NB, T1, T2, VS=1, mu=0.025, n=200, update_sel=False):
     # initial sfs
     gamma = get_gamma(a, 0, Ne, VS=VS)
     theta = 4 * Ne * mu
@@ -123,159 +108,203 @@ def equilibrium_VG(SD):
     return VG
 
 
-if __name__ == "__main__":
-    # stabilizing selection model
-
-    VS = 1
-    opt = optimum = 0
-    VG = 0.1
-    VS = 1
-
-    grid = (10, 2)
-    fig = plt.figure(1, figsize=(6.5, 4))
-
-    ax1 = plt.subplot2grid(grid, (0, 0), rowspan=5)
+def plot_stabilizing_selection_diagram(ax, VS=1, VG=0.1, optimum=0):
     z = np.linspace(-2, 2, 201)
     w = np.exp(-(z ** 2) / 2 / VS)
     f = np.exp(-(z ** 2) / 2 / VG)
-
-    ax1.plot(z, w, color=colors[0], label="$w(z)$")
-    ax1.plot(z, f, color=colors[1], label="$f(z)$")
-    ax1.vlines(0, 0, 1, colors="k", linestyles="--", lw=1)
-
-    ax1.annotate(
+    ax.plot(z, w, color=colors[0], label="Fitness\nfunction")
+    ax.plot(z, f, color=colors[1], label="Phenotypic\ndistribution")
+    ax.vlines(0, 0, 1, colors="k", linestyles="--", lw=1)
+    ax.annotate(
         "",
         xy=(0.43, 0.45),
         xytext=(0, 0.45),
[…]

### 2023-12-14 figures/toy_admixture.py

@@ -24,15 +24,16 @@ N2 = 1e3
 T1 = 500000 / 25
 T2 = 100000 / 25
 f = 0.05
-b = demes.Builder(time_units="years", generation_time=25)
+#b = demes.Builder(time_units="years", generation_time=25)
+b = demes.Builder(time_units="generations")
 b.add_deme("Deme1", epochs=[dict(end_time=0, start_size=N1)])
 b.add_deme(
     "Deme2",
-    start_time=(T1 + T2) * 25,
+    start_time=(T1 + T2),
     ancestors=["Deme1"],
     epochs=[dict(end_time=0, start_size=N2)],
 )
-b.add_pulse(sources=["Deme2"], dest="Deme1", time=T2 * 25, proportions=[f])
+b.add_pulse(sources=["Deme2"], dest="Deme1", time=T2, proportions=[f])
 g1 = b.resolve()
 
 b.data["pulses"][0]["sources"] = ["Deme1"]
@@ -250,7 +251,7 @@ if __name__ == "__main__":
     a_list = np.logspace(-3, np.log10(2 * SD))
 
     grid = (2, 8)
-    fig = plt.figure(2, figsize=(6.5, 3.5))
+    fig = plt.figure(3, figsize=(6.5, 3.5))
     fig.clf()
 
     ## from small to big
@@ -259,6 +260,8 @@ if __name__ == "__main__":
     # plot model
     ax1 = plt.subplot2grid(grid, (0, 0), colspan=2)
     demesdraw.tubes(g1, ax=ax1)
+    ax1.set_yticks(np.arange(0, 32000, 4000))
+    ax1.set_ylabel("Time ago (generations)")
 
     # plot simulation averages vs moments prediction
     ax2 = plt.subplot2grid(grid, (0, 2), colspan=3)
@@ -269,15 +272,17 @@ if __name__ == "__main__":
         theta=theta,
         admix_dest=admix_dest,
     )
-    t_bp = 2 * Ne * (t - t[-1]) / 1000
+    t_bp = 2 * Ne * (t - t[-1])
     # get simulation data
     VGs, VG_by_state = load_sim_data(admix_dest)
-    t_sim = -np.arange(len(VGs[0]))[::-1] / 1000
-    ax2.plot(25 * t_bp, VG2, ":", lw=2, c=colors[1], label=None)
[…]

### 2024-08-19 doc/manuscript.tex
paper edits

@@
 \usepackage{url}
+\usepackage{authblk}
+\usepackage{bm}
 
@@
 \DeclareMathOperator{\Var}{Var}
+\DeclareMathOperator{\Cov}{Cov}
 \newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
@@
 \title{Archaic introgression and the distribution
-of shared functional variation under stabilizing selection}
-\author{APR}
+of shared variation under stabilizing selection}
+\author[]{Aaron P. Ragsdale}
+\affil[]{Department of Integrative Biology, University of Wisconsin--Madison, WI, USA}
 \date{\today}
@@
 
-    Many phenotypic traits are under stabilizing selection, which maintains
-    individual phenotypic values near some optimum. Classical theory shows that
-    at loci contributing to the genetic variance of a trait under stabilizing
-    selection, the minor allele is selected against via symmetric
-    underdominance. Most studies of the dynamics of the genetic architecture
-    and variance of a trait under stabilizing selection have focused on single
-    populations. However, natural populations are typically structured in some
-    way, and admixture and introgression are common occurances, in particular
-    in human evolution. Because stabilizing selection results in selection
-    against the minor allele at a trait-affecting locus, the alleles from the
-    minor parental ancestry will be selected against after admixture. Here, we
-    show that the site frequency spectrum can be used to model the genetic
-    architecture of such traits, allowing for complex multi-population
-    histories including admixture. We develop a simple model with linkage to a
-    neutral locus to show that introgressed ancestry is expected to be depleted
-    around trait-affecting alleles. Using this model and individual-based
-    simulations, we show that introgressed ancestry deserts tend to form around
-    such loci. When introgression between two diverged populations occurs in
-    both directions, as has been inferred between humans and Neanderthals,
-    introgressed ancestry deserts will often be shared across populations.
-    Stabilizing selection for a shared phenotypic optimum may therefore explain
+    Many phenotypic traits are under stabilizing selection, which maintains a
+    population's mean phenotypic value near some optimum. The dynamics of
+    traits and trait architectures under stabilizing selection have bene
+    extensively studied for single populations at steady state. However,
+    natural populations are seldom at steady state. Instead, they are often
+    structured in some way, and admixture and introgression events may be
+    common, including over human evolutionary history. Because stabilizing
+    selection results in selection against the minor allele at a
+    trait-affecting locus, alleles from the minor parental ancestry will be
+    selected against after admixture. Here, we show that the site-frequency
+    spectrum can be used to model the genetic architecture of such traits,
+    allowing for the study of trait architecture dynamics in complex
+    multi-population settings. We use a simple deterministic two-locus model to
+    predict the reduction of introgressed ancestry around trait-contributing
+    loci. From this and individual-based simulations, we show that
+    introgressed-ancestry deserts are enriched around such loci. When
+    introgression between two diverged populations occurs in both directions,
+    as has been inferred between humans and Neanderthals, the locations of
+    introgressed-ancestry deserts will tend to be shared across populations. We
+    argue that stabilizing selection for shared phenotypic optima may explain
     recent observations in which regions of depleted human-introgressed
@@
 Genomic surveys of natural systems show that historical admixture among
-diverged populations or closely related taxa commonly occurs
+diverged populations and closely related taxa commonly occurs
 \citep{brandvain2014speciation, skoglund2015ancient, suvorov2022widespread} and
-is potentially widespread in primate \citep{tung2017contribution,
-sorensen2023genome} and hominin \citep{wolf2018outstanding, peter2020100}
-evolution. Admixture may therefore be an important driver of phenotypic and
-molecular variation and shapes the genetic basis of complex traits. In humans,
-archaic introgression involving Neanderthals and Denisovans has attracted
-considerable attention, including efforts to describe the historical processes
-leading to observed distributions of introgressed DNA in present-day
-populations \citep{prufer2014complete, villanea2019multiple,
-chen2020identifying} and the contribution of introgressed variation to
-quantitative traits \citep{sankararaman2016combined, wei2023lingering}.
+is widespread in primate \citep{tung2017contribution, sorensen2023genome} and
+hominin \citep{wolf2018outstanding, peter2020100} evolution. Admixture is
+therefore a frequent driver of phenotypic and molecular variation and can
+contribute to the genetic architectures of complex traits. In humans, archaic
+introgression from Neanderthals and Denisovans has attracted considerable
+attention, including efforts to describe the historical processes leading to
+observed distributions of introgressed DNA in present-day populations
+\citep{prufer2014complete, villanea2019multiple, chen2020identifying} and the
+contribution of introgressed variation to quantitative traits
+\citep{sankararaman2016combined, wei2023lingering}.
 
@@
 against. Some introgressed haplotypes were likely positively selected in modern
-humans \citep{huerta2014altitude, racimo2017signatures, enard2018evidence,
-gower2021detecting}, possibly due to locally adaptive variation that provided
-fitness advantages as humans encountered novel environments. Despite some cases
-of adaptive introgression, mostly introduced alleles were likely to have been
-selected against in humans \citep{harris2016genetic, juric2016strength}. Since
+\emph{Homo sapiens} (here, ``humans'') \citep{huerta2014altitude,
+racimo2017signatures, enard2018evidence, gower2021detecting}, possibly due to
+locally adaptive variation that provided fitness advantages as humans
+encountered novel environments. Despite some cases of adaptive introgression,
+most introduced alleles were likely to have been selected against in humans
+\citep{harris2016genetic, juric2016strength, veller2023recombination}. Since
 Neanderthal and Denisovan population sizes were relatively small for hundreds
-of thousands of years, theory predicts they would have accumulated fixed
-deleterious variation at an increased rate. Introgressed haplotypes loaded with
-more deleterious mutations would then have quickly been selected against after
+of thousands of years, theory predicts they would have accumulated deleterious
+variation at an increased rate. Introgressed haplotypes loaded with more
+deleterious mutations would have been rapidly removed by selection after
 admixture. Mapping the distribution of Neanderthal-introgressed haplotypes in
@@
 
-There is growing genetic evidence that early \emph{H. sapiens} reciprocally
-contributed to Neanderthal genomes \citep{kuhlwilm2016ancient,
-hubisz2020mapping, harris2023diverse}. Such introgression would have occurred
-tens or hundreds of thousands of years prior to Neanderthal introgression in
-humans during the global dispersal of modern humans around 60 ka. This is
-supported by ``near modern'' \emph{H. sapiens} outside of African around
-120--100 ka or earlier \citep{schwarcz1988esr, grun2005u, beyer2021climatic},
-potentially overlapping with Neanderthals and providing opportunities for early
-contacts. While estimates of the genomic contribution of early \emph{H.
-sapiens} to Neanderthals vary, around $6\%$ of later Neanderthal genomes may
-have been contributed through introgression from humans to Neanderthals
-\citep{harris2023diverse}. If \emph{H. sapiens}-related haplotypes carried
-fewer deleterious alleles due to their larger long-term effective population
-size, human-introgressed DNA would have been favored in Neanderthal genomes.
-The replacement of Neanderthal mitochondrial and Y chromosomes by early human
-haplotypes supports this model of post-admixture selection in the Neanderthal
-lineage \citep{posth2017deeply, petr2020evolutionary}.
+There is growing genetic evidence that humans reciprocally contributed to
+Neanderthal genomes \citep{kuhlwilm2016ancient, hubisz2020mapping,
+harris2023diverse, li2024recurrent}. This gene flow occurred tens to hundreds
+of thousands of years prior to Neanderthal introgression in humans during the
+global dispersal of modern humans around 60 ka. This is supported by early
+\emph{H. sapiens} outside of African around 200--100 ka \citep{schwarcz1988esr,
+grun2005u, harvati2019apidima, beyer2021climatic}, potentially overlapping with
+Neanderthals and providing opportunities for early contacts. While estimates of
+the genomic contribution of early humans to Neanderthals vary, around $6\%$ of
+later Neanderthal genomes may trace through this admixture event
+\citep{harris2023diverse}. Under a ``load'' model, if human-related haplotypes
+carried fewer deleterious alleles due to their larger long-term effective
+population size, human-introgressed DNA would have been favored in Neanderthal
+genomes. The replacement of Neanderthal mitochondrial and Y chromosomes by
+early human haplotypes appears to support this model of post-admixture positive
+selection in the Neanderthal lineage \citep{posth2017deeply,
+petr2020evolutionary}.
 
 Models for selection against introgressed alleles are often based on load
-arguments (in particular, differences in the rate of accumulation of
+arguments (that is, due to differences in the rate of accumulation of
 unconditionally deleterious variation in populations of different sizes) or
-hybrid incompatibilities \citep{muller1942isolating}. These arguments, founded
-in population genetics theory, rarely take into account selection operating on
-phenotypic variation or the relationship between genetic and phenotypic
-variation. Many phenotypic traits are thought to be under stabilizing selection
+hybrid incompatibilities \citep{muller1942isolating}. These models, founded in
+population genetics theory, rarely take into account selection operating on
+phenotypic traits or the relationship between genetic and phenotypic variation.
+Many phenotypic traits are thought to be under stabilizing selection
 \citep{sanjak2018evidence, sella2019thinking}, including gene regulation
@@
 in regulatory regions \citep{sankararaman2014genomic}, stabilizing selection on
-quantitative traits is particularly relevant to the dynamics of functional
-genetic variation after introgression.
+quantitative traits may be particularly relevant to the dynamics of functional
+genetic variation after introgression among hominins.
 
@@
 (Figure~\ref{fig:stab-sel}A). When the mean phenotype of the population is
-close to the phenotypic optimum, classical models predict that trait-affecting
-alleles are subject to underdominant selection, i.e., selection against the
-minor allele at the locus \citep{robertson1956effect}. This has proven to be a
-useful model for understanding genetic architectures of traits under
-stabilizing selection in single-population settings
+close to the phenotypic optimum, classical models predict that the minor allele
+at a trait-affecting locus is selected against, with allele-frequency dynamics
+equivalent to underdominant selection \citep{robertson1956effect}. This has
+proven to be a useful model for understanding genetic architectures of traits
+under stabilizing selection in single-population settings
 \citep[e.g.,][]{keightley1988quantitative, simons2018population,
@@
         \textbf{Additive genetic variance under stabilizing selection.}
-        (A) Stabilizing selection acts to maintain phenotypic values of individuals
-        in the population near some optimum. Throughout, we assume a Gaussian fitness
-        function.
-        (B) With low mutational variance ($V_M$), the expected additive genetic
-        variance ($V_G$) is proportional to the population-scaled mutation rate.
-        When $V_M$ is large so that mutational effects can be strong, $V_G$ is
-        independent of the mutation rate. The stochastic house-of-cards model
-        (Eq.~\ref{eq:SHC}) interpolates these regimes, assuming steady-state
-        dynamics \citep{burger1989much}. Expected $V_G$ computed using an SFS
-        \citep[using \moments,][]{jouganous2017inferring} matches
-        simulations assuming no linkage between loci affecting the trait, which
-        align closely the stochastic house-of-cards approximation. Here, the
-        mutation rate $\mu=0.01$ (per haploid), $V_S=1$, and $N_e=10,000$.
-        Mutation effects are normally distributed with mean zero and variance
-        $V_M$.
+        (A) Stabilizing selection acts to maintain phenotypic values of
+        individuals in the population near some optimum. Throughout, we assume
+        a Gaussian fitness function.
+        (B) With low mutational variance ($V_M$), the expected additive
+        genetic variance ($V_G$) is proportional to the
+        population-scaled mutation rate. When $V_M$ is large so that
+        mutational effects can be strong, $V_G$ is independent of the mutation
+        rate. The stochastic house-of-cards model (Eq.~\ref{eq:SHC})
+        interpolates these regimes, assuming steady-state dynamics
+        \citep{burger1989much}. Expected $V_G$ computed using an SFS approach
+        \citep[developed here using \moments,][]{jouganous2017inferring}
+        matches simulations assuming no linkage between loci affecting the
+        trait, which align closely with the stochastic house-of-cards
+        approximation.
+        Here $V_S=1$, $N_e=10{,}000$, the mutation rate
+        $\mu=0.01$ (per haploid), and mutation effects are drawn from a normal
+        distribution with mean zero and variance $V_M$.
     }
@@
 In diverged populations, the genetic variation contributing to a trait under
-stabilizing selection has a higher rate of turnover compared to neutrally
-evolving loci \citep{yair2022population}. We might therefore expect a rapid
-divergence of trait architecture and an accumulation of fixed differences in
-the \emph{H. sapiens} and Neanderthal lineages at trait-contributing loci, even
-when the mean phenotype in each population remains close to the same trait
-optimum. When a derived allele at high frequency in one population is
-introduced to another population in which is was previously absent, it will be
-at low frequency (if the admixture proportion is low) and subsequently selected
-against. Likewise, if the ancestral allele is reintroduced to a population
-fixed for a derived allele, the ancestral allele will be at low frequency and
-will be selected against. In either case we should expect selection against the
-introgressed allele, whether ancestral or derived and regardless the historical
-relative sizes of the populations involved. This prediction contrasts with the
-population-genetics ``load'' model, in which haplotypes with fewer deleterious
-variants (such as those from the population with larger historical size) are
-favored after introgression in either direction.
+stabilizing selection has a higher rate of turnover compared to neutral
+evolution \citep{yair2022population}. We therefore expect a rapid divergence of
+trait architectures and an accumulation of fixed differences in the human and
+Neanderthal lineages at trait-contributing loci, even when the mean phenotype
+in each population remains close to the same trait optimum. When a derived
+allele at high frequency in one population is introduced to another population
+in which it was previously absent, it will be at low frequency (if the
+admixture proportion is low) and subsequently selected against. Likewise, if
+the ancestral allele is reintroduced to a population fixed for a derived
+allele, the ancestral allele will be at low frequency and will be selected
+against. Selection acts against the introgressed allele in either case, whether
+it is ancestral or derived and regardless of the historical relative sizes of
+the populations involved. In concurrent work, \citet{veller2024stabilizing}
+demonstrate this effect by deriving the expected decay of minor parental
+ancestry under stabilizing selection after admixture. This prediction contrasts
+with the population-genetics load model, in which haplotypes with fewer
+deleterious variants (such as those from the population with larger historical
+size) are favored after introgression in either direction.
 
 In this article, we show that admixture between diverged populations results in
-selection against the minor parental ancestry at trait-affecting loci. We
-develop a numerical approach based on the site frequency spectrum to predict
-genetic variation under complex demographic scenarios, and which we use to
-partition predicted trait heritability by introgressed vs. non-introgressed
-variation. Using simulations with linkage, we demonstrate that deserts of
-introgressed ancestry form around trait-contributing loci. When gene flow
-occurs bidirectionally, such deserts will tend to overlap in location in each
-population. As we argue in the Discussion, stabilizing selection on shared
-trait optima provides an explanation for the overlap of introgressed ancestry
-deserts in human and Neanderthal genomes after reciprocal introgression,
-recently observed in \citet{harris2023diverse}.
+selection against the minor parental ancestry at loci contributing to a trait
+under stabilizing selection. We develop a numerical approach based on the
+site-frequency spectrum to predict genetic variation under complex demographic
+scenarios, and which we use to partition predicted trait heritability by
+introgressed and non-introgressed variation. Using simulations with linkage, we
+demonstrate that deserts of introgressed ancestry form around
+trait-contributing loci. When gene flow occurs bidirectionally, such deserts
+will tend to overlap in location across populations. We argue that stabilizing
+selection on shared trait optima may explain the overlap of introgressed
+ancestry deserts in human and Neanderthal genomes after reciprocal
+introgression, as has been recently observed \citep{harris2023diverse}.
 
@@
 allele at locus $l$, and \(g_{l,i}\in\{0,1,2\}\) is their genotype at that
-locus (i.e., the number of alleles they carry at that locus). With no linkage
-between trait-affecting loci, the expected additive genetic variance is \(V_A =
-\sum_l 2p_l(1-p_l)a_l^2\), where $p_l$ is the allele frequency at locus $l$. We
-ignore dominance and epistatic effects, so that \(V_G = V_A\). We further
-ingore environmental effects \citep{simons2018population}, so that the
-phenotypic variance \(V_P=V_G\).
+locus (i.e., the number of derived alleles they carry at that locus). With
+linkage equilibrium between trait-affecting loci, the expected additive genetic
+variance is \(V_A = \sum_l 2p_l(1-p_l)a_l^2\), where $p_l$ is the allele
+frequency at locus $l$. We ignore dominance and epistatic effects (often argued
+to be a reasonable modeling choice, e.g., \citet{barton2002understanding,
+zhang2002pleiotropic}), so that \(V_G = V_A\). We further ignore environmental
+effects and noise \citep[under the assumption of additive environmental
+effects, as in][]{lande1975maintenance, simons2018population}, so that only
+genetic effects are considered.
 
 Stabilizing selection acts to reduce phenotypic variation around the optimum
-value $O$ (typally set to zero), and we assume a Gaussian fitness function
-(Figure~\ref{fig:stab-sel}A) so that relative fitness is given by \(f(G_i | O,
+value $O$, typically set to zero, and we assume a Gaussian fitness function
+(Figure~\ref{fig:stab-sel}A) so that relative fitness (relative to an
+individual with optimal phenotype) is given by \(f(G_i | O,
 V_S) = \exp{(-(G_i - O)^2 / 2 V_S)}\). $\frac{1}{V_S}$ is interpreted as the
 strength of selection on the trait, so larger $V_S$ implies weaker selection.
-For a population with mean phenotype at (or very close to) the optimum, the
-mean fitness of the population (assuming a roughly normal distribution of
-phenotypic values in the population) is
+For a population with mean phenotype at or very close to the optimum, the
+mean fitness of the population [assuming a normal distribution of phenotypic
+values in the population, \citep{turelli1994genetic, urban2013asymmetric}] is
 \[\bar{w} \approx \int_{-\infty}^\infty
-f(G | 0, V_S) \mathcal{N}(0, V_G) dG =
-\left(\frac{V_S}{V_S+V_G}\right)^{1/2}.\]
-Thus, as the genetic variance increases, mean fitness among individuals in the
-population decreases.
+f(G | O, V_S) \mathcal{N}(O, V_G) dG =
+\left(\frac{V_S}{V_S+V_G}\right)^{1/2},\]
+so that as the genetic variance increases, population mean fitness decreases.
 
@@
 If all alleles contribute equally to the trait
-(with effect sizes \(\pm a\) occurring in equal proportion),
+with effect sizes \(\pm a\) occurring in equal proportion,
 \citet{keightley1988quantitative} showed that the dynamics of $V_G$ can be
@@
 
-Mutations are not expected to each have the same effect size $|a|$, but rather
-follow some distribution. Here, when mutation effects are not constant, we
-assume effect sizes are drawn from a normal distribution with mean 0 and given
-variance $V_M$. When population sizes or effect sizes are small, drift
+Mutations will not generally all have the same effect size, but rather are
+drawn from some distribution. Here, when modeling mutation effects as
+non-constant, we assume effect sizes follow a normal distribution with mean 0
+and variance $V_M$. When population sizes or effect sizes are small, drift
 dominates the dynamics of $V_G$, and at steady state \citet{lande1976natural}
-found \[V_G \approx 4 N_e \mu V_M.\] Interpolating between the selection- and
-drift-dominated regimes,
+found \[V_G \approx 4 N_e \mu V_M.\] Interpolating between the drift- and
+selection-dominated regimes,
 \begin{align}\label{eq:SHC}
@@
 phenotype of the population is close to the optimal phenotypic value,
-stabilizing selection results in symmetric underdominant selection on alleles
-contributing additively to the trait. Thus, for an allele at frequency $p$ with
-selection coefficient $s$, the expected change in $p$ over one generation is
+stabilizing selection results in selection against the minor alleles at loci
+contributing additively to the trait, with dynamics mirroring symmetric
+underdominance. In general, for an allele at frequency $p$ with selection
+coefficient $s$, the expected change in $p$ over one generation is
 \begin{equation}
-    \E[\Delta p] = -sp(1-p)(1-2p).
+    \E[\Delta p] = sp(1-p).
 \end{equation}
-The selection coefficent $s$ depends on the strength of selection on the trait
-$V_S$ and the effect size $a$, as 
+
+In our case, the selection coefficient $s$ depends on the strength of selection
+on the trait $V_S$ and the effect size $a$, as well as the frequency of the
+allele, so that
 \begin{equation}
-    s\approx \frac{a^2}{2(V_S+V_G)} \approx \frac{a^2}{2V_S}.
+    s\approx -\frac{a^2(1-2p)}{2(V_S+V_G)} 
+    = \frac{\left(p-\frac{1}{2}\right)a^2}{V_S + V_G}
+    \approx \frac{\left(p-\frac{1}{2}\right)a^2}{V_S},
 \end{equation}
-Details are shown in Appendix~\ref{sec:underdominance}.
+when $V_G \ll V_S$. This model assumes linkage equilibrium between
+trait-contributing loci. We note that LD is expected to be generated between
+loci even in the absence of linkage \citep{bulmer1971effect}, and
+\citet{negm2024effect} recently showed how to correct for its effect on the
+selection dynamics of a focal allele.
+
+Because $a^2/(V_S+V_G)$ is always positive for any $a\not=0$, we see that
+selection pushes allele frequencies to zero if $p<1/2$ and to one if $p>1/2$,
+resulting in symmetric underdominance. Details are shown in
+Appendix~\ref{sec:underdominance}.
 
@@
 change in mean allele frequency per generation is
-\[\E[M_{\delta_p}] = E[\Delta p] \approx -s p(1-p)(1-2p),\]
-as shown above, and the expected change in variance of the allele frequency per
+\[\E[M_{\delta_p}] = E[\Delta p] \approx s p(1-p),\]
+with $s$ defined above, and the expected change in variance of the allele frequency per
 generation is
 \[\E[V_{\delta_p}] \approx \frac{1}{2N_e}p(1-p).\]
-Because $s$ is always positive for any $a\not=0$, we see that selection pushes
-allele frequencies to zero if $p<1/2$ and to one if $p>1/2$, resulting in
-symmetric underdominance.
-
-We extend the \moments-based solution for the sample site-frequency spectrum
+
+We extended the \moments-based solution for the sample site-frequency spectrum
 (SFS) \citep{jouganous2017inferring} to include underdominance with given
@@
 $V_G$ is found by computing the expected pairwise diversity from the SFS (with
-sample size $n$, denoted $\Phi_n$), as \[V_{G,a} \approx 2a^2\sum_{j=1}^{n-1}
-\frac{j(n-j)}{n(n-1)} \Phi_n(j|a,\mu_a) da.\] Here, $\mu_a$ is the mutation
-rate of alleles with effect size $a$. Then assuming a normal distribution of
-effect sizes for new mutations, the total genetic variance is \[V_G \approx
-\int_{-\infty}^\infty V_{G,a}\mathcal{N}(0,V_M).\]
+sample size $n$, denoted $\Phi_n$), as
+\[V_{G,a} \approx 2a^2\sum_{j=1}^{n-1} \frac{j(n-j)}{n(n-1)} \Phi_n(j|a,\mu_a) da.\]
+Here, $\mu_a$ is the mutation rate of alleles with effect size $a$. Then
+assuming a normal distribution of effect sizes for new mutations, the total
+genetic variance is
+\[V_G \approx \int_{-\infty}^\infty V_{G,a}\mathcal{N}(0,V_M).\]
+$V_G$ can be computed periodically in this way to predict trajectories in
+non-equilibrium settings (Figure~\ref{fig:one-pop}).
 
 If $V_G$ is non-negligible compared to $V_S$, ignoring $V_G$ and using
-$s=a^2/2V_S$ leads to deflated estimates of additive genetic variance
+$s=(p-1/2)a^2/V_S$ leads to deflated estimates of additive genetic variance
 (Figure~\ref{fig:supp-one-pop}). When mutation rates are large so that $V_G$ is
-not small compared to $V_S$, using $s=a^2/2(V_S+V_G)$ provides estimates of
-$V_G$ that closely match simulations assuming free recombination between loci
-(Figures~\ref{fig:toy-admixture} and
+not small compared to $V_S$, using $s=(p-1/2)a^2/(V_S+V_G)$ provides estimates
+of $V_G$ that closely match simulations assuming free recombination between
+loci (Figures~\ref{fig:toy-admixture} and
 \ref{fig:one-pop-0.01}--\ref{fig:one-pop-0.1}). Because $V_G$ can vary over
@@
         We used \moments to track expected genetic variance of a trait
-        under stabilizing selection as the population chages size. (A) The
+        under stabilizing selection as the population changes size. (A) The
         population goes through a 10-fold size reduction, followed by a recovery.
-        (B) All mutations had effect sizes $\pm0.01$, $0.04$, or $0.1$, with equal
-        probabilitiy of being trait increasing or decreasing. Depending on the
-        selection coefficient ($s\approx a^2/2V_S$) compared to $N_e$, $V_G$
+        (B) All mutations have effect sizes $\pm0.01$, $0.04$, or $0.1$, with equal
+        probability of being trait increasing or decreasing. Depending on the
+        selection coefficient ($s\approx (p-1/2)a^2/V_S$) compared to $N_e$, $V_G$
         could increase or decrease after a sudden population size change.
         Predictions using \moments match simulations with free recombination
-        between trait-affeting loci
+        between trait-affecting loci
         (Figures~\ref{fig:one-pop-0.01}--\ref{fig:one-pop-0.1}).
@@
 The second model is meant to more closely resemble inferred human-Neanderthal
-history, in which the ancestral population of size $N_e=10{,}000$ splits into the
-human and Neanderthal branches. At 250ka, an early human-to-Neanderthal
+history, in which the ancestral population of size $N_e=10{,}000$ splits at 600
+kya into the human and Neanderthal branches, with effective sizes $10{,}000$
+and $2{,}000$, respectively. At 250 ka, an early human-to-Neanderthal
 introgression contributes $5\%$ ancestry to Neanderthals. The human branch
-shrinks to size $1{,}000$ 60ka, followed by exponential growth to size $20,000$
-at present time. Neanderthals contribute $2\%$ ancestry to this bottlenecked
-and expanding human population at 50ka, after which they go extinct
-(Figure~\ref{fig:human-neand-h2}A).
-
-In each scenario, we track genetic diversity and phenotypic variance before and
-after admixture to understand allele frequency and haplotype dynamics. The
-trait optimum is kept at $0$ in all branches, so that no optimum shift occurs,
-and the strength of selection $V_S=1$ remains constant.
+shrinks to size $1{,}000$ 60 ka, followed by exponential growth to size
+$20,000$ at present time. Neanderthals contribute $2\%$ ancestry to this
+bottlenecked and expanding human population at 50 ka, after which they go
+extinct (Figure~\ref{fig:human-neand-h2}A).
+
+In each scenario, we tracked phenotypic variance and genetic variation to
+compare simulations to model predictions. The trait optimum was kept at $0$ in
+all populations (no optimum shift occurred), and the strength of selection
+$V_S=1$ remains constant.
 
@@
     \caption{
-        \textbf{Genetic variance of a trait under stabilizing seletion with
+        \textbf{Genetic variance of a trait under stabilizing selection with
         multi-population demography.}
@@
         generations.
-        (C,D) We compared predicted (additive) $V_G$ from the site frequency
+        (C,D) We compared predicted (additive) $V_G$ from the site-frequency
         spectra (using \moments) to simulations
@@
 
-We compare our \moments-based predictions for $V_G$ in non-equilibrium settings
-to simulations assuming linkage equilibrium as well as individual-based
-simulations with recombination (next section). For both, we allow mutations
-to arise at rate $mu$ per haploid genome copy per generation. These are drawn
-from some distribution of mutation effects, which then contribute to individual
-trait values.
+We compared our \moments-based predictions for $V_G$ in non-equilibrium
+settings to simulations assuming linkage equilibrium as well as
+individual-based simulations with recombination (next section). For both,
+mutations occur at rate $\mu$ per haploid genome copy per generation, with
+effects drawn from some distribution of mutation effects.
 
@@
 equilibrium between all trait-affecting alleles at all times), for each focal
-segregating allele we integrate over possible genetic backgrounds contributed
-by all other segregating alleles. We make the assumption that many alleles
-contribute to the trait, so the variance in genetic backgrounds is normally
+segregating allele we integrated over possible genetic backgrounds contributed
+by all other segregating alleles. Making the assumption that many alleles
+contribute to the trait, the variance in genetic backgrounds is normally
 distributed around the mean genetic value $E_{\tilde{G}}=\sum_l 2p_l a_l$ with
 variance $V_{\tilde{G}}=\sum_l 2p_l(1-p_l)a_l^2$, where the sums omit the focal
-locus. The expected change in frequency is then computed using the approach
+locus. The expected change in frequency was then computed using the approach
 outlined in Appendix~\ref{sec:underdominance} (without taking the first-order
 Taylor series approximation for the exponentials). Allele counts in the next
-generation are then binomially sampled with parameter $p+\E[\Delta p]$
+generation were then binomially sampled with parameter $p+\E[\Delta p]$
 independently for each allele.
@@
 and neutral alleles, we used \fwdpy \citep{thornton2019polygenic} to run
-Wright-Fisher simulations under the same demographic models considered above.
-In these simulations, we consider large (1 Morgan, or 100 Mb with a per-base
-recombination rate of $10^-8$) chromosomes with a uniform recombination
-landscape and specify per-haploid mutation rates. Mutations may be uniformly
-distributed across the chromosome, as in Figures~\ref{fig:linkage} and
-\ref{fig:supp-low-VM}--\ref{fig:supp-high-VM}, or they may fall within
-functional regions, as in Figures~\ref{fig:deserts} and
-\ref{fig:deserts-directional}. In these simulations, such regions are centered
-2 Mb apart and are 100kb in size, so that there are 50 evenly spaced regions
-across the chromosome. Mutation effect sizes are drawn either as constant
-values $\pm a$, or from a normal distribution with given mutational variance
-$V_M$.
-
-In these simulations, we track the empirical phenotypic variance (since there
+Wright-Fisher simulations under the demographic models described above. In
+these simulations, we considered large (1 Morgan, or 100 Mb with a per-base
+recombination rate of $10^{-8}$) chromosomes with a uniform recombination
+landscape. Trait-affecting mutations were either uniformly distributed across
+the chromosome (Figures~\ref{fig:linkage} and
+\ref{fig:supp-low-VM}--\ref{fig:supp-high-VM}) or fell within functional
+regions (Figures~\ref{fig:deserts} and
+\ref{fig:deserts-high-VM}--\ref{fig:deserts-directional-B}). For the latter,
+such regions were centered 2 Mb apart and were 100kb in size, so that there
+were 50 evenly spaced regions across the chromosome. Mutation effect sizes were
+drawn either as constant values $\pm a$, or from a normal distribution with
+mutational variance $V_M$.
+
+In these simulations, we tracked the empirical phenotypic variance (since there
 is no simulated environmental effect, this is equivalent to $V_G$) in each
-population each generation. We measure the effects of linked selection on
-(neutral) introgressed ancestry using \tskit to analyze genealogical
+population each generation. We measured the effects of linked selection on
+neutral introgressed ancestry using \tskit to analyze genealogical
 information \citep{ralph2020efficiently}. To measure the reduction in
 introgressed ancestry around introgressed variants (as shown in
-Figure~\ref{fig:linkage}), we find each locus with a fixed difference between
-the parental populations at the time of admixture by preserving the generation
-immediately preceding admixture. For such a locus, for each sample we determine
-which (preserved) parental population its ancestry traces to at varying
-distances from the selected locus. Ancestry proportions are then averaged over
-each fixed difference. To measure proportions of introgressed ancestry or
-probabilities of observed ancestry deserts (as in Figure~\ref{fig:deserts}), we
-take the same approach to average ancestral source population proportions in 50
-kb windows. Ancestry deserts are defined as any such window with no ancestry
-inherited from the source population of introgression events.
+Figure~\ref{fig:linkage}), we identified each locus with a fixed difference
+between the parental populations at the time of admixture by preserving the
+generation immediately preceding admixture. For such a locus, for each sample
+we determined which (preserved) parental population its ancestry traced to at
+varying distances from the selected locus. Ancestry proportions were then
+averaged over each fixed difference. To measure proportions of introgressed
+ancestry or probabilities of observed ancestry deserts (as in
+Figure~\ref{fig:deserts}), we again used \tskit to average ancestry proportions
+of the source population in 50 kb windows. Ancestry deserts were defined as
+any such window with no ancestry inherited from the source population.
 
@@
 
-All code to run analyses, create figures, and compile this manuscript are available
-at \url{https://github.com/apragsdale/neanderthal_stabilizing_selection}.
+All code to run analyses, create figures, and compile this manuscript is
+available at
+\url{https://github.com/apragsdale/neanderthal_stabilizing_selection}.
 
@@
 
-We expect genetic variance to increase after introgression. The amount that
-genetic variance increases depends on the allelic differences accumulated
-between populations and the effects of those alleles. Assuming no linkage,
-dominance or epistasis, \(V_G=\sum_l 2p_l(1-p_l)a_l^2\). After admixture, with
-proportion $f$ contributed by the source population (labeled 0) into the focal
-population (labeled 1), \(p_l = fp_{l,0} + (1-f)p_{l,1}\). Plugging into the
-expression for $V_G$ and after some simple algebra
-(Appendix~\ref{sec:VG-admixture}), we can express the expected genetic variance
-directly after admixture as \[V_G = fV_{G,0} + (1-f) V_{G,1} + 2f(1-f)\sum_l
-F_{2,l} a_l^2,\] where $F_2 = (p_0 - p_1)^2$ is the squared difference in
-allele frequencies at a locus \citep{peter2016admixture}. This result is known
+We typically expect genetic variance to increase after introgression. The
+amount that genetic variance increases depends on the allelic differences
+accumulated between populations and the effects of those alleles. Assuming
+linkage equilibrium, and ignoring dominance and epistasis, \(V_G=\sum_l
+2p_l(1-p_l)a_l^2\). After admixture, with proportion $f$ contributed by the
+population labeled 0 and $1-f$ by population 1, \(p_l = fp_{l,0} +
+(1-f)p_{l,1}\). Plugging into the expression for $V_G$ and after a bit of
+algebra (Appendix~\ref{sec:VG-admixture}), we can write the expected genetic
+variance directly after admixture as
+\[V_G = fV_{G,0} + (1-f) V_{G,1} + 2f(1-f)\sum_l F_{2,l} a_l^2,\]
+where $F_2 = (p_0 - p_1)^2$ is the squared difference in allele frequencies at
+a locus \citep{peter2016admixture}. This result is known
 \citep[e.g.,][]{tufto2000quantitative}, showing that additive variance is equal
@@
 that depends on the divergence at trait-affecting loci between the populations
-weighted by the quadratic factor $2f(1-f)$.
+weighted by the quadratic factor $2f(1-f)$. Notably, $V_G$ is expected to
+increase only in the second generation after admixture, so that selection
+against introgressed alleles is not immediate \citep{veller2024stabilizing}.
+This effect is not captured by this expression.
 
@@
 vanishingly small effect, dynamics at a given locus will be approximately
-neutral, so that $F_2$ depends only on the demography. In this case,
+neutral, so that $F_2$ depends only on demographic history. In this case,
 \begin{align} \label{eq:VG-admixture}
@@
 \end{align}
-where $L$ is the number of trait affecting loci.
-
-\subsubsection*{Predicted $V_G$ from the SFS}
-
-For $a \not\approx 0$, the assumption of neutral evolution will lead to
-overestimates of $F_2$ compared to underdominant selection. We can compute
-expected $V_G$ both before and after admixture using the diffusion
-approximation for the joint SFS with underdominance. Comparing to simulations
-with free recombination between loci (Methods), we find that this provides an
-excellent fit to average observed $V_G$ over time
-(Figure~\ref{fig:toy-admixture}C,D). In general, admixture causes a sudden
-increase in $V_G$ followed by a fairly rapid return to pre-admixture levels.
-
-Modeling the dynamics of genetic variance using the SFS lets us examine
-contributions to $V_G$ from different classes of mutations, such as those at
-different frequencies or arising at different times, and how those
-contributions change over time. Of interest is the contribution to $V_G$ from
-alleles that were already segregating in the recipient population, those that
-were introduced through introgression, and new mutations since the time of
-admixture (Figure~\ref{fig:toy-admixture}E,F). In many scenarios, the initial
-divergence of the two populations will have occurred long enough before the
-admixture event, so that the mutations underlying $V_G$ are largely unique in
-each population. After mixing, previously segregating and introgressed alleles
-go to fixation or loss, and the variance of the trait is increasingly due to
-new mutations.
-
-Introgressed variation can initially make up a considerable portion of $V_G$,
-with those alleles being either newly introduced derived alleles or
-reintroduced ancestral alleles. The relative sizes of the two populations
-impacts the number of each, as derived alleles will accumulate more readily in
-a population with smaller effective size. Nonetheless, the increase in $V_G$
-can be fairly similar in either direction of introgrssion, as the term
-\(2f(1-f)\sum_l F_{2,l} a_l^2\) contributes in either case and can be much
-larger than $f V_G$ from the source population.
-
-\subsubsection*{Complex demography and partitioning heritability by origin of alleles}
-
-\begin{figure}[t!]
+where $L$ is the number of trait affecting loci and assuming $\E[a]=0$.
+
+\begin{figure}[ht!]
     \centering
@@
 
-We used a historical model more closely resembling inferred human-Neanderthal
-history (Figure~\ref{fig:human-neand-h2}A) to explore the effects of population
-size changes and reciprocal admixture on the additive genetic architecture of
-traits under stabilizing selection. As expected
-(Figure~\ref{fig:toy-admixture}), population contractions decrease $V_G$ as the
-increased rate of drift reduces allelic diversity at trait-affecting loci, and
-introgression increases $V_G$ in both directions of gene flow.
+\subsubsection*{Predicted \bm{$V_G$} from the SFS}
+
+Expected allele frequency differences ($F_2$) for selected alleles differ from
+neutrality. For negative and underdominant selection, $F_2$ is reduced relative
+to neutral expectations (Figure~\ref{fig:underdominance-F2}). Because analytic
+solutions are unavailable for arbitrary evolutionary scenarios involving
+multiple populations, we numerically solve for the expected joint distribution
+of allele frequencies (the SFS) before and after admixture. This provides a
+numerical solution for expected $V_G$, which can be tracked over time
+(Methods). Comparing to simulations with free recombination between loci, we
+observe close agreement with average $V_G$ at all times
+(Figure~\ref{fig:toy-admixture}C,D). In the scenarios tested here, admixture
+causes a sudden increase in $V_G$ followed by a fairly rapid return to
+pre-admixture levels, which is recovered by our numerical approach.
+
+Modeling the dynamics of genetic variance using the SFS lets us examine
+contributions to $V_G$ from different classes of mutations, such as those at
+different frequencies or arising at different times, and how those
+contributions change over time. In particular, we may quantify the contribution
+to $V_G$ from alleles that were already segregating in the recipient
+population, those that were introduced through introgression, and new mutations
+since the time of admixture (Figure~\ref{fig:toy-admixture}E,F). In many
+scenarios of interest, in which populations are considerably diverged at the
+time of admixture, genetic architectures will be largely unique in each
+population. After mixing, previously segregating and introgressed alleles each
+contribute to $V_G$ before going to fixation or loss, and the variance of the
+trait is increasingly due to new mutations.
+
+Introgressed variation can initially make up a considerable portion of $V_G$,
+with those alleles being either newly introduced derived alleles or
+reintroduced ancestral alleles. The relative sizes of the two populations
+impact the numbers of each, as derived alleles will accumulate more readily in
+a population with smaller effective size. Nonetheless, the overall increase in
+$V_G$ is similar in both directions of introgression, as the symmetric term
+\(2f(1-f)\sum_l F_{2,l} a_l^2\) contributes in either case and can be much
+larger than $f V_G$ from the source population
+(Equation~\ref{eq:VG-admixture}).
+
+\subsubsection*{Complex demography and partitioning heritability by origin of
+alleles}
+
+We used a historical model resembling inferred human-Neanderthal history
+(Figure~\ref{fig:human-neand-h2}A) to explore the effects of population size
+changes and reciprocal admixture on the additive genetic architecture of traits
+under stabilizing selection. As expected (Figure~\ref{fig:toy-admixture}),
+population contractions decrease $V_G$ as drift removes allelic diversity at
+trait-affecting loci, while introgression increases $V_G$.
 
@@
 segregating, fixed or lost in either parental populations or if they arose as
-new mutations since the time of admixture. This allows us to partition $V_G$ by
-contributions from these different sets of mutations
-(Figure~\ref{fig:human-neand-VG}). At low introgression proportions, as modeled
-here, the majority of $V_G$ is still contributed by previously segregating,
+new mutations since the time of admixture. Partitioning $V_G$ by contributions
+from these different sets of mutations (Figure~\ref{fig:human-neand-VG}), we
+find it is still primarily contributed by previously segregating,
 non-introgressed mutations. $V_G$ due to existing mutations decays
-monotonically over time and is fairly rapidly replace by new mutations.
+monotonically over time and is rapidly replaced by new mutations.
 
@@
 (i.e., $h^2$ per SNP) can similarly by tracked over time. For the
-human-Neanderthal-like demographic model and genetic architectures considered
-here, the contribution per SNP of introgressed variants is initally lower than
+human-Neanderthal demographic model and genetic architectures considered
+here, the contribution per SNP of introgressed variants is initially lower than
 that of non-introgressed variants. These contributions change over time,
@@
 
-In the preceding sections, we find that approximating the dynamics of
+In the preceding sections, we found that approximating the dynamics of
 trait-affecting alleles using an underdominance model
 \citep{robertson1956effect} provides an excellent approximation of $V_G$ in
-complex demographic scenarios. However, this relies on an absence of linkage
-between trait-affecting alleles. The inclusion of linkage can lead to
-noticeable effects on expected $V_G$, so that Equation~\ref{eq:SHC} differs
-from observed $V_G$ at steady state
-\citep{burger1989much, burger1994distribution, walsh2018evolution}.
+complex demographic scenarios. However, this relies on linkage equilibrium
+between trait-affecting alleles. The inclusion of linkage can cause noticeable
+distortions of expected $V_G$, so that Equation~\ref{eq:SHC} differs from
+observed $V_G$ at steady state \citep{burger1989much, burger1994distribution,
+walsh2018evolution}.
 
@@
 individual-based simulations \citep{thornton2019polygenic}. By varying the
-mutation rate and the variance of effect sizes of new mutations, we include
+mutation rate and the variance of effect sizes of new mutations, we included
 scenarios ranging from low to high polygenicity and from weak to strong
-selection on individual alleles. In this and the following sections, we will
+selection on individual alleles. In this and the following sections, we
 highlight two effects of linkage. First, we observe deviations of $V_G$ from
-expectations without linkage, which can be large for highly polygenic traits.
-Second, selection on introgressed trait-affecting alleles results in a
+expectations under free recombination, which can be large for highly polygenic
+traits. Second, selection on introgressed trait-affecting alleles results in a
 reduction of introgressed ancestry in surrounding regions.
 
-With linkage between two or more alleles affecting a trait under stabilizing
-selection, linkage disequilibrium (LD) can develop between alleles. Classicaly,
-stabilizing selection will cause mutations of opposite-signed effects to be in
-coupling LD and those of same-signed effects in repulsion LD
-\citep{bulmer1971effect}. This has the effect of reducing $V_G$. For lower
+With linkage between two or more loci contributing to a trait under stabilizing
+selection, linkage disequilibrium (LD) can develop between alleles. Notably,
+stabilizing selection will lead to coupling LD between mutations of
+opposite-signed effects and repulsion LD between those of same-signed effects
+\citep{bulmer1971effect}. This is expected to decrease $V_G$. For lower
 mutational variances ($2N_e V_M\approx1$, Figure~\ref{fig:supp-low-VM}), we
-observe this phenomenon. With low mutational input, and thus low polygenicity,
-$V_G$ in individual simulations closely matches expectations without linkage.
-As we increase the mutation rate, $V_G$ is reduced relative to those
-expectations.  However, when the mutational variance is much larger
-($2N_eV_m\approx 50$), we see the reverse trend
+observe such a reduction in $V_G$. With low mutational input, and thus low
+polygenicity, $V_G$ in individual-based simulations closely matches
+expectations without linkage. With increasing mutation rate, $V_G$ is reduced
+relative to those expectations. However, when the mutational variance is much
+larger ($2N_eV_M\approx 50$), we see the reverse trend
 (Figure~\ref{fig:supp-high-VM}). At low mutation rate, there is a close match
@@
 effect sizes plays an important role, as seen by the opposite trends in $V_G$
-relative to unlinked predictions (Figure~X vs. Figure~Y), which are most
-apparent with high mutation rates, and thus high polygenicity. In such cases,
-there appear to be complex dynamics involving the Bulmer Effect
-\citep{bulmer1971effect} and interference between selected alleles
+relative to unlinked predictions (Figure~\ref{fig:supp-low-VM} vs.
+Figure~\ref{fig:supp-high-VM}), which are most apparent with high mutation
+rates, and thus high polygenicity. In such cases, there appear to be complex
+dynamics involving the reduction in $V_G$ due to the Bulmer Effect
+\citep{bulmer1971effect} and an increase in $V_G$ due to selective interference
 \citep{hill1966effect}.
 
-\begin{figure}[t!]
+\begin{figure}[ht!]
     \centering
@@
         the effect size of the selected allele and the probability of
-        recombination between them. LD (as measured by $D=Cov(p,q)$) decays
+        recombination between them. LD (as measured by $D=\Cov(p,q)$) decays
         to zero over time.
@@
 
-Because introgressed trait-affecting alleles are selection against as the minor
+Because introgressed trait-affecting alleles are selected against as the minor
 allele, introgressed ancestry segments in the regions surrounding the selected
-alleles will also be removed due to linkage. The expected reduction in such
-introgressed ancestry will depend on the effect size of the trait-affecting
-allele and the probability of recombination surrounding that locus. We first
-consider a deterministic model of the dynamics of introgressed allele and
-linked ancestry frequencies with variable recombination between them. This
+alleles will also be removed due to linkage. The expected reduction in
+introgressed ancestry will depend on the effect size of the linked
+trait-affecting allele and the local probability of recombination. We first
+consider a deterministic model for the frequency dynamics of introgressed
+alleles (one selected, one neutral) with variable rate of recombination. This
 simple model ignores the effects of drift and of interference between multiple
@@
 
-We model a trait-affecting site with a fixed difference between the two
+We model a trait-affecting locus with a fixed difference between the two
 parental populations, in which the derived allele may be fixed in either
-population. At the functional site, with admixture proportion $f$ from the
+population. At the functional locus, with admixture proportion $f$ from the
 minor parental source, the initial frequency of the derived allele is either
@@
 \end{align}
-with
-\(s=a^2/2V_S\). As discussed above, if \(p_0<1/2\), \(p_t\rightarrow0\), and if
+with \(s=a^2/2V_S\) for stabilizing selection.
+As discussed above, if \(p_0<1/2\), \(p_t\rightarrow0\), and if
 \(p_0>1/2\), \(p_t\rightarrow1\) as \(t\rightarrow\infty\).
@@
 ancestry is \(q_0=f\), which changes over time due to linked selection on the
-trait-affecting allele. Letting \(D=Cov(p,q)\) be the standard covariance
-measure of LD between the two loci, $q$ is expected to change as
+trait-affecting allele. Letting \(D=\Cov(p,q)\) be the standard covariance
+measure of LD between the alleles at the two loci, $q$ is expected to change as
 \begin{align}\label{eq:system-q}
@@
 positive if \(p_0=f\) and negative if \(p_0=1-f\). LD between the loci changes
-deterministically over time, due to both selection and recombination, so that
+deterministically over time due to both selection and recombination, so that
 \begin{align}\label{eq:system-D}
@@
 
-Using this model, we predict the changes in introgressed allele frequences and
-LD after admixture for given effect size $a$ and recombination rate $r$
+Using this model, we predicted the changes in introgressed allele frequencies
+and LD after admixture for given effect size $a$ and recombination rate $r$
 (Figure~\ref{fig:linkage}A,B). As expected, smaller effect sizes result in a
@@
 from the selected allele dynamics. Thus, the expected reduction in introgressed
-ancestry is largest for larger effect sizes (Figure~\ref{fig:linkage}C),
-and LD between the selected and linked neutral alleles is largest for neutral
-sites closest to the selected allele (Figure~\ref{fig:supp-LD}).
+ancestry is largest for larger effect sizes (Figure~\ref{fig:linkage}C), and LD
+between the selected and linked neutral alleles is largest for neutral sites
+closest to the selected allele (Figure~\ref{fig:supp-LD}).
 
@@
 Morgan, with all mutations having effect sizes \(\pm a\) (\(a=0.02\) or
-\(0.05\)), and we varied the total per-chromosome mutation rate (\(\mu=0.001\),
-\(0.0025\) and \(0.01\)). For each fixed-difference mutation in the parental
-populations, we determined the average introgressed ancestry surrounding such
-loci (Methods).
+\(0.05\), in separate simulations), and we varied the total per-chromosome
+mutation rate (\(\mu=0.001\), \(0.0025\) and \(0.01\)). For each
+fixed-difference mutation in the parental populations, we determined the
+average introgressed ancestry surrounding such loci $4{,}000$ generations after
+admixture (Methods).
 
@@
 provides a very good approximation when mutation rates are small. As the
-mutation rate increases in these simulations, polygenicity also increases so
-that trait-affecting alleles are more densely distributed along the chromosome.
-Deviations from the deterministic model are due to multiple selected alleles
-affecting local introgressed ancestry. For the highest mutation rate shown
-here, there are expected to be multiple other trait-affecting alleles within a
-1 cM window around any focal SNP, which interact with each other and also
-affect local ancestry proportions.
+mutation rate increases in these simulations, trait-affecting alleles are more
+densely distributed along the chromosome. Deviations from the deterministic
+model are due to multiple selected alleles affecting local introgressed
+ancestry. For the highest mutation rate shown here, there can be many other
+trait-affecting alleles within a 1 cM window around any focal SNP, distorting
+dynamics away from predictions under the two-locus model.
 
@@
 
-\begin{figure}[t!]
+\begin{figure}[ht!]
     \centering
@@
         \emph{shared} deserts when compared across samples from humans and
-        Neanderthals.
+        Neanderthals. In these simulations, $\sigma_M=0.02$ and $\mu=0.01$.
+        Simulations with different mutational variances are shown in
+        Figures~\ref{fig:deserts-high-VM}--\ref{fig:deserts-low-VM}.
+        The pattern of shared deserts is not seen in simulations with
+        directional selection (Figures~\ref{fig:deserts-directional-A}
+        and \ref{fig:deserts-directional-B}).
     }
@@
 To demonstrate this effect, we performed chromosome-scale simulations under a
-model of human-Neanderthal reciprocal introgression
+model of reciprocal introgression between humans and Neanderthals
 \citep[Figure~\ref{fig:human-neand-h2}A,][]{harris2023diverse}. Mutations
-affecting a trait under stabilizing selection in each population could arise in
-100 kb ``functional'' regions, spaced 1 Mb apart (Methods). Sampling
+affecting a trait under stabilizing selection in each population occurred in
+100 kb ``functional'' regions, spaced 2 Mb apart (Methods). Sampling
 individuals from both the human and Neanderthal lineages, we observed the
-average introgressed ancestry proportions were lowest within functional regions
-and dipped in the surrounding regions (Figure~\ref{fig:deserts}A). This
-corresponded to an increased proportion of ancestry deserts (defined as
-    contiguous 50kb regions with no observed introgressed ancestry in the
-sample) within and immediately surrounding the functional regions
-(Figure~\ref{fig:deserts}B). Functional regions also displayed an enrichment of
-\emph{shared} ancestry deserts, with the probability of observing ancestry
+average introgressed ancestry proportions were lowest within and surrounding
+functional regions (Figure~\ref{fig:deserts}A). This corresponded to an
+increased proportion of ancestry deserts within the functional regions
+(Figure~\ref{fig:deserts}B). Functional regions also displayed an enrichment
+of \emph{shared} ancestry deserts, with the probability of observing ancestry
 deserts (either within a population or shared) decaying to background levels as
@@
 selection around a shared optimum as a null model for the dynamics of alleles
-affecting polygenic traits \citep{yair2022population}. Stabilizing selection
-has a distinct effect on trait-affecting alleles, characterized by symmetric
-underdominance \citep{robertson1956effect, keightley1988quantitative}, that is,
-selection against the minor allele. Some theoretical and simulation studies
-consider the effect of population differentiation and migration on the genetic
-architecture of a trait under stabilizing selection
-\citep[e.g.,][]{tufto2000quantitative, yeaman2011genetic, yair2022population},
-but most previous work has focused on single population scenarios, often
-assuming steady state dynamics. Admixture and introgression events are common
-in natural populations, so understanding their effects on the architecture of
-complex traits is needed.
-
-Here, we have shown that multi-population, non-equilibrium approaches for the
-site frequency spectrum with underdominance can be used to model the additive
+affecting polygenic traits, including in multi-population settings
+\citep{yair2022population}. Stabilizing selection on a trait results in
+symmetric underdominant selection at trait-contributing loci
+\citep{robertson1956effect, keightley1988quantitative}, that is, selection
+against the minor allele. Some theoretical and simulation studies consider the
+effect of population differentiation and migration on the genetic architecture
+of a trait under stabilizing selection \citep[e.g.,][]{tufto2000quantitative,
+yeaman2011genetic, yair2022population}, but most previous work has focused on
+single population scenarios, often assuming steady state dynamics. Episodes of
+admixture and introgression commonly occur in many species' evolutionary
+histories, so understanding their effects on the architecture of complex traits
+is needed.
+
+Here, we have shown that a multi-population, non-equilibrium approach for the
+site-frequency spectrum with underdominance can be used to model the additive
 genetic variance of a trait under stabilizing selection. While this approach
-ignores biologically relevant aspects such as linkage between sites and
-pleiotropy, it can still provide important insights into the dynamics of trait
-architectures. By modeling allelic contributions to the trait as additive, it
-allows for the decomposition of contributions from alleles of different
-origins, either by source population or mutation time. It may therefore be
-useful for understanding the expected contributions of introgressed variants to
-complex traits, for example from Neanderthal-introgressed mutations in humans
-\citep{wei2023lingering}. In the limited scenarios explored here, we see that
-the expected contributions of introgressed variants to heritability are
-complicated even in the purely additive case, depending on the distribution of
-effect sizes, demographic history, and the time since admixture
-(Figures~\ref{fig:human-neand-h2}, \ref{fig:human-neand-VG}).
+ignores some biological relevancies, such as linkage between sites,
+non-additivity and pleiotropy, it still provides important insights into the
+dynamics of trait architectures. For example, we can readily decompose
+contributions from alleles of different origins, either by source population or
+mutation time. It may therefore be useful for understanding the contributions
+of introgressed variants to complex traits, such as hominin-introgressed
+mutations in humans \citep{reilly2022contribution, wei2023lingering}. In the
+limited scenarios explored here, we find that the expected contributions of
+introgressed variants to heritability are complicated even in the purely
+additive case, depending on the distribution of effect sizes, demographic
+history and the time since admixture (Figures~\ref{fig:human-neand-h2} and
+\ref{fig:human-neand-VG}).
 
@@
 divergence at trait-affecting loci, measured by $F_2$
-(Equation~\ref{eq:VG-admixture}). $V_G$ is then expected to decay fairly
-quickly to background levels. During this process, introgressed and
-pre-existing trait-affecting alleles are replaced by new mutations, turning
-over the genetic architecture of the trait (Figure~\ref{fig:toy-admixture}).
-Selection occurs against the minor alleles at trait-affecting loci, so that
-introgressed alleles from the minor parental ancestry, whether derived or
-reintroduced ancestral alleles, are selected against
-(Figure~\ref{fig:linkage}).
-
-Selection against introgressed alleles will also remove linked introgressed
+(Equation~\ref{eq:VG-admixture}) \citep[see also,][]{veller2024stabilizing}.
+$V_G$ is then expected to decay fairly quickly to background levels. During
+this process, introgressed and pre-existing trait-affecting alleles are
+replaced by new mutations, turning over the genetic architecture of the trait
+(Figure~\ref{fig:toy-admixture}).  Selection occurs against the minor alleles
+at trait-affecting loci, so that introgressed alleles from the minor parental
+ancestry, whether derived or reintroduced ancestral alleles, are selected
+against (Figure~\ref{fig:linkage}).
+
+Selection against introgressed alleles also removes linked introgressed
 ancestry in the surrounding regions. Thus, introgressed ancestry deserts are
@@
 form in the same regions under reciprocal introgression. When comparing
-distributions of introgressed haplotypes in two diverged populations with more
-recent gene flow, we should expect to see an overlap in deserts of introgressed
-ancestry at loci affecting traits under stabilizing selection.
+distributions of introgressed haplotypes in two diverged populations with
+recent bi-directional gene flow, we should expect to see an overlap of deserts
+in genomic regions contributing to traits under stabilizing selection.
 
 The expected pattern of shared ancestry deserts under stabilizing selection
-differs from models of both deleterious load and incompatibilities. Load-based
-models predict that haplotypes that have accumulated more deterious mutations,
-e.g., from a population with small long-term effective population size, will be
-selected against under either direction of gene flow. Introgressed ancestry at
-a given selected locus will decrease in frequency in one introgression scenario
-and increase the other. This may explain the replacement of MT and Y chromosome
-DNA in Neanderthals by human haplotypes after early human-to-Neanderthal
-introgression \citep{posth2017deeply, petr2020evolutionary}, but does not broadly
-match observations across the autosomal genome \citep{harris2023diverse}.
+differs from models of both deleterious load and incompatibilities, providing
+testable hypotheses. Load-based models predict that haplotypes that have
+accumulated more deleterious mutations, e.g., from a population with small
+long-term effective population size, will be selected against under either
+direction of gene flow. Introgressed ancestry at a given selected locus will
+decrease in frequency in one introgression scenario and increase in the other.
+This may explain the replacement of MT and Y chromosome DNA in Neanderthals by
+human haplotypes after early human-to-Neanderthal introgression and the absence
+of such Neanderthal haplotypes in modern humans \citep{posth2017deeply,
+petr2020evolutionary}, but does not broadly match observations across the
+autosomal genome \citep{harris2023diverse}.
 
@@
 hypothesized that such BDMIs should most often form between distant or unlinked
-loci, instead of within single or tightly linked loci. Because theory predicts
-that hybrid incompatibilities are resolved via selection against the minor
-parental ancestry \citep{moran2021genomic}, ancestry deserts should form in
-different genomic regions as the different incompatibility alleles are selected
-against, depending on the direction of introgression. While such a model
-predicts selection against introgressed ancestry, similar to that of
-stabilizing selection, an important distinction is that deserts due to
+loci, instead of within a single locus or tightly linked loci. Because theory
+and experiments show that hybrid incompatibilities are resolved via selection
+against the minor parental ancestry \citep{matute2020rapid, moran2021genomic},
+ancestry deserts should form in different genomic regions, since different
+incompatibility alleles are selected against depending on the parental ancestry
+proportions. While both BDMI and stabilizing selection models predict selection
+against introgressed alleles, an important distinction is that deserts due to
 selection against incompatibility loci are not expected to overlap under
@@
 distribution of BDMIs, studies point to interacting BDMI alleles being unlinked
-\citep[e.g.,][]{presgraves2003fine} and an asymmetry in the alleles under
-selection in different introgression scenarios \citep{maheshwari2011genetics,
-moran2021genomic}.
+\citep{presgraves2003fine, li2022imbalanced} and an asymmetry in the alleles
+under selection in different introgression scenarios
+\citep{maheshwari2011genetics, moran2021genomic}.
 
 Finally, turning to the distribution of introgressed ancestry deserts in humans
-and Neanderthals, \citet{harris2023diverse} find that regions of depleted
+and Neanderthals, \citet{harris2023diverse} found that regions of depleted
 Neanderthal ancestry in humans overlap more than would be expected by chance
@@
 Human-introgressed ancestry in the Neanderthal genome is also depleted in
-functional regions, as observed in humans \citep{sankararaman2014genomic}.
+functional regions \citep{harris2023diverse}, as is well-documented in humans
+\citep{sankararaman2014genomic, sankararaman2016combined}. 
 \citet{harris2023diverse} propose that epistatic interactions between
 introgressed alleles and the recipient backgrounds could drive these patterns,
-which they interpret as evidence for the inital process of speciation between
+which they interpret as evidence for the initial process of speciation between
 humans and Neanderthals. However, the observation of shared ancestry deserts
@@
 overlapping ancestry deserts are expected even when a trait is under
-stabilizing selection for the same phenotypic optimal value. The selective
-causes of ancestry deserts in both humans and Neanderthals remain unknown.
-Stabilizing selection will prove a useful null model for testing scenarios of
-epistasis, incompatibilities and adaptive introgression.
+stabilizing selection for the same phenotypic optimal value. While the
+underlying causes of selection on introgressed alleles in humans and
+Neanderthals remain largely unknown, stabilizing selection provides a
+well-grounded explanation for observed patterns that should be considered when
+testing for epistasis, incompatibilities and adaptive introgression.
 
@@
 \appendix
-
-\section{Stabilizing selection results in underdominant selection on
-trait-affecting alleles} \label{sec:underdominance}
+\section*{Appendix}
+
+\section{Stabilizing selection results in selection on trait-affecting alleles
+equivalent to symmetric underdominance} \label{sec:underdominance}
 
@@
 selection, and we assume the mean phenotype of the population is close to the
-optimal value (which will be the case if the optimum has not shifted recently).
-Because many segregating loci are assumed to contribute to the phenotypic
-variance of the trait, the phenotypic distribution is well-approximated as
-normally distributed around the optimum: \(f(G)=\frac{1}{\sqrt{2\pi V_G}}
-e^{-G^2/2V_G}\). We assume a Gaussian fitness function, so that the fitness
-of an individual with genotypic value $G$ is given by \(w(G)=e^{-G^2/2V_S}\).
+optimal value (0), which will be the case if the optimum has not shifted
+recently. Because many segregating loci are assumed to contribute to the
+phenotypic variance of the trait, the phenotypic distribution is
+well-approximated as normally distributed around the optimum:
+\(f(G)=\frac{1}{\sqrt{2\pi V_G}} e^{-G^2/2V_G}\). We assume a Gaussian fitness
+function, so that the fitness of an individual with genotypic value $G$ is
+given by \(w(G)=e^{-G^2/2V_S}\).
 
 For a derived allele (labeled 1) with frequency $p$, the expected change in
-frequency over one generation due to selection in general is
+frequency over one generation due to selection (in general) is
 \[\E[\Delta p]=p(1-p)\frac{w_{1\cdot}-w_{0\cdot}}{\bar{w}},\]
@@
 and
-\[w_{0\cdot}=p w_{01} + (1-p)w_{00},\]
-respectively. Here, $w_{11}$, $w_{01}$, and $w_{00}$ are the relative fitnesses
-of individuals who are homozygous for the derived allele, heterozygous, or
+\[w_{0\cdot}=p w_{01} + (1-p)w_{00}.\]
+Here, $w_{11}$, $w_{01}$, and $w_{00}$ are the relative fitnesses of
+individuals who are homozygous for the derived allele, heterozygous, or
 homozygous for the ancestral allele, respectively.
@@
 We partition an individual's genetic values into contributions from their
-genetic background and the focal locus: $G=\tilde{G}+xa$, where $x\in{0,1,2}$.
-We then integrate over genetic backgrounds to find the expected fitnesses of
-each genotype. As described in \citet[][SI section 2.1]{simons2018population},
-because each individual locus contributes a small amount to $V_G$,
-$V_{\tilde{G}}\approx V_G$ and the fitness function is well approximated as
-\(f(\tilde{G})=\frac{1}{\sqrt{2\pi V_G}}e^{-\tilde{G}^2/2V_G}\). Then,
-mean fitness
+genetic background ($\tilde{G}$) and the focal locus: $G=\tilde{G}+xa$, where
+$x\in{0,1,2}$.  We then integrate over genetic backgrounds to find the expected
+fitnesses of each genotype. As described in SI section 2.1 of
+\citet{simons2018population}, because each individual locus contributes a small
+amount to $V_G$, $V_{\tilde{G}}\approx V_G$ and the fitness function is well
+approximated as \(f(\tilde{G})=\frac{1}{\sqrt{2\pi
+V_G}}e^{-\tilde{G}^2/2V_G}\). Then, mean fitness
 \[
@@
 \[ 
-    w_{11} \approx f(\tilde{G}) w(\tilde{G} + 2a) d\tilde{G}
+    w_{11} \approx \int_{-\infty}^\infty f(\tilde{G}) w(\tilde{G} + 2a) d\tilde{G}
     = \bar{w}\exp{\left(\frac{-4a^2(1-p)^2}{2(V_S+V_G)}\right)},
@@
 \[
-    w_{01} \approx f(\tilde{G}) w(\tilde{G} + a) d\tilde{G}
+    w_{01} \approx \int_{-\infty}^\infty f(\tilde{G}) w(\tilde{G} + a) d\tilde{G}
     = \bar{w}\exp{\left(\frac{-a^2(1-2p)^2}{2(V_S+V_G)}\right)},
@@
 \[
-    w_{00} \approx f(\tilde{G}) w(\tilde{G}) d\tilde{G}
+    w_{00} \approx \int_{-\infty}^\infty f(\tilde{G}) w(\tilde{G}) d\tilde{G}
     = \bar{w}\exp{\left(\frac{-4a^2p^2}{2(V_S+V_G)}\right)}.
@@
 
+This same result can be found by considering a haploid model
+\citep{negm2024effect}. While the selection coefficient takes an equivalent
+form to underdominance, we note that the mechanism differs: selection acts
+against the minor allele, rather than for the heterozygous genotype.
+
 \section{Moment equations for over- and underdominance}
@@
 For symmetric underdominance, the relative fitnesses of genotypes \(aa:Aa:AA\)
-are \(1:1-s:1\). We consider any value of $s$ (positive or negative,
+are \(1:1-t:1\). We consider any value of $s$ (positive or negative,
 corresponding to either under- or overdominance); with stabilizing selection,
-\(s=\frac{a^2}{2V_S}\). We extended \moments \citep{jouganous2017inferring} to
-compute the sample site frequency spectrum \(\Phi_{\mathbf{n}}\) for one or
-more (up to five) populations with sample sizes \(\mathbf{n}\). This provides a
-good approximation for the distribution of trait-affecting allele frequencies
-across multiple populations if the trait optimum is shared across populations
-and each population's mean phenotype remains close to that optimum. This is
-expected to be the case if there are no optimum shifts in any lineage.
-Hardy-Weinberg equilibrium is assumed at all loci. Accounting for optimum
-shifts in one or more lineages would require a combination of direct and
-underdominant selection \citep[e.g.,][]{hayward2022polygenic}, which we leave
-for future work.
+\(t=\frac{a^2}{2V_S}\), so that heterozygous individuals always have reduced
+fitness compared to either homozygote.
+
+We extended \moments \citep{jouganous2017inferring} to compute the sample
+site-frequency spectrum \(\Phi_{\mathbf{n}}\) for one or more (up to five)
+populations with sample sizes \(\mathbf{n}\). This provides a good
+approximation for the distribution of trait-affecting allele frequencies across
+multiple populations if the trait optimum is shared across populations and each
+population's mean phenotype remains close to that optimum. This is expected to
+be the case if there are no optimum shifts in any lineage.  Hardy-Weinberg
+equilibrium is assumed at all loci. Accounting for optimum shifts in one or
+more lineages would require a combination of direct and underdominant selection
+\citep[e.g.,][]{hayward2022polygenic}, which we leave for future work.
 
@@
 Here, we describe how underdominant selection changes \(\Phi\) over a single
-generation. With symmetric underdominance, selection acts heterozygotes. This
+generation. With symmetric underdominance, selection acts on heterozygotes. This
 can be formulated as some proportion of heterozygotes failing to reproduce in a
@@
 observed derived alleles in a haploid sample of size of $n$). $i$ can increase
-or decrease due to selection ``events''. Here, we assume $s$ is small enough so
+or decrease due to selection ``events''. Here, we assume $t$ is small enough so
 that at most a single selection event occurs among the $n$ lineages in any
-given generation. This is a reasonable approximation as long as $s$ is not
-extremely large \citep{jouganous2017inferring} -- for selection coefficients
-induced by stabilizing selection, this is typically a safe assumption.
+given generation. This is a reasonable approximation as long as $t$ is not
+extremely large \citep{jouganous2017inferring, krukov2021taming} -- for
+selection coefficients induced by stabilizing selection, this is typically a
+safe assumption.
 
@@
 
-For case (a), \(i\rightarrow i-1\) (i.e., \(\Phi_n(i)\) is reduced) with
-probability \[s\frac{i(n-i+2)(n-i+1)}{(n+2)(n+1)}\Phi_{n+2}(i),\] and
-\(i+1\rightarrow i\) (\(\Phi_n(i)\) increases) with probability
-\[s\frac{(i+1)(n-i+1)(n-i)}{(n+2)(n+1)}\Phi_{n+2}(i+1).\] For case (b),
+The expected increase and decrease of $\Phi_n(i)$ due to selection is found by
+enumerating the sampling probabilities of each selective event. For case (a),
+\(i\rightarrow i-1\) (i.e., \(\Phi_n(i)\) is reduced) with probability
+\[t\frac{i(n-i+2)(n-i+1)}{(n+2)(n+1)}\Phi_{n+2}(i),\] and \(i+1\rightarrow i\)
+(\(\Phi_n(i)\) increases) with probability
+\[t\frac{(i+1)(n-i+1)(n-i)}{(n+2)(n+1)}\Phi_{n+2}(i+1).\] For case (b),
 \(i\rightarrow i+1\) with probability
-\[s\frac{(n-i)(i+2)(i+1)}{(n+2)(n+1)}\Phi_{n+2}(i+2),\] and \(i-1 \rightarrow
-i\) with probability \[s\frac{(n-i+1)(i+1)i}{(n+2)(n+1)}\Phi_{n+2}(i+1).\]
+\[t\frac{(n-i)(i+2)(i+1)}{(n+2)(n+1)}\Phi_{n+2}(i+2),\] and \(i-1 \rightarrow
+i\) with probability \[t\frac{(n-i+1)(i+1)i}{(n+2)(n+1)}\Phi_{n+2}(i+1).\]
 These can be combined (with negative rates for the reduction of \(\Phi_n(i)\))
 to describe the change of \(\Phi_n(i)\) for all \(0\leq i \leq n\).
-Figures~\ref{fig:underdominance-validation-small-s} and
-\ref{fig:underdominance-validation-large-s} show that this approach,
-implemented in \moments, is accurate compared to discrete Wright-Fisher
-simulations.
+Figures~\ref{fig:underdominance-validation-small-s}--\ref{fig:overdominance-validation}
+show that this approach, implemented in \moments, is accurate compared to
+discrete Wright-Fisher simulations.
 
@@
 
-Assuming no linkage, dominance or epistasis (so \(V_G=V_A\)),
+Assuming random mating and no linkage, dominance or epistasis (so \(V_G=V_A\)),
 \[V_G = \sum_l 2p_l(1-p_l)a_l^2 = \sum_l \pi_l a_l^2,\]
-where $\pi$ denotes expected pairwise diverisity.
+where $\pi$ denotes expected pairwise diversity.
 At a given locus (dropping the $l$), after admixture the allele frequency is
@@
     2p(1-p) & = 2(f p_0 + (1-f) p_1)(1 - f p_0 - (1-f) p_1) \\
-    & = f^2 2p_0(1-p0) + (1-f)^2 p_1(1-p1) + 2f(1-f) (p_0(1-p_1) + p_1(1-p_0)) \\
+    & = f^2 2p_0(1-p_0) + (1-f)^2 p_1(1-p_1) + 2f(1-f) (p_0(1-p_1) + p_1(1-p_0)) \\
     & = f^2 \pi_{0,0} + (1-f)^2 \pi_{1,1} + 2f(1-f)\pi_{0,1}.
 \end{align*}
-
-Plugging in to the definition for $V_G$, we get after admixture
+Here, $\pi_{i,j}$ is the expected pairwise diversity between two samples, one
+drawn from population $i$ and the other from population $j$. If $i=j$, this is
+pairwise diversity within a single population.
+
+Plugging into the definition for $V_G$, we get after admixture
 \[V_G = f^2 V_{G,0} + (1-f)^2 V_{G,1} + 2f(1-f)\sum_l \pi_{0,1,l}a_l^2.\]
 We can write $\pi_{0,1}$ at a given locus in terms of $\pi_{0,0}$, $\pi_{1,1}$,
-and $F_2(0,1)=(p_0-p_1)^2$ as \citep{peter2016admixture}
-\[\pi_{0,1} = F_2(0,1) + \frac{1}{2}\pi_{0,0} + \frac{1}{2}\pi_{1,1}.\]
-Then
+and $F_2(0,1)=(p_0-p_1)^2$, a measure of single-locus allele frequency
+differentiation \citep{peter2016admixture}. Then
+\[\pi_{0,1} = F_2(0,1) + \frac{1}{2}\pi_{0,0} + \frac{1}{2}\pi_{1,1},\]
+and
 \begin{align*}
@@
     + \frac{1}{2}V_{G,1} + \sum_l F_{2,l}(0,1)a_l^2\right] \\
-    & = \left(f^2 + f(1-f)\right) V_{G,0} + \left((1-f)^2 + f(1-f)\right)V_{G,1}
-    + 2f(1-f) \sum_l F_{2,i}(0,1)a_l^2 \\
+    %& = \left(f^2 + f(1-f)\right) V_{G,0} + \left((1-f)^2 + f(1-f)\right)V_{G,1}
+    %+ 2f(1-f) \sum_l F_{2,i}(0,1)a_l^2 \\
     & = f V_{G,0} + (1-f)V_{G,1} + 2f(1-f)\sum_i F_{2,l}(0,1)a_l^2.

### 2024-08-19 doc/supplement.tex
paper edits

@@
 
-\title{Supporting Information}
-\author{APR}
+\title{Supporting Information for\\
+``Archaic introgression and the distribution
+of shared variation under stabilizing selection''}
+\author[]{Aaron P. Ragsdale}
+\affil[]{Department of Integrative Biology, University of Wisconsin--Madison, WI, USA}
 \date{\today}
@@
 
-\begin{figure}[t!]
-    \centering
-    \includegraphics{../figures/human-neand-VG.pdf}
-    \caption{
-        Genetic variance of a trait under stabilizing selection in a model with
-        human (blue) and Neanderthal (orange) demography. We consider two
-        mutational variances ($\sigma_M=0.05$ and $0.01$). The demographic model
-        (Figure~\ref{fig:human-neand-h2}A) includes $5\%$ admixture from humans
-        to Neanderthals 250 ka, and $2\%$ admixture from Neanderthals to humans
-        50 ka.
-    }
-    \label{fig:human-neand-VG}
-\end{figure}
-
 \begin{figure}[ht!]
@@
     \label{fig:supp-high-VM}
+\end{figure}
+
+\begin{figure}[t!]
+    \centering
+    \includegraphics{../figures/human-neand-VG.pdf}
+    \caption{
+        Genetic variance of a trait under stabilizing selection in a model with
+        human (blue) and Neanderthal (orange) demography. We consider two
+        mutational variances ($\sigma_M=0.05$ and $0.01$). The demographic model
+        (Figure~\ref{fig:human-neand-h2}A) includes $5\%$ admixture from humans
+        to Neanderthals 250 ka, and $2\%$ admixture from Neanderthals to humans
+        50 ka.
+    }
+    \label{fig:human-neand-VG}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics{../figures/introgression_deserts.SD_0.05.pdf}
+    \caption{
+        Introgressed ancestry and probability of introgression deserts in and
+        around functional regions, from simulations with high mutational variance.
+        The demographic model is shown in Figure~\ref{fig:human-neand-h2}A.
+    }
+    \label{fig:deserts-high-VM}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics{../figures/introgression_deserts.SD_0.01.pdf}
+    \caption{
+        Introgressed ancestry and probability of introgression deserts in and
+        around functional regions, from simulations with moderate mutational variance.
+        The demographic model is shown in Figure~\ref{fig:human-neand-h2}A.
+    }
+    \label{fig:deserts-med-VM}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics{../figures/introgression_deserts.SD_0.005.pdf}
+    \caption{
+        Introgressed ancestry and probability of introgression deserts in and
+        around functional regions, from simulations with low mutational variance.
+        The demographic model is shown in Figure~\ref{fig:human-neand-h2}A.
+    }
+    \label{fig:deserts-low-VM}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics{../figures/introgression_deserts.directional.mean_-0.0002.shape_0.5.pdf}
+    \caption{
+        Introgressed ancestry and probability of introgression deserts in and
+        around functional regions, from simulations with directional selection.
+        Deleterious mutations were drawn from a gamma distribution with mean
+        $-0.0002$ and shape $0.05$.
+        The demographic model is shown in Figure~\ref{fig:human-neand-h2}A.
+    }
+    \label{fig:deserts-directional-A}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics{../figures/introgression_deserts.directional.mean_-0.001.shape_0.5.pdf}
+    \caption{
+        Introgressed ancestry and probability of introgression deserts in and
+        around functional regions, from simulations with directional selection.
+        Deleterious mutations were drawn from a gamma distribution with mean
+        $-0.001$ and shape $0.05$.
+        The demographic model is shown in Figure~\ref{fig:human-neand-h2}A.
+    }
+    \label{fig:deserts-directional-B}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics[width=0.6\textwidth]{../figures/underdominance_F2.pdf}
+    \caption{
+        Underdominance, like negative selection, reduces expected $F_2$ compared
+        to neutral divergence.
+    }
+    \label{fig:underdominance-F2}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.6\textwidth]{../figures/underdominance.Ne_10000.s_-0.001.pdf}
+    \includegraphics[width=0.6\textwidth]{../figures/underdominance.Ne_5000.s_-0.001.pdf}
     \caption{
         Comparison of predicted (\moments) and simulated SFS with underdominant
-        selection. In this comparison, $N_e=10^4$ and $s=-1\times10^{-3}$, so that
-        the population-size scale selection coefficient $\gamma=-20$.
+        selection. In this comparison, $N_e=5000$ and $s=-0.001$, so that
+        the population-size scale selection coefficient $\gamma=-10$.
         Simulations were performed under a Wright-Fisher model without linkage.
@@
 
+\begin{figure}[ht!]
+    \centering
+    \includegraphics[width=0.6\textwidth]{../figures/underdominance.Ne_5000.s_0.0005.pdf}
+    \caption{
+        Comparison of predicted (\moments) and simulated SFS with \emph{over}dominant
+        selection. In this comparison, $N_e=5000$ and $s=0.0005$, so that
+        the population-size scale selection coefficient $\gamma=5$.
+        Thus, with positive selection, heterozygotes are favored.
+        Simulations were performed under a Wright-Fisher model without linkage.
+    }
+    \label{fig:overdominance-validation}
+\end{figure}
+
 \end{document}

### 2024-08-19 figures/single_population_stabilizing_selection.py

@@ -13,7 +13,6 @@ matplotlib.rc("axes", labelsize=7)
 matplotlib.rc("axes", titlesize=7)
 matplotlib.rc("legend", fontsize=6)
 
-
 colors = plt.rcParams["axes.prop_cycle"].by_key()["color"]
 
 
@@ -131,7 +130,7 @@ def plot_stabilizing_selection_diagram(ax, VS=1, VG=0.1, optimum=0):
     ax.text(0.605, 0.53, "$V_S$", va="center", ha="center", color="k", fontsize=6)
     ax.set_ylim(bottom=0)
     ax.set_xlim([-2, 2])
-    ax.set_xlabel("Phenotype ($z$)")
+    ax.set_xlabel("Phenotype")
     ax.set_xticks([0])
     ax.set_xticklabels(["Optimum"])
     ax.set_yticks([])
@@ -191,8 +190,8 @@ def plot_shoc(ax, VS=1):
 
     data = pickle.load(open("../unlinked/unlinked_VGs_array.pkl", "rb"))
     x = np.logspace(-5, 0, 101)
-    ax.plot(x ** 2 * Ne, 4 * Ne * mu * x ** 2, ":", color="gray", lw=0.5, zorder=0)
-    ax.plot(x ** 2 * Ne, 4 * mu * VS * np.ones(len(x)), ":", color="gray", lw=0.5, zorder=0)
+    ax.plot(x ** 2 * Ne, 4 * Ne * mu * x ** 2, ":", color="gray", lw=0.75, zorder=0)
+    ax.plot(x ** 2 * Ne, 4 * mu * VS * np.ones(len(x)), ":", color="gray", lw=0.75, zorder=0)
     SHOC = 4 * mu * VS / (1 + VS / Ne / x ** 2)
     ax.plot(
         x ** 2 * Ne,
@@ -206,22 +205,22 @@ def plot_shoc(ax, VS=1):
     ax.scatter(
         SDs ** 2 * Ne,
         VGs,
-        s=16,
-        marker="+",
-        color=colors[0],
-        linewidths=0.75,
+        s=14,
+        marker="o",
+        edgecolors=colors[0],
+        facecolors="none",
+        linewidths=1,
         label="moments",
-        zorder=3,
+        zorder=2,
     )
     ax.scatter(
         data["SD"] ** 2 * Ne,
         data["VG"],
         color=colors[1],
-        s=10,
-        marker="x",
-        linewidths=0.75,
[…]

### 2024-08-20 doc/manuscript.tex
careful reading

@@
     extensively studied for single populations at steady state. However,
-    natural populations are seldom at steady state. Instead, they are often
-    structured in some way, and admixture and introgression events may be
-    common, including over human evolutionary history. Because stabilizing
-    selection results in selection against the minor allele at a
-    trait-affecting locus, alleles from the minor parental ancestry will be
-    selected against after admixture. Here, we show that the site-frequency
-    spectrum can be used to model the genetic architecture of such traits,
-    allowing for the study of trait architecture dynamics in complex
-    multi-population settings. We use a simple deterministic two-locus model to
-    predict the reduction of introgressed ancestry around trait-contributing
-    loci. From this and individual-based simulations, we show that
-    introgressed-ancestry deserts are enriched around such loci. When
-    introgression between two diverged populations occurs in both directions,
-    as has been inferred between humans and Neanderthals, the locations of
-    introgressed-ancestry deserts will tend to be shared across populations. We
-    argue that stabilizing selection for shared phenotypic optima may explain
-    recent observations in which regions of depleted human-introgressed
-    ancestry in the Neanderthal genome overlap with Neanderthal-ancestry
-    deserts in humans.
+    natural populations are seldom at steady state and are often structured in
+    some way. Admixture and introgression events may be common, including over
+    human evolutionary history. Because stabilizing selection results in
+    selection against the minor allele at a trait-affecting locus, alleles from
+    the minor parental ancestry will be selected against after admixture. We
+    show that the site-frequency spectrum can be used to model the genetic
+    architecture of such traits, allowing for the study of trait architecture
+    dynamics in complex multi-population settings. We use a simple
+    deterministic two-locus model to predict the reduction of introgressed
+    ancestry around trait-contributing loci. From this and individual-based
+    simulations, we show that introgressed-ancestry deserts are enriched around
+    such loci. When introgression between two diverged populations occurs in
+    both directions, as has been inferred between humans and Neanderthals, the
+    locations of introgressed-ancestry deserts will tend to be shared across
+    populations. We argue that stabilizing selection for shared phenotypic
+    optima may explain recent observations in which regions of depleted
+    human-introgressed ancestry in the Neanderthal genome overlap with
+    Neanderthal-ancestry deserts in humans.
 
@@
 encountered novel environments. Despite some cases of adaptive introgression,
-most introduced alleles were likely to have been selected against in humans
+most introduced functional alleles likey were selected against in humans
 \citep{harris2016genetic, juric2016strength, veller2023recombination}. Since
@@
 
-There is growing genetic evidence that humans reciprocally contributed to
-Neanderthal genomes \citep{kuhlwilm2016ancient, hubisz2020mapping,
-harris2023diverse, li2024recurrent}. This gene flow occurred tens to hundreds
-of thousands of years prior to Neanderthal introgression in humans during the
-global dispersal of modern humans around 60 ka. This is supported by early
-\emph{H. sapiens} outside of African around 200--100 ka \citep{schwarcz1988esr,
-grun2005u, harvati2019apidima, beyer2021climatic}, potentially overlapping with
-Neanderthals and providing opportunities for early contacts. While estimates of
-the genomic contribution of early humans to Neanderthals vary, around $6\%$ of
-later Neanderthal genomes may trace through this admixture event
-\citep{harris2023diverse}. Under a ``load'' model, if human-related haplotypes
-carried fewer deleterious alleles due to their larger long-term effective
-population size, human-introgressed DNA would have been favored in Neanderthal
-genomes. The replacement of Neanderthal mitochondrial and Y chromosomes by
-early human haplotypes appears to support this model of post-admixture positive
-selection in the Neanderthal lineage \citep{posth2017deeply,
-petr2020evolutionary}.
-
-Models for selection against introgressed alleles are often based on load
-arguments (that is, due to differences in the rate of accumulation of
-unconditionally deleterious variation in populations of different sizes) or
-hybrid incompatibilities \citep{muller1942isolating}. These models, founded in
-population genetics theory, rarely take into account selection operating on
-phenotypic traits or the relationship between genetic and phenotypic variation.
-Many phenotypic traits are thought to be under stabilizing selection
-\citep{sanjak2018evidence, sella2019thinking}, including gene regulation
-\citep{gilad2006natural, hodgins2015gene, price2022detecting}. Because some of
-the strongest signals of selection against Neanderthal-introduced alleles are
-in regulatory regions \citep{sankararaman2014genomic}, stabilizing selection on
-quantitative traits may be particularly relevant to the dynamics of functional
-genetic variation after introgression among hominins.
+There is growing genetic evidence that Neanderthals reciprocally received
+genetic material from early humans \citep{kuhlwilm2016ancient,
+hubisz2020mapping, harris2023diverse, li2024recurrent}. This gene flow occurred
+tens to hundreds of thousands of years prior to Neanderthal introgression in
+humans during the global dispersal of modern humans around 60 ka. The earlier
+admixture is supported by \emph{H. sapiens} outside of African around 200--100
+ka \citep{schwarcz1988esr, grun2005u, harvati2019apidima, beyer2021climatic},
+potentially overlapping with Neanderthals and providing opportunities for early
+contacts. While estimates of the genomic contribution of early humans to
+Neanderthals vary, around $6\%$ of later Neanderthal genomes may trace through
+this admixture event \citep{harris2023diverse}. Under a ``load'' model, if
+human-related haplotypes carried fewer deleterious alleles due to their larger
+long-term effective population size, human-introgressed DNA would have been
+favored in Neanderthal genomes. The replacement of Neanderthal mitochondrial
+and Y chromosomes by early human haplotypes appears to support this model of
+post-admixture positive selection in the Neanderthal lineage
+\citep{posth2017deeply, petr2020evolutionary}.
+
+Models for selection against introgressed alleles are often based on
+deleterious load or hybrid incompatibilities \citep{muller1942isolating}. These
+models, founded in population genetics theory, rarely take into account
+selection operating on phenotypic traits or the relationship between genetic
+and phenotypic variation.  Many phenotypic traits are thought to be under
+stabilizing selection \citep{sanjak2018evidence, sella2019thinking}, including
+gene regulation \citep{gilad2006natural, hodgins2015gene, price2022detecting}.
+Because some of the strongest signals of selection against
+Neanderthal-introduced alleles are in regulatory regions
+\citep{sankararaman2014genomic}, stabilizing selection on quantitative traits
+may be particularly relevant to the dynamics of functional genetic variation
+after introgression among hominins.
 
@@
         genetic variance ($V_G$) is proportional to the
-        population-scaled mutation rate. When $V_M$ is large so that
-        mutational effects can be strong, $V_G$ is independent of the mutation
-        rate. The stochastic house-of-cards model (Eq.~\ref{eq:SHC})
+        population-scaled mutation rate. When $V_M$ is large ($N_e V_M > 1$) so that
+        mutational effects can be strong, $V_G$ is independent of $V_M$.
+        The stochastic house-of-cards model (Eq.~\ref{eq:SHC})
         interpolates these regimes, assuming steady-state dynamics
@@
         \citep[developed here using \moments,][]{jouganous2017inferring}
-        matches simulations assuming no linkage between loci affecting the
+        matches simulations assuming linkage equilibrium between loci affecting the
         trait, which align closely with the stochastic house-of-cards
@@
 evolution \citep{yair2022population}. We therefore expect a rapid divergence of
-trait architectures and an accumulation of fixed differences in the human and
-Neanderthal lineages at trait-contributing loci, even when the mean phenotype
-in each population remains close to the same trait optimum. When a derived
-allele at high frequency in one population is introduced to another population
-in which it was previously absent, it will be at low frequency (if the
-admixture proportion is low) and subsequently selected against. Likewise, if
-the ancestral allele is reintroduced to a population fixed for a derived
-allele, the ancestral allele will be at low frequency and will be selected
-against. Selection acts against the introgressed allele in either case, whether
-it is ancestral or derived and regardless of the historical relative sizes of
-the populations involved. In concurrent work, \citet{veller2024stabilizing}
-demonstrate this effect by deriving the expected decay of minor parental
-ancestry under stabilizing selection after admixture. This prediction contrasts
-with the population-genetics load model, in which haplotypes with fewer
-deleterious variants (such as those from the population with larger historical
-size) are favored after introgression in either direction.
+trait architectures in the human and Neanderthal lineages at trait-contributing
+loci, even when the mean phenotype in each population remains close to the same
+trait optimum. When a derived allele at high frequency in one population is
+introduced to another population in which it was previously absent, it will be
+at low frequency (if the admixture proportion is low) and subsequently selected
+against. Likewise, if the ancestral allele is reintroduced to a population
+fixed for a derived allele, the ancestral allele will be at low frequency and
+will be selected against. Selection acts against the introgressed allele in
+either case, whether it is ancestral or derived and regardless of the
+historical relative sizes of the populations involved. In concurrent work,
+\citet{veller2024stabilizing} demonstrate this effect by deriving the expected
+decay of minor parental ancestry under stabilizing selection after admixture.
+This prediction contrasts with the population-genetics load model, in which
+haplotypes with fewer deleterious variants (such as those from the population
+with larger historical size) are favored after introgression in either
+direction.
 
@@
 site-frequency spectrum to predict genetic variation under complex demographic
-scenarios, and which we use to partition predicted trait heritability by
+scenarios, which we use to partition predicted trait heritability by
 introgressed and non-introgressed variation. Using simulations with linkage, we
@@
 will tend to overlap in location across populations. We argue that stabilizing
-selection on shared trait optima may explain the overlap of introgressed
-ancestry deserts in human and Neanderthal genomes after reciprocal
-introgression, as has been recently observed \citep{harris2023diverse}.
+selection on shared trait optima may explain the overlap of
+introgressed-ancestry deserts in human and Neanderthal genomes after reciprocal
+introgression \citep{harris2023diverse}.
 
@@
 allele at locus $l$, and \(g_{l,i}\in\{0,1,2\}\) is their genotype at that
-locus (i.e., the number of derived alleles they carry at that locus). With
-linkage equilibrium between trait-affecting loci, the expected additive genetic
+locus (i.e., the number of derived alleles they carry). With linkage
+equilibrium between trait-affecting loci, the expected additive genetic
 variance is \(V_A = \sum_l 2p_l(1-p_l)a_l^2\), where $p_l$ is the allele
-frequency at locus $l$. We ignore dominance and epistatic effects (often argued
-to be a reasonable modeling choice, e.g., \citet{barton2002understanding,
+frequency at locus $l$. We ignore dominance and epistasis (often argued to be a
+reasonable modeling choice, e.g., \citet{barton2002understanding,
 zhang2002pleiotropic}), so that \(V_G = V_A\). We further ignore environmental
@@
 effects, as in][]{lande1975maintenance, simons2018population}, so that only
-genetic effects are considered.
+genetic effects are considered (so that \(V_P=V_G\)).
 
@@
 \left(\frac{V_S}{V_S+V_G}\right)^{1/2},\]
-so that as the genetic variance increases, population mean fitness decreases.
+so that as the genetic variance increases, mean fitness decreases.
 
@@
 allele, so that
-\begin{equation}
+\begin{equation}\label{eq:selcoeff}
     s\approx -\frac{a^2(1-2p)}{2(V_S+V_G)} 
@@
 sample size $n$, denoted $\Phi_n$), as
-\[V_{G,a} \approx 2a^2\sum_{j=1}^{n-1} \frac{j(n-j)}{n(n-1)} \Phi_n(j|a,\mu_a) da.\]
+\[V_{G,a} \approx 2a^2\sum_{j=1}^{n-1} \frac{j(n-j)}{n(n-1)} \Phi_n(j|a,\mu_a).\]
 Here, $\mu_a$ is the mutation rate of alleles with effect size $a$. Then
@@
 genetic variance is
-\[V_G \approx \int_{-\infty}^\infty V_{G,a}\mathcal{N}(0,V_M).\]
+\[V_G \approx \int_{-\infty}^\infty V_{G,a} dF(a),\]
+where \(F(a)\) is the cumulative distribution function of \(\mathcal{N}(0,V_M)\).
 $V_G$ can be computed periodically in this way to predict trajectories in
@@
 not small compared to $V_S$, using $s=(p-1/2)a^2/(V_S+V_G)$ provides estimates
-of $V_G$ that closely match simulations assuming free recombination between
+of $V_G$ that closely match simulations assuming linkage equilibrium between
 loci (Figures~\ref{fig:toy-admixture} and
@@
         probability of being trait increasing or decreasing. Depending on the
-        selection coefficient ($s\approx (p-1/2)a^2/V_S$) compared to $N_e$, $V_G$
+        selection coefficient (Equation~\ref{eq:selcoeff}) compared to $N_e$, $V_G$
         could increase or decrease after a sudden population size change.
-        Predictions using \moments match simulations with free recombination
+        Predictions using \moments match simulations with linkage equilibrium
         between trait-affecting loci
@@
 subsequent introgression events. We focus on parameter regimes relevant to
-human-Neanderthal history. In a simple toy model (which is not meant to
-perfectly match any particular known or inferred history, but instead
-demonstrate the effect of reciprocal introgression following population
-divergence) a population of size $N_e=10{,}000$ splits into two, one remaining
-size $10{,}000$ and the other shrinking to size $1{,}000$. They remain isolated
-for $2N_e$ generations (or 500 thousand years, assuming an average generation
-time of 25 years) and then introgression occurs from one branch to the other,
-contributing $5\%$ ancestry to the recipient population
-(Figure~\ref{fig:toy-admixture}A,B).
+human-Neanderthal history. In a simple toy model a population of size
+$N_e=10{,}000$ splits into two, one remaining size $10{,}000$ and the other
+shrinking to size $1{,}000$. They remain isolated for $2N_e$ generations (or
+500 thousand years, assuming an average generation time of 25 years) and then
+introgression occurs from one branch to the other, contributing $5\%$ ancestry
+to the recipient population (Figure~\ref{fig:toy-admixture}A,B).
 
@@
         without linkage between trait-affecting loci. In Deme 2, $V_G$ decreases
-        after divergence and population size reduction. After admixture in both
-        directions, $V_G$ increases and then decays again toward levels at steady
-        state. Predictions from \moments closely match observed $V_G$ in
-        simulations.
-        Here, mutations were drawn from a normal distribution of fitnesses
-        with $V_M=0.0025$.
+        due to their population size reduction. After admixture in both
+        directions, $V_G$ increases in the recipient deme
+        and then decays to steady-state levels. Predictions from \moments closely
+        match observed $V_G$ in simulations.
+        Here, mutations were drawn from a normal distribution with $V_M=0.0025$.
         Other parameters: $\mu=0.025$, $V_S=1$, and the optimal phenotype remained
@@
         reintroduced ancestral alleles under the reverse direction of gene flow.
-        In both cases, $V_G$ is increasingly due to new mutations and the genetic
+        In both cases, $V_G$ is increasingly due to new mutations as the genetic
         architecture of the trait turns over with time since admixture.
@@
 mutations occur at rate $\mu$ per haploid genome copy per generation, with
-effects drawn from some distribution of mutation effects.
+effects drawn from a normal distribution with mean 0 and variance $V_M$.
 
@@
 variance directly after admixture as
-\[V_G = fV_{G,0} + (1-f) V_{G,1} + 2f(1-f)\sum_l F_{2,l} a_l^2,\]
+\begin{align}
+V_G = fV_{G,0} + (1-f) V_{G,1} + 2f(1-f)\sum_l F_{2,l} a_l^2,
+\end{align}
 where $F_2 = (p_0 - p_1)^2$ is the squared difference in allele frequencies at
@@
 that depends on the divergence at trait-affecting loci between the populations
-weighted by the quadratic factor $2f(1-f)$. Notably, $V_G$ is expected to
+weighted by the quadratic factor $2f(1-f)$. We note that $V_G$ is expected to
 increase only in the second generation after admixture, so that selection
@@
 \end{align}
-where $L$ is the number of trait affecting loci and assuming $\E[a]=0$.
+where $L$ is the number of trait affecting loci.
 
@@
         reciprocal introgression.}
-        (A) A simple model of divergence and admixture between humans and
+        (A) A model of divergence and admixture between humans and
         Neanderthals. Using \moments, we computed predicted $V_G$ over time,
@@
         SNP) is plotted over the 50 thousand years following introgression.
-        For non-introgressed variants, we also plot $h^2$ per SNP weighted by
+        For non-introgressed variants, we also plot $h^2$-per-SNP weighted by
         allele frequencies matching those of introgressed variants. These
@@
 The average contribution of introgressed vs. non-introgressed SNPs to $V_G$
-(i.e., $h^2$ per SNP) can similarly by tracked over time. For the
+(i.e., $h^2$-per-SNP) can similarly by tracked over time. For the
 human-Neanderthal demographic model and genetic architectures considered
-here, the contribution per SNP of introgressed variants is initially lower than
+here, the contribution per-SNP of introgressed variants is initially lower than
 that of non-introgressed variants. These contributions change over time,
@@
 In the preceding sections, we found that approximating the dynamics of
-trait-affecting alleles using an underdominance model
+trait-affecting alleles using an underdominant selection model
 \citep{robertson1956effect} provides an excellent approximation of $V_G$ in
@@
 polygenicity, $V_G$ in individual-based simulations closely matches
-expectations without linkage. With increasing mutation rate, $V_G$ is reduced
-relative to those expectations. However, when the mutational variance is much
-larger ($2N_eV_M\approx 50$), we see the reverse trend
+expectations under linkage equilibrium. With increasing mutation rate, $V_G$ is
+reduced relative to those expectations. However, when the mutational variance
+is much larger ($2N_eV_M\approx 50$), we see the reverse trend
 (Figure~\ref{fig:supp-high-VM}). At low mutation rate, there is a close match
-between observed $V_G$ and unlinked expectations, although simulated values are
-slightly higher. As the mutation rate increases, $V_G$ increases to be
-relatively much larger than expectations, rather than smaller.
+between observed $V_G$ and expectations, although simulated values are slightly
+higher. As the mutation rate increases, $V_G$ increases to be relatively much
+larger than expectations, rather than smaller.
 
@@
 depends on a number of factors. The total mutation rate affects not only the
-polygenicity, with higher mutation rates causing higher number of segregating
-alleles, but also the mean recombination rate between those alleles, as they
-will be more or less densely distributed in the genome. The distribution of
-effect sizes plays an important role, as seen by the opposite trends in $V_G$
-relative to unlinked predictions (Figure~\ref{fig:supp-low-VM} vs.
-Figure~\ref{fig:supp-high-VM}), which are most apparent with high mutation
-rates, and thus high polygenicity. In such cases, there appear to be complex
-dynamics involving the reduction in $V_G$ due to the Bulmer Effect
-\citep{bulmer1971effect} and an increase in $V_G$ due to selective interference
-\citep{hill1966effect}.
+polygenicity, with higher mutation rates producing more segregating alleles,
+but also influences the mean recombination rate between alleles, as they will
+be more densely distributed in the genome. The distribution of effect sizes
+plays an important role, as seen by the opposite trends in $V_G$ distortions
+(Figure~\ref{fig:supp-low-VM} vs. Figure~\ref{fig:supp-high-VM}), which are
+most apparent with high mutation rates and thus high polygenicity. In these
+cases, there appear to be complex dynamics involving the reduction in $V_G$ due
+to the Bulmer Effect \citep{bulmer1971effect} and an increase in $V_G$ due to
+selective interference \citep{hill1966effect}.
 
@@
         \textbf{Reduced introgressed ancestry around alleles contributing to a
-        trait under stabilizing selection.}
-        Using a deterministic model
+        trait under stabilizing selection.} Using a deterministic model
         (Equations~\ref{eq:system-p}--\ref{eq:system-D}), we model the
@@
         neutral allele, initially absent from the recipient population so that
-        their frequencies equal the introgression proportion ($f=0.05$).
-        The neutral allele frequency is reduced at a rate that depends on
-        the effect size of the selected allele and the probability of
-        recombination between them. LD (as measured by $D=\Cov(p,q)$) decays
-        to zero over time.
-        (C) Alleles with strong effects are expected to result in a larger
-        depletion of introgressed ancestry around the selected locus.
-        (D,E) Compared to individual-based simulations, deterministic model
+        their frequencies equal the introgression proportion ($f=0.05$). The
+        neutral allele frequency is reduced at a rate that depends on the
+        effect size of the selected allele and the probability of recombination
+        between them. LD (as measured by $D=\Cov(p,q)$) decays to zero over
+        time. (C) Alleles with strong effects are expected to result in a
+        larger depletion of introgressed ancestry around the selected locus.
+        (D,E) Compared to individual-based simulations, the deterministic model
         predicts the dip in introgressed ancestry around trait-affecting loci,
-        when the mutation rate is low, so that trait-affecting are sparsely
-        distributed. When mutation rates (and thus polygenicity) are high,
-        selected alleles are close together, so that selective interference
-        is more pronounced and local ancestry is affected by multiple selected
-        alleles.
+        when the mutation rate is low, so that trait-affecting loci are
+        sparsely distributed. When mutation rates (and thus polygenicity) are
+        high, selected alleles tend to be more tightly linked, so that
+        selective interference is more pronounced and local ancestry is
+        affected by multiple selected alleles.
     }
@@
 \begin{align}\label{eq:system-p}
-    p_{t+1} & = p_t - s p_t(1-p_t)(1-2p_t),
+    p_{t+1} & = p_t + s_tp_t(1-p_t),
 \end{align}
-with \(s=a^2/2V_S\) for stabilizing selection.
-As discussed above, if \(p_0<1/2\), \(p_t\rightarrow0\), and if
-\(p_0>1/2\), \(p_t\rightarrow1\) as \(t\rightarrow\infty\).
+with \(s_t=(p_t-1/2)a^2/V_S\) for stabilizing selection
+(Equation~\ref{eq:selcoeff}).
 
 We consider a neutral locus separated from the selected locus by recombination
-probability $r$. Initially, the expected frequency of linked introgressed
-ancestry is \(q_0=f\), which changes over time due to linked selection on the
+fraction $r$. Initially, the expected frequency of linked introgressed ancestry
+is \(q_0=f\), which changes over time due to linked selection on the
 trait-affecting allele. Letting \(D=\Cov(p,q)\) be the standard covariance
@@
 \begin{align}\label{eq:system-q}
-    q_{t+1} & = q_t - s D_t(1-2p_t).
+    q_{t+1} & = q_t + s_t D_t.
 \end{align}
-Initially, \(D_0=\pm f(1-f)\), with \(D\) being
-positive if \(p_0=f\) and negative if \(p_0=1-f\). LD between the loci changes
-deterministically over time due to both selection and recombination, so that
+Initially, \(D_0=\pm f(1-f)\), with \(D\) being positive if \(p_0=f\) and
+negative if \(p_0=1-f\). LD between the loci changes deterministically over
+time due to both selection and recombination, so that
 \begin{align}\label{eq:system-D}
-    D_{t+1} = D_t - r D_t - s D_t (1-2p_t)^2.
+    D_{t+1} = D_t - r D_t + s_t D_t (1-2p_t).
 \end{align}
-Together, this forms a
-simple nonlinear system of equations for the deterministic change in allele
-frequencies at the two loci and LD between them.
+Together, this forms a simple nonlinear system of equations for the
+deterministic change in allele frequencies at the two loci and LD between them.
 
@@
 We assessed the accuracy of the deterministic two-locus model using
-individual-based forward-in-time simulations \citep{thornton2019polygenic}
-under a simple demographic model (Figure~\ref{fig:toy-admixture}A), with
-introgression fraction \(f=0.05\). We simulated a single chromosome of length 1
-Morgan, with all mutations having effect sizes \(\pm a\) (\(a=0.02\) or
-\(0.05\), in separate simulations), and we varied the total per-chromosome
-mutation rate (\(\mu=0.001\), \(0.0025\) and \(0.01\)). For each
-fixed-difference mutation in the parental populations, we determined the
-average introgressed ancestry surrounding such loci $4{,}000$ generations after
-admixture (Methods).
+individual-based simulations \citep{thornton2019polygenic} under a simple
+demographic model (Figure~\ref{fig:toy-admixture}A), with introgression
+fraction \(f=0.05\). We simulated a single chromosome of length 1 Morgan, with
+all mutations having effect sizes \(\pm a\) (\(a=0.02\) or \(0.05\), in
+separate simulations), and we varied the total per-chromosome mutation rate
+(\(\mu=0.001\), \(0.0025\) and \(0.01\)). For each fixed-difference mutation in
+the parental populations, we determined the average introgressed ancestry
+surrounding such loci $4{,}000$ generations after admixture (Methods).
 
 The deterministic model (Equations~\ref{eq:system-p}--\ref{eq:system-D})
-provides a very good approximation when mutation rates are small. As the
-mutation rate increases in these simulations, trait-affecting alleles are more
-densely distributed along the chromosome. Deviations from the deterministic
-model are due to multiple selected alleles affecting local introgressed
-ancestry. For the highest mutation rate shown here, there can be many other
-trait-affecting alleles within a 1 cM window around any focal SNP, distorting
-dynamics away from predictions under the two-locus model.
+provides a very good approximation when mutation rates are small
+(Figure~\ref{fig:linkage}D,E). As the mutation rate increases in these
+simulations, trait-affecting alleles are more densely distributed along the
+chromosome. Deviations from the deterministic model are due to multiple
+selected alleles affecting local introgressed ancestry. For the highest
+mutation rate shown here, there can be many other trait-affecting alleles
+within a 1 cM window around any focal SNP, distorting dynamics away from
+predictions under the two-locus model.
 
@@
 \citep{yair2022population}. Stabilizing selection on a trait results in
-symmetric underdominant selection at trait-contributing loci
-\citep{robertson1956effect, keightley1988quantitative}, that is, selection
-against the minor allele. Some theoretical and simulation studies consider the
-effect of population differentiation and migration on the genetic architecture
-of a trait under stabilizing selection \citep[e.g.,][]{tufto2000quantitative,
-yeaman2011genetic, yair2022population}, but most previous work has focused on
-single population scenarios, often assuming steady state dynamics. Episodes of
-admixture and introgression commonly occur in many species' evolutionary
-histories, so understanding their effects on the architecture of complex traits
-is needed.
-
-Here, we have shown that a multi-population, non-equilibrium approach for the
-site-frequency spectrum with underdominance can be used to model the additive
-genetic variance of a trait under stabilizing selection. While this approach
-ignores some biological relevancies, such as linkage between sites,
-non-additivity and pleiotropy, it still provides important insights into the
-dynamics of trait architectures. For example, we can readily decompose
-contributions from alleles of different origins, either by source population or
-mutation time. It may therefore be useful for understanding the contributions
-of introgressed variants to complex traits, such as hominin-introgressed
-mutations in humans \citep{reilly2022contribution, wei2023lingering}. In the
-limited scenarios explored here, we find that the expected contributions of
-introgressed variants to heritability are complicated even in the purely
-additive case, depending on the distribution of effect sizes, demographic
-history and the time since admixture (Figures~\ref{fig:human-neand-h2} and
-\ref{fig:human-neand-VG}).
+selection against the minor allele, mirroring symmetric underdominance
+\citep{robertson1956effect, keightley1988quantitative}. Some theoretical and
+simulation studies have considered the effect of population differentiation or
+migration on the genetic architecture of a trait under stabilizing selection
+\citep[e.g.,][]{tufto2000quantitative, yeaman2011genetic, yair2022population},
+but most previous work has focused on single population scenarios, often
+assuming steady state dynamics. Episodes of admixture and introgression
+commonly occur in many species' evolutionary histories, so understanding their
+effects on the architecture of complex traits is needed.
+
+We find a multi-population, non-equilibrium approach for the site-frequency
+spectrum with underdominance can be used to model the additive genetic variance
+of a trait under stabilizing selection. While this approach ignores some
+biological relevancies, such as linkage between sites, non-additivity and
+pleiotropy, it still provides important insights into the dynamics of trait
+architectures. For example, we can readily decompose contributions from alleles
+of different origins, either by source population or mutation time. It may
+therefore be useful for understanding the contributions of introgressed
+variants to complex traits, such as hominin-introgressed mutations in humans
+\citep{reilly2022contribution, wei2023lingering}. In the limited scenarios
+explored here, we find that the expected contributions of introgressed variants
+to heritability are complicated even in the purely additive case, depending on
+the distribution of effect sizes, demographic history and the time since
+admixture (Figures~\ref{fig:human-neand-h2} and \ref{fig:human-neand-VG}).
 
@@
 
-Finally, turning to the distribution of introgressed ancestry deserts in humans
-and Neanderthals, \citet{harris2023diverse} found that regions of depleted
-Neanderthal ancestry in humans overlap more than would be expected by chance
-with regions lacking human-introgressed alleles in the Altai Neanderthal.
+Recently, \citet{harris2023diverse} observed that regions of depleted Neanderthal
+ancestry in humans overlap more than would be expected by chance with regions
+lacking human-introgressed alleles in the Altai Neanderthal.
 Human-introgressed ancestry in the Neanderthal genome is also depleted in
@@
 
-%\section*{Acknowledgements}
-%\begin{itemize}
-%    \item Lab members
-%    \item Kevin Thornton
-%    \item Bret Payseur
-%\end{itemize}
+\section*{Acknowledgements}
+
+I am grateful to Kevin Thornton, Bret Payseur, Carl Veller and Yuval Simons for
+valuable discussions, and Carl Veller for helpful comments on an earlier
+version of this manuscript. Support for this research was provided by the
+Office of the Vice Chancellor for Research and Graduate Education at the
+University of Wisconsin–Madison with funding from the Wisconsin Alumni Research
+Foundation.
 
@@
 
-This same result can be found by considering a haploid model
-\citep{negm2024effect}. While the selection coefficient takes an equivalent
-form to underdominance, we note that the mechanism differs: selection acts
-against the minor allele, rather than for the heterozygous genotype.
+This same result can be found by directly considering the marginal fitness of a
+focal allele against the expected haploid background, as pointed out by
+\citet{negm2024effect}. Thus, while the selection coefficient takes an
+equivalent form to underdominance, we note that the mechanism differs:
+selection acts against the minor allele, rather than against the heterozygous
+genotype.
 
@@
 For symmetric underdominance, the relative fitnesses of genotypes \(aa:Aa:AA\)
-are \(1:1-t:1\). We consider any value of $s$ (positive or negative,
+are \(1:1-t:1\). We consider any value of $t$ (positive or negative,
 corresponding to either under- or overdominance); with stabilizing selection,
@@
 
-Assuming random mating and no linkage, dominance or epistasis (so \(V_G=V_A\)),
+Assuming random mating, linkage equilibrium, and no dominance or epistasis
+(so \(V_G=V_A\)),
 \[V_G = \sum_l 2p_l(1-p_l)a_l^2 = \sum_l \pi_l a_l^2,\]
@@
 
+\hfill$\square$
+
 \end{document}
